### 2019-11-14 manuscript8.Rmd
First commit

---
title: "Measuring Stability and Change in Personal Culture Using Panel Data"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, individuals' dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple model for comparing their prevalence using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. The pattern of results is more consistent with the settled dispositions model than the active updating model, and most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes in belief. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggdag)
library(ggrepel)
library(here)
'%!in%' <- function(x,y)!('%in%'(x,y))

```

# Introduction

Culture is an important part of social life but cultures are continuously evolving. In 1972, for example, more than 40 percent of US adults supported a law outlawing interracial marriage. Three decades later, this opinion had become so uncommon that the question was removed from the US General Social Survey. How does this kind of cultural change happen?

Attempts to account for opinion change at the societal level have produced an array of often conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent, real attitudes. Within sociology, a number of models suggest that individuals lack the cognitive tools to maintain consistent beliefs on social and political issues. As a result, they construct responses in the interview setting, drawing on ideas they hear from opinion leaders and changing their attitudes when elite discourse changes [@zaller; @converse; @perrinmcfadden]. Cohort replacement theories of attitude change assume that individuals hold opinions, that people are unwilling to alter them in the face of societal change, and that public opinion changes in response to generational turnover [@mannheim; @ryder]. Another broad array of models argue that individuals and the public at large hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p. xi; @achen1992]. More recently, sociologists have suggested that individuals attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change that have their roots in theories of cultural sociology. The first of these individual-level models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another. Separate from this distinction is a second dimension of attitude formation that reflects degree to which individuals can consistently report the same opinion over time. Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller]. Attitudes can also be in a process of active updating but relatively easy to report, such as when individuals change their registered partisan affiliation.

In this paper, we make these two dimensions of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for comparing the prevalence of active updating using panel data. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By focusing on the pattern of attitude change at the individual level, we can elucidate the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that individuals lack clear opinions, some survey items appear elicit inconsistent responses and a few appear to elicit random responses. At the same time, individuals are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and individuals are more likely to report these attitudes consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

These findings generally are more consistent with the settled dispositions model than with the active updating model and support theories claiming that cultural change tends to come through generational turnover rather than persuasion. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change and the life course. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions. We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.

# Theoretical Background

## Belief Formation in Cultural Sociology

How does cultural change happen at the individual level? Let us make the question concrete by imagining a person who answers the same question each year for several years. The question could be anything, but assume it is this GSS question: "Please tell me whether you strongly agree, agree, disagree, or strongly disagree [with this statement]: 'a working mother can establish just as warm and secure a relationship with her children as a mother who does not work.'"

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all. We will explain this in greater detail below. For now, consider the following two models:

\begin{equation}
  y_{it} = y_{it-1} + \nu_{it}
  (\#eq:AR)
\end{equation}

\begin{equation}
  y_{it} = U_i + \nu_{it}
  (\#eq:FE)
\end{equation}

These models may seem similar at first glance, but they have very different implications for the pattern of individuals' responses over time. Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below. Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences. In the next two sections, we consider each of these models and briefly discuss their connections to sociological theory. In the section after that, we discuss the conditions under which each model is more likely to apply.

```{r dag, message=FALSE, fig.width=6.8, fig.cap="Different Models of Response Over Time"}
library(ggdag)
library(dagitty)
aum <- 
  dagitty("dag {
        y1 -> y2 -> y3 ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( aum ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )

aum_dag <- aum %>% tidy_dagitty() %>% ggdag() + 
  theme_dag_blank() + 
  labs(title = "Active Updating Model") +
  scale_y_continuous(limits = c(.8,3.2)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.margin = unit(c(1,1,.2,.2), "cm"))

sdm <- 
  dagitty("dag {
        U -> {y1 y2 y3} ;
        v1 -> y1 ;
        v2 -> y2 ;
        v3 -> y3 
        }")
coordinates( sdm ) <-
  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3, U = 2),
        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1, U = 3) )

sdm_dag <- sdm %>% tidy_dagitty() %>% ggdag() + 
  theme_dag_blank() + 
  labs(title = "Settled Dispositions Model") +
  scale_y_continuous(limits = c(.8,3.2)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.margin = unit(c(1,.2,.2,1), "cm"))

gridExtra::grid.arrange(aum_dag, sdm_dag, ncol = 2)

```

## Active Updating Model

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property, and this active updating model generally, is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings]. Concerns about panel conditioning, in which a person's participation in a survey influences her subsequent answer choices, either by changing her non-survey behavior or by altering how they respond to the survey without changing behavior, also imply some sort of active updating on the part of the respondent [@warrenhalpernmanners].

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities [@gross09; @degloma]. As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.

The AUM makes no assumption about the distribution of $\nu_{it}$. Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$. This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.

## Settled Dispositions Model

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are temporary influences only.

The SDM is most consistent with the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp. 66ff]. In other words, updating happened, but it happened in the past, prior to the time of the study. In this sense, $U_i$ reflects the "habitus." We will say more about this below.

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within individuals these shifts will be erased over time as individuals return to their baselines.

Metaphorically, one can think of the active updating model as a "traveler" and the settled dispositions model as a "visitor." That is, in the active updating model, the traveler wakes up in the morning, walks in a direction for a while, goes to sleep, and then wakes up the next day to pick up the journey exactly where she left off. In the settled dispositions model, the visitor also travels in a particular direction each day but goes home at the end of the day and wakes up in the same bed each morning.

The models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time. In the SDM, changes tend to revert because each person returns to their existing baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_. We consider the implications of these predictions below.

## Consistency in Response
Neither model makes any assumptions about the magnitude of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall amount of change in responses from wave to wave and the persistence of this change. Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make much greater inference about the applicability of different theories to these items.

## Combining the Models

These two dimensions of attitude change give us a two-by-two matrix of patterns for attitude change. The first dimension reflects what we call active updating, or the idea that more recent responses are better predictors of current responses that older responses. In other words, active updating means that when changes happen, they are more likely to persist than revert. We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses. Individuals might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their older positions. The second dimension reflects what we call stability, or how consistent individuals are in reporting their beliefs once we account for the amount of active updating.

With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them. Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix. We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes. Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.

\begin{center}
\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
\hline
 & Low Active Updating & High Active Updating \\ 
\hline
High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\ 
\hline
Low Consistency &  \\ 
\hline
\end{tabular}
\end{center}

### Age and Attitude Formation

In broad terms, cohort replacement theories of public opinion change suggest that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo]. Implied in these models is the notion that individuals have a period of "cohortization" in which they form beliefs.

To continue the metaphor above, cohort replacement theories imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition). The concept of a "cohort effect" is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman]. Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo]. To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick]. Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].

Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender. Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort” change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts. Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].

As Table ## suggests, items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults in the population, but high active updating for young people. Because these approaches generally assume that beliefs are prominent in the minds of respondents, the implication is that these beliefs are reported with high consistency in adulthood. At younger ages, however, we expect more volatility as attitude development occurs. We note, however, that there is some evidence that volatility of settled beliefs increases as individuals change as well [@krosnickalwin], but we do not expect this to be the general pattern.

### Inconsistent Opinions

Not all cultural change occurs through cohort replacement processes, and there is abundant evidence of large-scale public opinion shifts that occur in a span of time too short to be driven entirely by cohort replacement. This suggests that, at times, even middle-aged individuals update their opinions in persisting ways.

In contrast to cohort replacement theories, a range of theories posit that, outside of a well informed elite, individuals lack consistent opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse]. For example, when asked whether the government should spend more on health care, forming an opinion requires an individual to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result. Most individuals do not have this information readily stored in their minds.

Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment. This explains why individuals are susceptible to question-order, priming, and recency effects [@zaller]. However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues. Shifts in public opinion can also come about when elites shift their opinions on issues, providing new signals to the general public.

In general, then, we expect low-profile and complex public policy issues to show a low rate of active updating and a low rate of consistency. By definition, most questions of public policy should fall into this group, since only a few issues can sustain a relatively high level of public attention. The small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do spotlight one issue that saw significant change in elite opinion. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of rights for homosexuals. Leaders of the Democratic Party including President Barack Obama came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, then, we might expect that public opinion shifts notably. This should manifest as a high degree of active updating for the whole population, not just young people.

Outside of a lack of information on public policy issues, there are two other reasons why individuals might demonstrate low consistency and low active updating. First, individuals might be unwilling to report socially undesirable beliefs if they hold them, and as a result obfuscate their true position. Similarly, some issues might not require a lot of information to comprehend, but might be so far from the public consciousness as to be nonsensical to respondents. We expect both of these conditions to produce a similar pattern of low active updating and low consistency.

### Updating Opinions

The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, individuals update their opinions across the life course in response to gathering new information and interacting with changing social environments.

Some of these approaches, common in political science, suggest that individuals consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman]. These models do not tend to make distinctions about what kinds of attitudes shift over time and what do not. However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.

Other theories suggest that only some beliefs are open to updating during adulthood. A number of theories of attitude development posit that individuals hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein]. These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find little support for Lakoff's model and significant support for ideological identification as this "core" belief in the political space. If these  individuals have the ability to regularly assess new information against these core beliefs, they should also be able to recognize their position in a multiple-choice survey questionnaire and report them with limited error [@vaisey09]. In other words, we expect that core beliefs such as ideological identification should have high consistency and low active updating, and conversely that items that show these traits to have the potential to be core beliefs.

In a similar vein of research focused on political polarization, Baldassari and Gelman [@baldassarigelman] find little increase in the correlation between specific policy positions over time, suggesting limited active updating of these items. What they do find is an increased alignment between partisan identification and policy positions. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, partisan affiliation should demonstrate the highest level of active updating, and that it should be reported with a high degree of consistency.

### Public vs. Private Culture

A final dimension that likely influences the consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances, which might lead to a low degree of consistency and active updating. However, when public culture provides a strong "scaffold" for individuals through clear signals of how their identities and social locations should influence how they should respond to a survey question, it becomes easier for them to maintain consistency [@lizardostrand].

We have touched on this idea already with our discussion of the idea that high-profile elite opinion change might drive opinion change in the broader public, or by noting that high-profile issues might be easier to report consistently because individuals receive signals that support that position. Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not in an individual's head might demonstrate both higher rates of active updating and greater consistency.

### Toward Theoretical Synthesis

The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating. As such, this disagreement forms the primary structure of our analysis. A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development. We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating, but we should still observe high consistency and low active updating.

If what we find is a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions. Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work. We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

# Research Questions and Expectations

With these considerations in mind, we ask the following questions:

First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ As we just mentioned, in a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18). However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.

Third, _are there systematic differences in item content between questions that fall in different parts of our matrix?_ The preceding sections have made some specific predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs. As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.

# Analytic Strategy

To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below. While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.

## Statistical Model

### Basic Models

Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time. We first consider measuring the amount of active updating. Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:

\begin{equation}
  E(y_{i3}) = y_{i2}
  (\#eq:AUM2)
\end{equation}

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response. If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.

The SDM makes the following prediction:

\begin{equation}
  E(y_{i3}) = U_i
  (\#eq:SDM2)
\end{equation}

Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction like this:

\begin{equation}
  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
  (\#eq:SDM2B)
\end{equation}

That is, the SDM predicts that the average response of previous waves is the best predictor of the next response. If change is non-persisting (i.e., if our metaphorical visitor eventually always returns home), then taking the average of the last two places we saw her will be our best guess about the location of her home base.

### Combined Model

Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor. If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition. Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$. If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model. However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, we begin to observe evidence that at least some individuals in the population engage in active updating. We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:

\begin{equation}
  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
  (\#eq:MODEL)
\end{equation}

Rather than estimate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B). If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.

At the same time, our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating in responses. We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model. If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the population average, and $\beta = 0$. If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.

### Limitations of the Method

Three challenges limit our ability to evaluate the relative presence of settled dispositions and active updating models using our approach. These challenges are not fatal, but put some limitations on the conclusions we will be able to reach.

The first challenge is that our model is designed to allocate variance explained to each of the prior predictors rather than to assign probabilities to each data-generating process per se. Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes. If we were willing to assume that persisting and non-persisting changes were drawn from the same distribution, we could make inferences about the relative proportion of people coming from each data generating process, but we have no reason to believe this assumption.

The second challenge is measurement error, which is a form non-persisting change. In the attitude change literature, measurement error means different things to different authors. For some, it represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure. For other theories, measurement error reflects individuals' inability to accurately report their response. It is also possible that measurement error simply reflects errors of selection and interpretation (misunderstanding the question or incorrect coding).

Because measurement error "looks the same" as non-persisting real change (and because the latter is sometimes interpreted as the former), estimates of $\phi$ will be biased toward .5. There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings]. On the other hand, these previous studies of measurement reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them completely. Because of this issue of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.

The third challenge of our analysis is that we focus exclusively on predicting wave 3. If for some reason individuals have a high likelihood of changing between waves 2 and 3, then our ability to predict responses at wave 3 with either wave will be limited and $\beta$ will be low. Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3. If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error. Then again, it may not be reasonable to consider this sort of change "persistent" in any meaningful way.

In addition to these three challenges, there are two forms of change that our model is not well designed to account for. The first is change in the overall population. Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates. The second is change in the variance of responses. If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will be absorbed into $\beta$ but not enter into $\phi$.

Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error. Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model. We can only detect whether there is any evidence of persisting belief changes, and therefore whether there is any evidence that active updating is taking place in the population.

## Analysis Steps

Our analysis proceeds in three steps to answer our three research questions. First, we evaluate the overall evidence in support of the active updating model. To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$. We calculate the posterior probability that the model with the free parameter fits the data better. If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.

Second, for variables that do show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is disproportionately concentrated among younger respondents. To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff. Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45. We test a range of cutoffs to ensure robustness of the overall pattern to specific ages.

Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for AUM across variables as suggested by existing theories. Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be largely inductive.

## Item Selection

We want to test our model on as broad a range of items tapping attitudes, beliefs, and social behaviors as possible. To do so, we estimate the model in equation \@ref(eq:MODEL) on data from the three different three-wave panel data sets of the General Social Survey, conducted from 2006 through 2014.

We sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves. These questions tended to come from the "core" of the GSS, a set of questions asked in each wave. Rotating topical modules asked during the panels were only asked in select waves or were not asked consistently to the same people over time, leaving too few cases with complete data to analyze.

Since our theoretical framework focuses on attitudes, beliefs, and social behaviors, we excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer. We also exclude questions that ask about an individuals' childhood. While testing these questions with our method is possible, they are beyond the scope of our theoretical framework. To organize our presentation of such a large number of variables, we follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on their subject material. Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutional leaders, questions about public spending, and questions about social life.

We also follow Hout and Hastings [-@houthastings] in recreating some commonly used scales designed to capture attitudes about gender roles, access to abortion, and social trust. This includes Rossi's six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany"). We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are. We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman].

Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergymen. We combine four parallel questions about how frequently individuals socialize to create a "social life" scale. We combine four questions about support for suicide under different circumstances. We also created a scale of support for police use of violence against criminal suspects by averaging five binary questions about the conditions under which individuals support police use of violence.

In total, we test the model on 184 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves. Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.

[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options. These results are reported in an the Appendix.

# Results

```{r}
load(here("data", "results_df.Rdata"))
load(here("data", "label_vars.Rdata"))
load(here("data", "aum_vars.Rdata"))

```

Our model generates two parameter estimates of interest for each GSS item in our analysis: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3. High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating. $\phi$ is our measure of active updating and captures the relative proportion of wave 3 explained by wave 2. If responses are generated through a settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3). As the evidence of active updating increases, $\phi$ will increase toward 1. Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3. This is a highly unlikely scenario.

## Evidence for Active Updating

To evaluate the evidence in favor of the active updating model, we compare for all 184 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$. If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.

Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 184 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.

```{r phihist, message=FALSE, fig.width=6.8, fig.cap="Distribution of phi estimates and probabilities that items show evidence of active updating."}

res.df %>%
  filter(grepl("_c", var)==FALSE) %>%
  select(var, p, prob_aum) %>%
  gather(key = "key", value = "value", -var) %>%
  mutate(key = recode(key, "p"="Phi estimates", "prob_aum"="Probability of some active updating")) %>%
  ggplot(aes(x = value, y=..count../sum(..count..))) +
  geom_histogram(fill = "gray", color = "black", bins = 10) + 
  geom_vline(xintercept = .5, color = "black", linetype = 2) + 
  theme_classic() + 
  facet_wrap(~key, scales = "free_x") +
  labs(x = "", y = "Proportion of items") + 
  theme(axis.title=element_text(size=9))
```

\begin{figure}[t]
\includegraphics{figures/aum_prob_dist.pdf}
\centering
\caption{Distribution of phi estimates and probabilities that items show evidence of active updating.}
\label{fig:phi_hist}
\end{figure}

On the left side of the figure, we see that the majority of $\phi$ estimates fall between .5 and .6, meaning that wave 2 is only a slightly better predictor than wave 1 for most items. This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.

To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates. Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2, and only 147 (57 percent) of those individuals maintained the same response at wave 3. Under the settled disposition model, we would expect that about 50 percent or 129 individuals would maintain the same response, so we really only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting attitude change. In other words, even for the items that do show strong evidence of active updating, the overall rate of attitude change in the population is small.

The right side of Figure \@ref(fig:phihist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period. That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they will tend to revert to their previously held position. This group includes many items addressing abortion, civil liberties, confidence in institutions, and views on race and gender.

We will discuss in more detail below how different items perform. To answer our first question, however, we need only focus on the overall distribution. We find that about 40 percent of items show no evidence at all for active updating among GSS respondents. And among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent changes. We can only be really confident in detecting persistent change among a small minority of items, perhaps 1 in 5. This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.

## Age Heterogeneity

Our second research question is about whether there is evidence of cohort formation on some items. That is, we want to see if there is evidence that younger respondents are updating their views more than older respondents. Although it's impossible for any item to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.

Of the 111 items that showed evidence for active updating in the last section, 23 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not. Figure \@ref(fig:agegroupcomparison) plots the estimates of these 23 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.

```{r agegroupcomparison, message=FALSE, fig.width=6.8, fig.height=8, fig.cap="Comparison of phi estimates for individuals over and equal to to less than 30 years old."}

load(here("data", "age_results.Rdata"))
load(here("data", "good_vars.Rdata"))

bind_rows(age.results) %>%
  mutate(ci.lower = estimate - 1.645 * std.error,
         ci.upper = estimate + 1.645 * std.error) %>%
  left_join(good_vars) %>%
  mutate(age = recode(age, "old"="Over 30", "young"="30 & under")) %>%
  ggplot(aes(x = reorder(var.name, estimate), y = estimate, color = age)) + 
  geom_hline(yintercept = .5, color = "black", linetype = 2) + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  coord_flip() + 
  theme_minimal() + 
  scale_color_manual(values = c("black", "gray")) + 
  labs(x = "", y = "Phi estimate", color = "") +
  theme(axis.text=element_text(size=7),
        axis.title=element_text(size=9),
        legend.position = "bottom")

```

\begin{figure}[t]
\includegraphics{figures/age_differences_phi.pdf}
\centering
\caption{Comparison of $\phi$ estimates for individuals over and equal to to less than 30 years old.}
\label{fig:age_group_comparison}
\end{figure}

The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents. These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die. These items tend to be in categories where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable for the rest of life.

For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address pass discrimination, all evidence of active updating disappears for individuals over 30. This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but that over time they harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick]. For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals. This pattern is consistent with the "increasing persistence" hypotheses.

Eight of these items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals. These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, views on whether aging parents should live with their children, and views on whether police should be allowed to use force against suspects who are verbally abusive toward them. Some of these items, such as views on whether aging parents should live with their children, might be things young people are not forced to consider until later in life and as a result do not form clear opinions until that exposure. This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is relatively unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previous theorizing has accounted for.

The remaining 89 items explored in this analysis (just under half of all items) show evidence for some active updating but do not show consistent evidence for age heterogeneity, again suggesting a more complicated relationship between age and attitude change than previously theorized. However, this does not mean these show strong evidence of an active updating model. These items may simply be susceptible to updating for a small proportion of the population.

## Item heterogeneity

While there is a high proportion (40 percent) of items that show no evidence of active updating, and those that do show evidence show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change. Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.

Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items colored by whether they preferred the $\phi = .5$ constraint or not. Although we discuss the pattern of items below, we label a few items that stand out. Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.

```{r phibetascatter, message=FALSE, fig.width=6.8, fig.height = 6.0, fig.cap="Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint."}
res.df %>%
  filter(grepl("_c", var)==FALSE) %>%
  mutate(aum = ifelse(var %in% aum_vars, "\nPersistent\nChange\n", "\nNo Persistent\nChange\n")) %>%
  ggplot(aes(x = p, y = b1)) + 
  geom_point(aes(fill = aum, shape = aum)) + 
  scale_shape_manual(values = c(21, 22)) + 
  scale_fill_manual(values = c("gray", "black")) + 
  geom_text_repel(data = label.vars, aes(label = var.name), size = 3) + 
  theme_minimal() +
  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
  theme(axis.title=element_text(size=9))
```

\begin{figure}[t]
\includegraphics{figures/phibeta_scatter.pdf}
\centering
\caption{Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint.}
\label{fig:phibeta_scatter}
\end{figure}

Items showing evidence for active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6. A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.

Which items show the strongest evidence for active updating? There is no way we can possibly discuss all 184 items in detail without the discussion becoming tedious. Although we include $\phi$ estimates for all items in Appendix B, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question. We constrain items that showed no evidence of active updating to $\phi = .5$. In addition to showing the median and interquartile range of each distribution, the figure also highlights the item in each group that shows the greatest degree of evidence for active updating.

```{r summary, message=FALSE, fig.width=6.8, fig.height=9.0, fig.cap="Summary of phi estimates for all items, by topical group."}
load(here("data", "summary_plot_data.Rdata"))

plot.data %>%
  ggplot(aes(x = reorder(small.cat, median))) + 
  geom_pointrange(aes(y = median, ymin = q25, ymax = q75)) + 
  geom_point(aes(y = min), shape = 21) + 
  geom_point(aes(y = max),shape = 21) + 
  geom_text(aes(y = max, label = var.name), size = 2, position = position_nudge(y = .07)) + 
  coord_flip() + 
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=9),
    legend.text=element_text(size=9)
    )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")

```

\begin{figure}[!ht]
\includegraphics{figures/summary_plot.pdf}
\centering
\caption{$\phi$ estimates for all items by topical group.}
\label{fig:summary}
\end{figure}

There is a lot to process, even in this summary figure. The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low in absolute terms. Only two groups of items have median $\phi$ values above 0.6: public spending and religious activity. In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]

[^error]: While we cannot control for measurement error in our analysis, we can take steps to mitigate its impact. Appendix B presents results comparing items with more than three scale points to coarsened versions of these question with either two or three scale points. As noted previously, low measurement error might be a reasonable assumption for some items. Previous studies using different approaches to measuring the reliability of survey reports suggest that some items captured in our study, such as whether a person owns a gun, are measured with a high degree of reliability ($> 0.9$) [@houthastings]. For other items, such as confidence in the leadership of major companies ($\phi = .58$), reliability might be as low as 0.5. There is very little correlation ($\rho = .165$) between our $\phi$ estimates and reliability estimates.

Consistent with the findings and theory of Vaisey and Lizardo [-@vaiseylizardo], we see that over half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B). These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample. Views on these "core" issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood (and thus becomes eligible to participate in the GSS). This means that, for most of these items, social change must occur through cohort succession rather than through individual change.

There are several exceptions to this general pattern, however, even in categories with otherwise very low $\phi$ values. The items with the largest values generally have one or more attributes in common. We consider these attributes one by one in order to give some general impressions of the pattern.

Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them, tapping the idea that public culture is more persistent than personal culture. If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue. This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, as well as how often individuals engage in "religious activities", which are more nebulous questions that display less active updating. Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind). Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.

Other high-$\phi$ items have a _changing referent_. That is, although the item wording is the same, the object to which the question refers may change between survey waves. The most obvious example of this is the item about confidence in the executive branch of the federal government (which has the highest $\phi$ value of all items in the analysis). The president changed between the 2008 and 2010 waves of the GSS, meaning that the question no longer referred to the same administration. If we generate separate estimates for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, we see that it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$). In the 2006-10 panel, waves 1 and 2 have almost no predictive power ($\beta = .10$). For the 2010-14 panel, which takes place entirely during the Obama administration, $\phi$ moves much closer to .5 and predictive power increases ($\phi = .57$, $\beta = .65$).

Likewise, all of public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas. The change of administration and changing federal spending policies likely affected all spending items. The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession). If real external change is happening, we should see exactly this sort of pattern.

Perhaps the most striking pattern in our findings is about questions about _gay rights_. There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating. Questions about civil liberties for gays and about gay marriage are the highest in their categories. The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views. By comparison, abortion attitudes show almost no evidence for change.

Finally, questions about religious activity, identity and (to a lesser extent) religious beliefs generally show some evidence for active updating. In some cases, this fits with the general pattern of external mechanisms (e.g., attendance and identification are largely public acts).

### Political Beliefs

Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, we examine these beliefs in greater detail than other items. Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for most of the questions in our data set tapping questions of political identity, the role of government, and specific policies, broken into four general areas. We break these items up into categories and remove a few items solely for ease of viewing.

```{r polvariables, message=FALSE, fig.width=7, fig.height=8, fig.cap="Phi and beta estimates for 43 political attitudes and identities."}
left_join(res.df, good_vars) %>%
  filter(small.cat %in% c("Politics & government", "Public spending",
                          "Sex, sexuality & abortion") |
           var %in% c("cappun", "divlaw", "gunlaw", "grass", "courts",
                      "letin1a")) %>%
  mutate(small.cat = ifelse(small.cat %in% c("Race & immigration", 
                                             "Guns, laws, crime & police")
                            | var %in% c("prayer", "uswary"), 
         "Misc. policy" , small.cat)) %>%
  filter(var %!in% c("abscale6", "absingle", "teensex", "xmarsex",
                     "homosex", "premarsx")) %>%
  mutate(var.name = gsub("Government", "Gov't", var.name),
         var.name = gsub("Abortion", "Abort", var.name),
         small.cat = ifelse(small.cat == "Sex, sexuality & abortion",
                            "Abortion & sex", small.cat)) %>%
  mutate(small.cat = factor(small.cat, c("Politics & government", "Public spending",
                                        "Abortion & sex", "Misc. policy"))) %>%
  ggplot(aes(x = p, y = b1)) + 
  geom_vline(xintercept = .5, linetype = 2, color = "gray") + 
  geom_point(aes(fill = small.cat), shape = 21) + 
  geom_text_repel(aes(label = var.name), size = 3) + 
  theme_bw() + 
  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
  theme(axis.title=element_text(size=9)) +
  facet_wrap(~small.cat) + 
  theme(legend.position = "none")

```

There are several notable features of the figure. First, political party affiliation is a clear outlier in the plot, with much greater consistency and much greater active updating than other items. In contrast, questions asking about specific public spending priorities tend to have very low consistency, at least compared to other political items. Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government –  display higher consistency than specific policy questions but weak evidence of active updating. It is important to note that for political views, however, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.

While individuals tend to be inconsistent in their responses to spending priorities, views on some other specific policy questions with both high and low salience are reported quite consistently. Views on abortion, especially when considered as a scale, are quite consistently reported. As with all other scales, aggregating the composite items increases consistency but does not affect the amount of active updating. As discussed previously, views on gay marriage display high levels of active updating. In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating. This is notable, as public opinion and policies regarding both have shifted considerably in recent years, with many people assuming change in both is being driven by the same underlying process. Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but change in views on marijuana use have been driven primarily by cohort replacement.

# Discussion and Conclusions

This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making. These two models of individual change are often implicit in analyses of belief change at the society level, but they are rarely examined empirically. We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models. We now revisit our research questions to summarize what we have learned.

First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model. Around 40 percent of all items show no evidence for updating, and even those items that do show a relatively low degree of updating. Adult Americans are highly consistent in their attitudes from year to year, and those who change tend to bounce back to where they were. What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago. Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is. But the average level of updating appears for most views appears to be low.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ For a limited subset of items, there seems to be evidence that younger respondents are updating their views whereas older respondents are not. This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014). Because the GSS begins with 18-year-olds, we may lack power to detect updating that occurs even earlier, so this may be an underestimate. This may be the case for the roughly 70 items that show no evidence of persistent change over time. Overall, however, there is significant heterogeneity in the relationship between age and attitude and behavior change.

Third, _are there systematic differences in item content between questions that are better described by each model?_ With 184 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population. We find patterns of responses that provide a range of support for diverse theories of attitude formation.  At the same time, our results are consistent with some general expectations about patterns in updating. Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating. Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18. The overall pattern of results, while diverse and at times hard to reconcile, does shed light on theoretical debates, to which we now turn.

## Implications for Cultural Sociology

What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area? We see two major implications, one theoretical and one methodological.

In the domain of cultural theory, our findings support the view that most cultural change happens slowly, through the mechanism of cohort succession. Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled during adulthood. The settled nature of these beliefs is generally coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report these beliefs wave after wave. In contrast, a range of views about race were so inconsistent that it would be difficult to call them either settled or updating. Even for those items that did display strong evidence for active updating (e.g. "whites rich or poor" and "whites work hard") were still extremely hard to predict wave after wave.

In contrast to other beliefs, however, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging some people to update their views who might not otherwise do so. By definition, salience is a limited resource, meaning only few beliefs and behaviors could change at this rate at a time. The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views. This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.

While the dominant pattern is stability during adulthood, the results suggest greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories. The most prominent and well supported theories of attitude change suggest a peak of susceptibility to attitude change early in adulthood and either a rapid or gradual decline in attitude change with age [@alwinkrosnick; @visserkrosnick]. An additional view, the "life stages" hypothesis, suggests a similar pattern with a surge in attitude change late in life [@visserkrosnick]. While this age-related decline appears to be true for many political attitudes, which are the those items most frequently studied, the pattern is not nearly as consistent for other kinds of attitudes. Some items, such views on the Bible, suggest perpetual openness to attitude change as individuals age. Others, such as views on most abortion questions, suggest that early adult socialization is so strong that opinions are hardened by the time most people reach even 18 years old.

For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change actually becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change. Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to factors that encourage openness to change at different ages. Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes. Doing so will expand our understanding of the social and psychological factors that that give rise to stable or variable attitudes [@howekrosnick].

Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is a strong start [@eatonetal; @vissermirabile]. However, this line of work should be cognizant of which attitudes stabilize in middle age and the relationship between social factors and these specific attitudes and explore how these factors link to attitudes that seem to change more frequently at these ages.

## Implications for Political Sociology

We said at the outset that these two models of individual culture underlie significant debates about public opinion formation, the formation of individual political beliefs over the life course, and the ability of individuals to hold and express clear opinions over time. What do our findings suggest for these debates?

A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30. These respondents might express different positions from wave-to-wave, but in guessing what an individual will say in the future, we are better off guessing the mean of their previous responses than their most recent response. While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30. With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still quite limited. Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.

These results are consistent with theories that posit that, at least in the current era, individuals organize their beliefs around ideological identity [@boutylinevaisey; @converse], rather than some other organizing heuristic such as partisan identification, moral views, religion, or particular policy positions. It is also consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification. Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time. This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.

The fact that there is not significant active updating for most political opinions should not be taken to imply that individuals are consistent in their positions. Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions. This is clearest when it comes to questions that deal with spending priorities. Prior responses to questions about government spending on foreign aid, science, drugs, child care, crime prevention, mass transit, national parks, and the problems of cities are bad predictors of what individuals will say later on.

That being said, individuals are not wholly without consistent opinions. On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next. While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.

The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents hold a general political identity and a few clear views on issues like abortion, generally attempt to align their identification to their views, and respond to elite opinion change when it provides clear signals. It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates. Individuals tend to have more-or-less settled views, especially on core questions such as ideology, and report (with a high degree of inconsistency) more specific policy positions based on these.

## Methodological Implications

Methodologically, our results highlight the challenges of evaluating population-wide attitude change using short-term panel studies. The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change. It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that change is too rare in a sample of adults to measure accurately on the vast majority of items. This strongly argues against using two-wave panels to measure attitude change, which do not allow researchers to separate persisting from non-persisting changes.

The fact that persistent change is practically nonexistent for many items bolsters the case for using repeated survey responses to measure the reliability of survey items [@alwin; @houthastings], since it is often a valid assumption that the underlying view is unchanging. At the same time, our results call for greater focus on methodological tools that can separate short-term attitude change from measurement error. While we generally assume that lasting changes in attitudes are more likely to influence behavior, this is not necessarily true. Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying this kind of change is difficult. Even for items related to abortion, where there is almost no evidence of persistent change and the overall picture is one of stability of belief, individuals do bounce around from year-to-year.

Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of younger individuals to understand how these attitudes are formed. Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful. In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly. Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike as as particularly useless. Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights). Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production" that give rise to persistent beliefs in adulthood.

\pagebreak

# Appendix A: $\phi$ values for all variables

Figures \ref{fig:phi_relig} through \ref{fig:phi_racgen} plot $\phi$ estimates for all items included in the analysis, grouped by subject material.

```{r loadfiguredata}
load(here("data", "figure_data.Rdata"))
```

```{r religphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about religious activity and beliefs, social life, subjective SES, and suicide."}
clean.pestimates %>%
  filter(grepl("_c", var)==FALSE) %>%
  filter(small.cat %in% c("Relig. Activity", "Relig. identity & beliefs",
                          "Social life", "Subjective SES", "Suicide")) %>%
  ggplot(aes(x = reorder(var.name, estimate), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(small.cat ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")

```

\begin{figure}[!ht]
\includegraphics{figures/relig_phi_estimates.pdf}
\centering
\caption{$\phi$ estimates for items about religious activity and beliefs, social life, subjective SES, and suicide.}
\label{fig:phi_relig}
\end{figure}

```{r poliphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about guns, law, crime and policing; politics and government; and public spending."}

clean.pestimates %>%
  filter(grepl("_c", var)==FALSE) %>%
  filter(small.cat %in% c("Guns, laws, crime & police", "Politics & government",
                          "Public spending")) %>%
  filter(var != "polhitok") %>%
  ggplot(aes(x = reorder(var.name, estimate), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(small.cat ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")

```

\begin{figure}[!ht]
\includegraphics{figures/poli_phi_estimates.pdf}
\centering
\caption{$\phi$ estimates for items about guns, law, crime and policing; politics and government; and public spending.}
\label{fig:phi_poli}
\end{figure}

```{r civlibphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about civil liberties, confidence in leadership, health, morale, and social trust."}
clean.pestimates %>%
  filter(grepl("_c", var)==FALSE) %>%
  filter(small.cat %in% c("Civil liberties", "Trust", "Confidence in leadership",
                          "Health & morale")) %>%
  ggplot(aes(x = reorder(var.name, estimate), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(small.cat ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")
```

\begin{figure}[!ht]
\includegraphics{figures/civlib_phi_estimates.pdf}
\centering
\caption{$\phi$ estimates for items about civil liberties, confidence in leadership, health, morale, and social trust.}
\label{fig:phi_civlib}
\end{figure}

```{r racgenphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about race, gender, sex, sexuality, and abortion."}
clean.pestimates %>%
  filter(grepl("_c", var)==FALSE) %>%
  filter(small.cat %in% c("Gender & family", "Race & immigration", "Sex, sexuality & abortion")) %>%
  ggplot(aes(x = reorder(var.name, estimate), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(small.cat ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")

```

\begin{figure}[!ht]
\includegraphics{figures/racgen_phi_estimates.pdf}
\centering
\caption{$\phi$ estimates for items about race, gender, sex, sexuality, and abortion.}
\label{fig:phi_racgen}
\end{figure}

# Appendix B: Coarsened variable estimates

```{r loadcoarse}
load(here("data", "coarse_vars.Rdata"))

```

$\phi$ estimates for coarsened items are presented in Figures \ref{fig:coarse_relig} through \ref{fig:coarse_racgen}. If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree). If scales included a clear midpoint, we included that and coarsened responses on either side. For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median. If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.

```{r coarserelig, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about religion, subjective socioeconomic status, and morale." }
coarse_vars %>%
  filter(small.cat %in% c("Relig. Activity", "Relig. identity & beliefs",
                          "Social life", "Subjective SES", "Suicide",
                          "Health & morale")) %>%
  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.text.y = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")

```

\begin{figure}[t]
\includegraphics{figures/coarse_relig.pdf}
\centering
\caption{$\phi$ estimates for coarsened questions about religion, subjective socioeconomic status, and morale.}
\label{fig:coarse_relig}
\end{figure}

```{r coarsepoli, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about politics, government, sex and sexuality."}
coarse_vars %>%
  filter(small.cat %in% c("Guns, laws, crime & police", "Politics & government",
                          "Public spending", "Sex, sexuality & abortion")) %>%
  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.text.y = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")
```

\begin{figure}[t]
\includegraphics{figures/coarse_poli.pdf}
\centering
\caption{$\phi$ estimates for coarsened questions about politics, government, sex and sexuality.}
\label{fig:coarse_poli}
\end{figure}

```{r coarseracgen, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about race and gender."}
coarse_vars %>%
  filter(small.cat %in% c("Gender & family", "Race & immigration")) %>%
  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
  geom_point() + 
  coord_flip() + 
  theme_minimal() + 
  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
  scale_color_manual(values = c("black", "darkgray")) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    strip.text.y = element_blank(),
    strip.placement = "outside",
    legend.position = "none",
    legend.title = element_blank(),
    axis.text=element_text(size=7),
    legend.text=element_text(size=7)
  )  +
  expand_limits(y = .9) +
  labs(x = "", y = "Phi estimate")
```

\begin{figure}[t]
\includegraphics{figures/coarse_racgen.pdf}
\centering
\caption{$\phi$ estimates for coarsened questions about race and gender.}
\label{fig:coarse_racgen}
\end{figure}

Generally speaking, most items with no evidence of active updating continue to have no evidence of active updating when coarsened. For items that showed evidence for active updating, the coarsened estimates tended to be very similar to the uncoarsened estimates, suggesting that persisting and non-persisting changes happen about as often between ends of these scales as they do within ends of these scales.

There are a couple notable departures from this general pattern. Several political questions -- general political views, views on whether the government should reduce inequality, and views on whether the government should help blacks -- show decreased evidence of persisting change when coarsened. This suggests that changes around the midpoint of the scale tend to be measurement error, a finding that is consistent with previous work suggesting that individuals without settled political views tend to choose points around the middle of the response scale [@converse].

\pagebreak

# References

### 2019-11-14 manuscript8.Rmd
finished background editing

@@
 ```{r setup, include=FALSE}
+# Before the first commit, I added "here", changed the load paths with here(), and added the %!in% function
+
 knitr::opts_chunk$set(echo = FALSE)
@@
 library(here)
+library(tidyverse)
 '%!in%' <- function(x,y)!('%in%'(x,y))
@@
 
-Attempts to account for opinion change at the societal level have produced an array of often conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent, real attitudes.
-Within sociology, a number of models suggest that individuals lack the cognitive tools to maintain consistent beliefs on social and political issues.
-As a result, they construct responses in the interview setting, drawing on ideas they hear from opinion leaders and changing their attitudes when elite discourse changes [@zaller; @converse; @perrinmcfadden].
-Cohort replacement theories of attitude change assume that individuals hold opinions, that people are unwilling to alter them in the face of societal change, and that public opinion changes in response to generational turnover [@mannheim; @ryder].
+Attempts to account for opinion changes in society have produced an array of often conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes.
+Some models suggest that individuals lack the cognitive tools to maintain consistent beliefs on social and political issues.
+As a result, they construct responses in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden].
+By contrast, cohort replacement theories of change assume that individuals hold opinions, that people are unwilling to alter them in the face of societal change, and that public opinion changes only with generational turnover [@mannheim; @ryder].
 Another broad array of models argue that individuals and the public at large hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
@@
 
-Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change that have their roots in theories of cultural sociology.
-The first of these individual-level models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation.
-This model is rooted in a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001).
+Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology.
+The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation.
+This model is rooted in a broadly pragmatist approach to  action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001).
 The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life.
-This model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another.
+This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
+Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???).
 Separate from this distinction is a second dimension of attitude formation that reflects degree to which individuals can consistently report the same opinion over time.
@@
 
-In this paper, we make these two dimensions of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for comparing the prevalence of active updating using panel data.
+In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for estimating the prevalence of active updating using panel data.
 We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
-By focusing on the pattern of attitude change at the individual level, we can elucidate the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change apply.
+By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change apply.
 Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
-We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary society.
+We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary U.S.
+society.
 
@@
 Second, settled opinions vary in how consistently individuals report the same answer.
-Consistent with theories that argue that individuals lack clear opinions, some survey items appear elicit inconsistent responses and a few appear to elicit random responses.
+Consistent with theories that argue that individuals lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses.
 At the same time, individuals are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
@@
 
-These findings generally are more consistent with the settled dispositions model than with the active updating model and support theories claiming that cultural change tends to come through generational turnover rather than persuasion.
-However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change and the life course.
+These findings are generally somewhat more consistent with the settled dispositions model than with the active updating model.
+This is, they offer broad support to theories claiming that cultural change tends to come through generational turnover rather than persuasion.
+However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change.
 The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions.
-We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.
+We argue that there is a place for both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
 
@@
 We will explain this in greater detail below.
-For now, consider the following two models:
+For now, consider the following two simple models:
 
@@
 As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses.
-Such considerations are temporary influences only.
+Such considerations are thus temporary influences only.
 
@@
 ## Consistency in Response
-Neither model makes any assumptions about the magnitude of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
-This means that we must consider separately the overall amount of change in responses from wave to wave and the persistence of this change.
+Neither model makes any assumptions about the magnitude of of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
+This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change.
 Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development.
-However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make much greater inference about the applicability of different theories to these items.
+However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.
 
@@
 We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses.
-Individuals might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their older positions.
+Individuals might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions.
 The second dimension reflects what we call stability, or how consistent individuals are in reporting their beliefs once we account for the amount of active updating.
@@
 
-In broad terms, cohort replacement theories of public opinion change suggest that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
-Implied in these models is the notion that individuals have a period of "cohortization" in which they form beliefs.
+In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
+Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.
 
@@
 Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender.
-Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort” change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts.
+Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts.
 Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].
@@
 These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse].
-For example, when asked whether the government should spend more on health care, forming an opinion requires an individual to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result.
-Most individuals do not have this information readily stored in their minds.
+For example, when asked whether the government should spend more on health care, forming an opinion requires someone to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result.
+Most people do not have this information readily stored in their minds.
 
 Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment.
-This explains why individuals are susceptible to question-order, priming, and recency effects [@zaller].
+This explains why survey respondents are susceptible to question-order, priming, and recency effects [@zaller].
 However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues.
@@
 However, we do spotlight one issue that saw significant change in elite opinion.
-In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of rights for homosexuals.
-Leaders of the Democratic Party including President Barack Obama came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
+In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights.
+Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
 For this issue, then, we might expect that public opinion shifts notably.
@@
 
-The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, individuals update their opinions across the life course in response to gathering new information and interacting with changing social environments.
+The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, individuals update their opinions across their lives in response to gathering new information and interacting with changing social environments.
 
 Some of these approaches, common in political science, suggest that individuals consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman].
-These models do not tend to make distinctions about what kinds of attitudes shift over time and what do not.
+These models do not tend to make distinctions about which kinds of attitudes shift over time and which do not.
 However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.
@@
 The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating.
-As such, this disagreement forms the primary structure of our analysis.
+This disagreement thus forms the primary structure of our analysis.
 A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development.
-We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating, but we should still observe high consistency and low active updating.
+We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating.
+However we should still observe high consistency and low active updating.
 

### 2019-11-15 manuscript8.Rmd
first complete pass

@@
 One approach theorizes that people actively update their beliefs in the face of new information.
-The other argues that, following early socialization experiences, individuals' dispositions are stable.
-We formalize these two models, elaborate empirical implications of each, and derive a simple model for comparing their prevalence using panel data.
+The other argues that, following early socialization experiences, dispositions are stable.
+We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data.
 We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey.
-The pattern of results is more consistent with the settled dispositions model than the active updating model, and most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes in belief.
+Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model.
+Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes.
 When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes.
-We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.
+We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
 ---
@@
 
-Attempts to account for opinion changes in society have produced an array of often conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes.
-Some models suggest that individuals lack the cognitive tools to maintain consistent beliefs on social and political issues.
-As a result, they construct responses in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden].
-By contrast, cohort replacement theories of change assume that individuals hold opinions, that people are unwilling to alter them in the face of societal change, and that public opinion changes only with generational turnover [@mannheim; @ryder].
-Another broad array of models argue that individuals and the public at large hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
+Attempts to account for opinion changes in society have produced conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes.
+Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues.
+As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden].
+By contrast, cohort replacement theories posit that people do hold opinions, are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder].
+Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
 xi; @achen1992].
-More recently, sociologists have suggested that individuals attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].
+More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].
 
@@
 Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???).
-Separate from this distinction is a second dimension of attitude formation that reflects degree to which individuals can consistently report the same opinion over time.
+Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time.
 Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller].
-Attitudes can also be in a process of active updating but relatively easy to report, such as when individuals change their registered partisan affiliation.
+Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.
 
@@
 We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
-By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change apply.
+By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change are more likely to apply.
 Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
@@
 Our analysis yields several results.
-First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change.
+First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change.
 Simply put, there is little evidence that U.S.
@@
 Second, settled opinions vary in how consistently individuals report the same answer.
-Consistent with theories that argue that individuals lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses.
-At the same time, individuals are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
-third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
+Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses.
+At the same time, people are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
+Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
 On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life.
-Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and individuals are more likely to report these attitudes consistently.
+Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently.
 This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.
@@
 That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time.
-The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within individuals these shifts will be erased over time as individuals return to their baselines.
+The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that these shifts will be erased over time as people return to their baselines.
 
@@
 We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses.
-Individuals might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions.
-The second dimension reflects what we call stability, or how consistent individuals are in reporting their beliefs once we account for the amount of active updating.
+People might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions.
+The second dimension reflects what we call stability, or how consistent people are in reporting their beliefs once we account for the amount of active updating.
 
@@
 At younger ages, however, we expect more volatility as attitude development occurs.
-We note, however, that there is some evidence that volatility of settled beliefs increases as individuals change as well [@krosnickalwin], but we do not expect this to be the general pattern.
+We note, however, that there is some evidence that volatility of settled beliefs increases as people change as well [@krosnickalwin], but we do not expect this to be the general pattern.
 
@@
 
-The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, individuals update their opinions across their lives in response to gathering new information and interacting with changing social environments.
-
-Some of these approaches, common in political science, suggest that individuals consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman].
+The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, people update their opinions across their lives in response to gathering new information and interacting with changing social environments.
+
+Some of these approaches, common in political science, suggest that people consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman].
 These models do not tend to make distinctions about which kinds of attitudes shift over time and which do not.
@@
 Other theories suggest that only some beliefs are open to updating during adulthood.
-A number of theories of attitude development posit that individuals hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein].
+A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein].
 These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams.
@@
 Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
-Any item where the burden of maintaining consistency is externalized and not in an individual's head might demonstrate both higher rates of active updating and greater consistency.
+Any item where the burden of maintaining consistency is externalized and not only "in a person's head" might demonstrate both higher rates of active updating and greater consistency.
 
@@
 
-The second challenge is measurement error, which is a form non-persisting change.
+The second challenge is measurement error, which is a form of non-persisting change.
 In the attitude change literature, measurement error means different things to different authors.
@@
 
-Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for AUM across variables as suggested by existing theories.
+Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
 Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be largely inductive.
@@
 To organize our presentation of such a large number of variables, we follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on their subject material.
-Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutional leaders, questions about public spending, and questions about social life.
+Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about public spending, and questions about social life.
 
@@
 
-Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergymen.
+Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy.
 We combine four parallel questions about how frequently individuals socialize to create a "social life" scale.
@@
 [^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options.
-These results are reported in an the Appendix.
+These results are reported in Appendix B.
 
@@
 
-Our model generates two parameter estimates of interest for each GSS item in our analysis: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
+Our model estimates two parameters of interest for each GSS item: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
 High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
-$\phi$ is our measure of active updating and captures the relative proportion of wave 3 explained by wave 2.
+$\phi$ is our measure of active updating and captures the relative proportion of wave 3 variance predicted by wave 2.
 If responses are generated through a settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
@@
 For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
-This pattern is consistent with the "increasing persistence" hypotheses.
+This pattern is consistent with the "increasing persistence" hypothesis.
 
@@
 There is no way we can possibly discuss all 184 items in detail without the discussion becoming tedious.
-Although we include $\phi$ estimates for all items in Appendix B, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
+Although we include $\phi$ estimates for all items in Appendix A, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
 We constrain items that showed no evidence of active updating to $\phi = .5$.
@@
 
-Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them, tapping the idea that public culture is more persistent than personal culture.
+Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them, perhaps making changes "stickier" than private forms of personal culture.
 If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
-This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, as well as how often individuals engage in "religious activities", which are more nebulous questions that display less active updating.
+This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, as well as how often individuals engage in "religious activities," which are more nebulous questions that display less active updating.
 Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
@@
 The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession).
-If real external change is happening, we should see exactly this sort of pattern.
+If the environment is changing we should see exactly this sort of pattern.
 
@@
 This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making.
-These two models of individual change are often implicit in analyses of belief change at the society level, but they are rarely examined empirically.
+These two models of individual change are often implicit in analyses of belief change at the society level, but they are rarely compared empirically.
 We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models.
@@
 Around 40 percent of all items show no evidence for updating, and even those items that do show a relatively low degree of updating.
-Adult Americans are highly consistent in their attitudes from year to year, and those who change tend to bounce back to where they were.
+Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were.
 What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
 Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is.
-But the average level of updating appears for most views appears to be low.
+But the average level of updating for most views appears to be low.
 
@@
 This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014).
-Because the GSS begins with 18-year-olds, we may lack power to detect updating that occurs even earlier, so this may be an underestimate.
+Because the GSS begins with 18-year-olds, we may lack power to detect updating that occurs even earlier, so this is probably an underestimate.
 This may be the case for the roughly 70 items that show no evidence of persistent change over time.
@@
 
-In the domain of cultural theory, our findings support the view that most cultural change happens slowly, through the mechanism of cohort succession.
-Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled during adulthood.
+In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly, through the mechanism of cohort succession.
+Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood.
 The settled nature of these beliefs is generally coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report these beliefs wave after wave.
@@
 In contrast to other beliefs, however, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging some people to update their views who might not otherwise do so.
-By definition, salience is a limited resource, meaning only few beliefs and behaviors could change at this rate at a time.
+By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given historical period.
 The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
@@
 While this age-related decline appears to be true for many political attitudes, which are the those items most frequently studied, the pattern is not nearly as consistent for other kinds of attitudes.
-Some items, such views on the Bible, suggest perpetual openness to attitude change as individuals age.
-Others, such as views on most abortion questions, suggest that early adult socialization is so strong that opinions are hardened by the time most people reach even 18 years old.
+Some items, such views on the Bible, suggest continued openness to attitude change as individuals age.
+Others, such as views on most abortion questions, suggest that early adult socialization is so strong that lifetime opinions are settled by the time most people reach even 18 years old.
 
@@
 Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is a strong start [@eatonetal; @vissermirabile].
-However, this line of work should be cognizant of which attitudes stabilize in middle age and the relationship between social factors and these specific attitudes and explore how these factors link to attitudes that seem to change more frequently at these ages.
+However, this line of work should be aware of which attitudes stabilize in middle age and the relationship between social factors and these specific attitudes and explore how these factors link to attitudes that seem to change more frequently at these ages.
 
@@
 
-We said at the outset that these two models of individual culture underlie significant debates about public opinion formation, the formation of individual political beliefs over the life course, and the ability of individuals to hold and express clear opinions over time.
+We said at the outset that these two models of personal culture underlie significant debates about public opinion formation, the formation of individual political beliefs over the life course, and the ability of individuals to hold and express clear opinions over time.
 What do our findings suggest for these debates?
@@
 This is clearest when it comes to questions that deal with spending priorities.
-Prior responses to questions about government spending on foreign aid, science, drugs, child care, crime prevention, mass transit, national parks, and the problems of cities are bad predictors of what individuals will say later on.
-
-That being said, individuals are not wholly without consistent opinions.
+Prior responses to questions about government spending on foreign aid, science, drugs, child care, crime prevention, mass transit, national parks, and the problems of cities are bad predictors of what people will say later on.
+
+That said, respondents are not wholly without consistent opinions.
 On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next.
@@
 
-Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of younger individuals to understand how these attitudes are formed.
+Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.
 Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful.

### 2019-12-06 manuscript9.Rmd
kevin's revisions

--- 2019-11-15 manuscript8.Rmd
@@
 ---
-title: "Measuring Stability and Change in Personal Culture Using Panel Data"
+title: "Stability and Change in Personal Culture"
 author:
@@
 
-Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology.
+Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology.
 The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation.
@@
 This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???).
-Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time.
-Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller].
-Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.
-
-In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for estimating the prevalence of active updating using panel data.
+
+In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data.
+In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors.
 We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
-By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change are more likely to apply.
+By classifying the pattern of change at the individual level, we can clarify when different accounts of aggregate change are more likely to apply.
 Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
-We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary U.S.
+We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S.
 society.
@@
 First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change.
-Simply put, there is little evidence that U.S.
+Simply put, there is little evidence that large numbers of U.S.
 adults are changing their beliefs or practices in lasting ways over this period of study.
 Second, settled opinions vary in how consistently individuals report the same answer.
-Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses.
-At the same time, people are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
+Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses.
+At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
 Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
@@
 
-These findings are generally somewhat more consistent with the settled dispositions model than with the active updating model.
-This is, they offer broad support to theories claiming that cultural change tends to come through generational turnover rather than persuasion.
-However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change.
-The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions.
-We argue that there is a place for both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
+Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence.
+However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course.
+The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to specific policy positions.
+We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
 
@@
 
-These models may seem similar at first glance, but they have very different implications for the pattern of individuals' responses over time.
+These models may seem similar at first glance, but they have different implications for the pattern of individuals' responses we would observe over time.
 Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below.
 Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences.
-In the next two sections, we consider each of these models and briefly discuss their connections to sociological theory.
-In the section after that, we discuss the conditions under which each model is more likely to apply.
+In the next two sections, we consider each of these models and briefly discuss their connections to sociological theories.
 
@@
 The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$).
-There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.
+There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.
 Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state.
-This formal property, and this active updating model generally, is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings].
-Concerns about panel conditioning, in which a person's participation in a survey influences her subsequent answer choices, either by changing her non-survey behavior or by altering how they respond to the survey without changing behavior, also imply some sort of active updating on the part of the respondent [@warrenhalpernmanners].
+This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].
 
 More informally, this AUM posits an agent that is updating her views in the face of social experience.
-There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
+There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
 Following Gross's [-@gross09, pp.
 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."
+
+The concept of active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
+When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model.
+When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time.
+Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.
 
@@
 This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.
+
+A number of theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the population is becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
+In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.
 
@@
 Such considerations are thus temporary influences only.
-
-The SDM is most consistent with the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp.
+At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp.
 66ff].
 In other words, updating happened, but it happened in the past, prior to the time of the study.
-In this sense, $U_i$ reflects the "habitus." We will say more about this below.
+In this sense, $U_i$ reflects the "habitus."
+
+The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007].
+The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
+These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965).
+Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.
+
+On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case.
+The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
+These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have (Zaller 1992; Converse 1964), and suggest that peoples’ responses at any particular time are a random deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
+As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if these wave-to-wave differences are quite large.
 
@@
 
-Metaphorically, one can think of the active updating model as a "traveler" and the settled dispositions model as a "visitor." That is, in the active updating model, the traveler wakes up in the morning, walks in a direction for a while, goes to sleep, and then wakes up the next day to pick up the journey exactly where she left off.
-In the settled dispositions model, the visitor also travels in a particular direction each day but goes home at the end of the day and wakes up in the same bed each morning.
+Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level.
+For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported.
+Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].
 
@@
 ## Consistency in Response
-Neither model makes any assumptions about the magnitude of of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
+
+Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
 This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change.
-Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development.
+Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development.
 However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.
-
-## Combining the Models
-
-These two dimensions of attitude change give us a two-by-two matrix of patterns for attitude change.
-The first dimension reflects what we call active updating, or the idea that more recent responses are better predictors of current responses that older responses.
-In other words, active updating means that when changes happen, they are more likely to persist than revert.
-We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses.
-People might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions.
-The second dimension reflects what we call stability, or how consistent people are in reporting their beliefs once we account for the amount of active updating.
-
-With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them.
-Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix.
-We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes.
-Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.
-
-\begin{center}
-\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
-\hline
- & Low Active Updating & High Active Updating \\ 
-\hline
-High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\
-\hline
-Low Consistency &  \\ 
-\hline
-\end{tabular}
-\end{center}
-
-### Age and Attitude Formation
-
-In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
-Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.
-
-To continue the metaphor above, cohort replacement theories imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition).
-The concept of a "cohort effect" is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman].
-Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo].
-To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
-
-There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick].
-Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].
-
-Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender.
-Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts.
-Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].
-
-As Table ## suggests, items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults in the population, but high active updating for young people.
-Because these approaches generally assume that beliefs are prominent in the minds of respondents, the implication is that these beliefs are reported with high consistency in adulthood.
-At younger ages, however, we expect more volatility as attitude development occurs.
-We note, however, that there is some evidence that volatility of settled beliefs increases as people change as well [@krosnickalwin], but we do not expect this to be the general pattern.
-
-### Inconsistent Opinions
-
-Not all cultural change occurs through cohort replacement processes, and there is abundant evidence of large-scale public opinion shifts that occur in a span of time too short to be driven entirely by cohort replacement.
-This suggests that, at times, even middle-aged individuals update their opinions in persisting ways.
-
-In contrast to cohort replacement theories, a range of theories posit that, outside of a well informed elite, individuals lack consistent opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse].
-For example, when asked whether the government should spend more on health care, forming an opinion requires someone to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result.
-Most people do not have this information readily stored in their minds.
-
-Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment.
-This explains why survey respondents are susceptible to question-order, priming, and recency effects [@zaller].
-However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues.
-Shifts in public opinion can also come about when elites shift their opinions on issues, providing new signals to the general public.
-
-In general, then, we expect low-profile and complex public policy issues to show a low rate of active updating and a low rate of consistency.
-By definition, most questions of public policy should fall into this group, since only a few issues can sustain a relatively high level of public attention.
-The small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these.
-However, we do spotlight one issue that saw significant change in elite opinion.
+As such, we designate a second term, _consistency_, to reflect this second dimension of change.
+
+## Toward Theoretical Synthesis
+
+We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating.
+Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.
+
+The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern.
+This disagreement thus forms the primary structure of our analysis.
+However, it is possible that some items display active updating while others display settled dispositions.
+There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.
+
+_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman].
+To put it differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
+
+Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].
+
+_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein].
+Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe].
+In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint.
+Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time.
+They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
+442].
+This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate a highest level.
+
+These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate a cascade of changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs.
+In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating.
+Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions.
+Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.
+
+_Public vs.
+Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
+The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp.
+91].
+This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances.
+However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].
+
+Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse].
+This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these.
+However, we do spotlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population.
 In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights.
 Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
-For this issue, then, we might expect that public opinion shifts notably.
-This should manifest as a high degree of active updating for the whole population, not just young people.
-
-Outside of a lack of information on public policy issues, there are two other reasons why individuals might demonstrate low consistency and low active updating.
-First, individuals might be unwilling to report socially undesirable beliefs if they hold them, and as a result obfuscate their true position.
-Similarly, some issues might not require a lot of information to comprehend, but might be so far from the public consciousness as to be nonsensical to respondents.
-We expect both of these conditions to produce a similar pattern of low active updating and low consistency.
-
-### Updating Opinions
-
-The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, people update their opinions across their lives in response to gathering new information and interacting with changing social environments.
-
-Some of these approaches, common in political science, suggest that people consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman].
-These models do not tend to make distinctions about which kinds of attitudes shift over time and which do not.
-However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.
-
-Other theories suggest that only some beliefs are open to updating during adulthood.
-A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein].
-These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams.
-In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find little support for Lakoff's model and significant support for ideological identification as this "core" belief in the political space.
-If these individuals have the ability to regularly assess new information against these core beliefs, they should also be able to recognize their position in a multiple-choice survey questionnaire and report them with limited error [@vaisey09].
-In other words, we expect that core beliefs such as ideological identification should have high consistency and low active updating, and conversely that items that show these traits to have the potential to be core beliefs.
-
-In a similar vein of research focused on political polarization, Baldassari and Gelman [@baldassarigelman] find little increase in the correlation between specific policy positions over time, suggesting limited active updating of these items.
-What they do find is an increased alignment between partisan identification and policy positions.
-They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
-442].
-This suggests that among items tapping issues of politics, partisan affiliation should demonstrate the highest level of active updating, and that it should be reported with a high degree of consistency.
-
-### Public vs. Private Culture
-
-A final dimension that likely influences the consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
-The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp.
-91].
-This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances, which might lead to a low degree of consistency and active updating.
-However, when public culture provides a strong "scaffold" for individuals through clear signals of how their identities and social locations should influence how they should respond to a survey question, it becomes easier for them to maintain consistency [@lizardostrand].
-
-We have touched on this idea already with our discussion of the idea that high-profile elite opinion change might drive opinion change in the broader public, or by noting that high-profile issues might be easier to report consistently because individuals receive signals that support that position.
-Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
-Any item where the burden of maintaining consistency is externalized and not only "in a person's head" might demonstrate both higher rates of active updating and greater consistency.
-
-### Toward Theoretical Synthesis
-
-The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating.
-This disagreement thus forms the primary structure of our analysis.
-A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development.
-We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating.
-However we should still observe high consistency and low active updating.
-
-If what we find is a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
-Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
-We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
+For this issue, we predict active updating throughout the population, not just young people.
+
+Items that tap some public dimension of behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
+Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
+
+As we noted previously, we do not expect that all items will align with just one model.
+If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
+Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
+We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
 
@@
 
-Third, _are there systematic differences in item content between questions that fall in different parts of our matrix?_ The preceding sections have made some specific predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs.
+Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors.
 As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.
@@
 
-That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response.
-If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.
+That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0).
+If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.
 
@@
 That is, the SDM predicts that the average response of previous waves is the best predictor of the next response.
-If change is non-persisting (i.e., if our metaphorical visitor eventually always returns home), then taking the average of the last two places we saw her will be our best guess about the location of her home base.
+If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.
 
@@
 
+### Comparison to Other Approaches
+
+Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
+Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS survey responses over time.
+This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
+These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier.
+While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves.
+This approach also combines persisting and non-persisting change into two similar but distinct kinds of change "structural" and "non-structural," both of which can be persisting and non-persisting.
+This distinction, while important for some theoretical questions, is not our focus.
+
+A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief.
+This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
+Since we at times invoke both these models, we include a handful of composite scales (discussed below) of related items.
+If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in persistence.
+
+Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
+While Warren and Halpern-Manners [@-warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
+One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
+If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.
+
+We do not view this as a problem for our theoretical models.
+If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
+This might lead to a higher estimate of active updating than we might observe in the population, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update their beliefs over time.
+
+A second form of panel conditioning posits that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time, even if they actually change.
+In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change.
+This would be problematic for our study, as it would under-estimate the amount of real change in the population.
+However, if this is the case, we should observe no active updating in responses and high levels of consistency, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.
+
+Finally, a number of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
+These tools are well designed to address the questions they set out to.
+However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
+
 ### Limitations of the Method
 
-Three challenges limit our ability to evaluate the relative presence of settled dispositions and active updating models using our approach.
-These challenges are not fatal, but put some limitations on the conclusions we will be able to reach.
-
-The first challenge is that our model is designed to allocate variance explained to each of the prior predictors rather than to assign probabilities to each data-generating process per se.
+Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach.
+These challenges put some limitations on the conclusions we can reach.
+
+The first is that our model is designed to allocate variance explained to each of the prior predictors rather than to assign probabilities to each data-generating process.
 Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.
-If we were willing to assume that persisting and non-persisting changes were drawn from the same distribution, we could make inferences about the relative proportion of people coming from each data generating process, but we have no reason to believe this assumption.
-
-The second challenge is measurement error, which is a form of non-persisting change.
+
+The second is measurement error, which is a form of non-persisting change.
 In the attitude change literature, measurement error means different things to different authors.
@@
 For other theories, measurement error reflects individuals' inability to accurately report their response.
-It is also possible that measurement error simply reflects errors of selection and interpretation (misunderstanding the question or incorrect coding).
-
-Because measurement error "looks the same" as non-persisting real change (and because the latter is sometimes interpreted as the former), estimates of $\phi$ will be biased toward .5.
+It is also possible that measurement error reflects errors of selection and interpretation, such as misunderstanding the question or incorrect coding.
+
+Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), estimates of $\phi$ will be biased toward .5, since wave 1 would be informative for all individuals who do change.
 There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
-On the other hand, these previous studies of measurement reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them completely.
-Because of this issue of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
-
-The third challenge of our analysis is that we focus exclusively on predicting wave 3.
-If for some reason individuals have a high likelihood of changing between waves 2 and 3, then our ability to predict responses at wave 3 with either wave will be limited and $\beta$ will be low.
+On the other hand, previous studies of measurement reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
+Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
+
+The third challenge of our analysis is that we focus on predicting wave 3.
+If individuals have a high likelihood of changing between waves 2 and 3, our ability to predict responses at wave 3 will be limited and $\beta$ will be low.
 Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3.
-If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error.
-Then again, it may not be reasonable to consider this sort of change "persistent" in any meaningful way.
+If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error, and it may not be reasonable to consider this sort of change "persistent."
 
 In addition to these three challenges, there are two forms of change that our model is not well designed to account for.
-The first is change in the overall population.
+The first is a uni-directional shock to the population.
 Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates.
@@
 
-Second, for variables that do show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is disproportionately concentrated among younger respondents.
+Second, for variables that show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is concentrated among younger respondents.
 To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff.
@@
 We want to test our model on as broad a range of items tapping attitudes, beliefs, and social behaviors as possible.
-To do so, we estimate the model in equation \@ref(eq:MODEL) on data from the three different three-wave panel data sets of the General Social Survey, conducted from 2006 through 2014.
-
-We sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves.
-These questions tended to come from the "core" of the GSS, a set of questions asked in each wave.
-Rotating topical modules asked during the panels were only asked in select waves or were not asked consistently to the same people over time, leaving too few cases with complete data to analyze.
-
-Since our theoretical framework focuses on attitudes, beliefs, and social behaviors, we excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
-We also exclude questions that ask about an individuals' childhood.
-While testing these questions with our method is possible, they are beyond the scope of our theoretical framework.
-To organize our presentation of such a large number of variables, we follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on their subject material.
+We sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
+We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
+We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on their subject material.
 Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about public spending, and questions about social life.
 
-We also follow Hout and Hastings [-@houthastings] in recreating some commonly used scales designed to capture attitudes about gender roles, access to abortion, and social trust.
+We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust.
 This includes Rossi's six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
@@
 We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman].
-
 Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy.
@@
 Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3.
-This is a highly unlikely scenario.
 
@@
 To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates.
-Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2, and only 147 (57 percent) of those individuals maintained the same response at wave 3.
-Under the settled disposition model, we would expect that about 50 percent or 129 individuals would maintain the same response, so we really only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting attitude change.
-In other words, even for the items that do show strong evidence of active updating, the overall rate of attitude change in the population is small.
+Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2.
+Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3.
+Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3.
+Therefore, we only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting change.
+In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.
 
 The right side of Figure \@ref(fig:phihist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
-That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they will tend to revert to their previously held position.
-This group includes many items addressing abortion, civil liberties, confidence in institutions, and views on race and gender.
+That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
+This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.
 
@@
 To answer our first question, however, we need only focus on the overall distribution.
-We find that about 40 percent of items show no evidence at all for active updating among GSS respondents.
-And among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent changes.
+Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent change.
 We can only be really confident in detecting persistent change among a small minority of items, perhaps 1 in 5.
@@
 
-Our second research question is about whether there is evidence of cohort formation on some items.
-That is, we want to see if there is evidence that younger respondents are updating their views more than older respondents.
-Although it's impossible for any item to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
+Our second research question asks whether there is evidence that younger respondents are updating their views more than older respondents.
+Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
 
@@
 
-For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address pass discrimination, all evidence of active updating disappears for individuals over 30.
-This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but that over time they harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
+For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address pass discrimination, all evidence of active updating disappears for people over 30.
+This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but beliefs quickly harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
 For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
-This pattern is consistent with the "increasing persistence" hypothesis.
+This pattern is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
 
 Eight of these items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
-These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, views on whether aging parents should live with their children, and views on whether police should be allowed to use force against suspects who are verbally abusive toward them.
-Some of these items, such as views on whether aging parents should live with their children, might be things young people are not forced to consider until later in life and as a result do not form clear opinions until that exposure.
-This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is relatively unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previous theorizing has accounted for.
-
-The remaining 89 items explored in this analysis (just under half of all items) show evidence for some active updating but do not show consistent evidence for age heterogeneity, again suggesting a more complicated relationship between age and attitude change than previously theorized.
+These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, and views on whether aging parents should live with their children.
+Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on early in life.
+This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
+
+The remaining 89 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
 However, this does not mean these show strong evidence of an active updating model.
@@
 
-While there is a high proportion (40 percent) of items that show no evidence of active updating, and those that do show evidence show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
+While there is a high proportion (40 percent) of items that show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
 Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
 
-Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items colored by whether they preferred the $\phi = .5$ constraint or not.
-Although we discuss the pattern of items below, we label a few items that stand out.
+Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items shaded by whether they preferred the $\phi = .5$ constraint or not and label a few items that stand out.
 Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.
@@
 
-Consistent with the findings and theory of Vaisey and Lizardo [-@vaiseylizardo], we see that over half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
+Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that over half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
 These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
@@
 
-There are several exceptions to this general pattern, however, even in categories with otherwise very low $\phi$ values.
+There are several exceptions to this general pattern, however, even in categories with otherwise low $\phi$ values.
 The items with the largest values generally have one or more attributes in common.
-We consider these attributes one by one in order to give some general impressions of the pattern.
-
-Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them, perhaps making changes "stickier" than private forms of personal culture.
+We consider these attributes to give some general impressions of the pattern.
+
+Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them, perhaps making changes "stickier" than forms of personal culture.
 If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
-This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, as well as how often individuals engage in "religious activities," which are more nebulous questions that display less active updating.
+This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, which are more nebulous questions that display less active updating.
 Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
 Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
+Questions about religious activity, identity and (to a lesser extent) religious beliefs generally show some evidence for active updating.
+In some cases, this fits with the general pattern of external mechanisms (e.g., attendance and identification are largely public acts).
 
@@
 The president changed between the 2008 and 2010 waves of the GSS, meaning that the question no longer referred to the same administration.
-If we generate separate estimates for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, we see that it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$).
+If we generate estimates for this item for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$).
 In the 2006-10 panel, waves 1 and 2 have almost no predictive power ($\beta = .10$).
@@
 
-Likewise, all of public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
+Likewise, all public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
 The change of administration and changing federal spending policies likely affected all spending items.
@@
 
-Perhaps the most striking pattern in our findings is about questions about _gay rights_.
+Perhaps the most striking pattern in our findings pertains to questions about _gay rights_.
 There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating.
@@
 The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
-By comparison, abortion attitudes show almost no evidence for change.
-
-Finally, questions about religious activity, identity and (to a lesser extent) religious beliefs generally show some evidence for active updating.
-In some cases, this fits with the general pattern of external mechanisms (e.g., attendance and identification are largely public acts).
+This pattern is strongly consistent with Zaller's [@-zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
+
+A few items lack external mechanisms, represent forms of private culture, and did not achieve salience during this period, but still display active updating, including some views on abortion and the morality of different forms of sex.
+We address these items in greater detail in the discussion section.
 
@@
 
-Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, we examine these beliefs in greater detail than other items.
-Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for most of the questions in our data set tapping questions of political identity, the role of government, and specific policies, broken into four general areas.
-We break these items up into categories and remove a few items solely for ease of viewing.
+Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, and because many of these beliefs display evidence of active updating, we examine these beliefs in greater detail than other items.
+Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for questions about political identity, the role of government, and specific policies.
+We break these items up into categories and remove a few items for ease of viewing.
 
@@
 There are several notable features of the figure.
-First, political party affiliation is a clear outlier in the plot, with much greater consistency and much greater active updating than other items.
-In contrast, questions asking about specific public spending priorities tend to have very low consistency, at least compared to other political items.
+First, political party affiliation is a clear outlier, with greater consistency and greater active updating than other items.
+In contrast, questions asking about specific public spending priorities tend to have low consistency compared to other items.
 Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government –  display higher consistency than specific policy questions but weak evidence of active updating.
-It is important to note that for political views, however, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
-
-While individuals tend to be inconsistent in their responses to spending priorities, views on some other specific policy questions with both high and low salience are reported quite consistently.
-Views on abortion, especially when considered as a scale, are quite consistently reported.
-As with all other scales, aggregating the composite items increases consistency but does not affect the amount of active updating.
+It is important to restate that for political views, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
+
+While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently.
+Views on abortion, especially when considered as a scale, are consistently reported.
+As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change.
 As discussed previously, views on gay marriage display high levels of active updating.
 In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating.
-This is notable, as public opinion and policies regarding both have shifted considerably in recent years, with many people assuming change in both is being driven by the same underlying process.
-Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but change in views on marijuana use have been driven primarily by cohort replacement.
+This is notable, as public opinion and policies regarding both have shifted considerably in recent years, with many people assuming change in both is being driven by the same underlying process [@schnabelsevell].
+Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana use have been driven primarily by cohort replacement.
 
@@
 This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making.
-These two models of individual change are often implicit in analyses of belief change at the society level, but they are rarely compared empirically.
+These two models of individual change are implicit in analyses of belief change at the society level, but they are rarely compared empirically.
 We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models.
@@
 First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
-Around 40 percent of all items show no evidence for updating, and even those items that do show a relatively low degree of updating.
+Around 40 percent of all items show no evidence for updating.
+For the items that do show evidence of active updating, the overall rate of change in the population is low.
 Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were.
@@
 Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
+Items assumed to be "peripheral" beliefs, such as specific policy questions, also showed evidence for updating and low consistency.
 Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18.
+
+There are a handful of items that demonstrate active updating and enough consistency to merit further consideration, but are not easily explained by theoretical mechanisms outlined so far.
+These include questions about abortion in the case of rape or poverty, the morality of premarital and teen sex, the ideal number of children in a family, and whether children should obey adults or think for themselves.
+These items did not achieve high salience during the time of the study and there were not, to the best of our knowledge, clear changes in elite opinion on these issues.
+This suggests additional mechanisms that can drive persisting change at the individual level, such as exposure to these ideas through mass media or by directly interacting with people in these groups.
+In the case of questions about children and family structures, it might be the experience of having children that leads to changes in these attitudes.
+It is important to note that even though these items demonstrate active updating, it is a small proportion of the population that are changing their views during the study window.
+
 The overall pattern of results, while diverse and at times hard to reconcile, does shed light on theoretical debates, to which we now turn.
@@
 
-What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area? We see two major implications, one theoretical and one methodological.
+What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area?
 
@@
 Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood.
-The settled nature of these beliefs is generally coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report these beliefs wave after wave.
-In contrast, a range of views about race were so inconsistent that it would be difficult to call them either settled or updating.
+The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time.
+In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating.
 Even for those items that did display strong evidence for active updating (e.g.
-"whites rich or poor" and "whites work hard") were still extremely hard to predict wave after wave.
-
-In contrast to other beliefs, however, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging some people to update their views who might not otherwise do so.
-By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given historical period.
+"whites rich or poor" and "whites work hard") were still hard to predict wave after wave.
+
+In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
+By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period.
 The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
@@
 An additional view, the "life stages" hypothesis, suggests a similar pattern with a surge in attitude change late in life [@visserkrosnick].
-While this age-related decline appears to be true for many political attitudes, which are the those items most frequently studied, the pattern is not nearly as consistent for other kinds of attitudes.
-Some items, such views on the Bible, suggest continued openness to attitude change as individuals age.
+While this age-related decline appears to be true for many political attitudes, the pattern is not nearly as consistent for other kinds of attitudes.
+Some items, such as views on the Bible, suggest equal openness to attitude change as individuals age.
 Others, such as views on most abortion questions, suggest that early adult socialization is so strong that lifetime opinions are settled by the time most people reach even 18 years old.
 
-For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change actually becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
+For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
 Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to factors that encourage openness to change at different ages.
@@
 Doing so will expand our understanding of the social and psychological factors that that give rise to stable or variable attitudes [@howekrosnick].
-
 Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is a strong start [@eatonetal; @vissermirabile].
-However, this line of work should be aware of which attitudes stabilize in middle age and the relationship between social factors and these specific attitudes and explore how these factors link to attitudes that seem to change more frequently at these ages.
 
@@
 
-We said at the outset that these two models of personal culture underlie significant debates about public opinion formation, the formation of individual political beliefs over the life course, and the ability of individuals to hold and express clear opinions over time.
-What do our findings suggest for these debates?
-
 A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
-These respondents might express different positions from wave-to-wave, but in guessing what an individual will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
+These respondents might express different positions from wave to wave, but in guessing what an individual will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
 While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
@@
 
-The fact that there is not significant active updating for most political opinions should not be taken to imply that individuals are consistent in their positions.
-Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions.
-This is clearest when it comes to questions that deal with spending priorities.
-Prior responses to questions about government spending on foreign aid, science, drugs, child care, crime prevention, mass transit, national parks, and the problems of cities are bad predictors of what people will say later on.
+Similarly, our findings are not consistent with the idea that individuals adopt a partisan affiliation based on their social groupings and subsequently adjust their ideological commitments to conform to that [@greenetal].
+For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much persistent change as partisan identification, which is simply not true in our data.
+
+Our work cannot speak to the strength of ideological commitments or whether individuals have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions.
+The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
+What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."
 
@@
 
+We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation.
+One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.
+
+We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
+We see too much inconsistency in responses to most questions for models of commitment bias to be applicable.
+Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
+499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p.
+500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p.
+501], tend to show low active updating and low consistency.
+
+We cannot and would not want to rule out the possibility that panel conditioning is taking place in the GSS, and we believe it would be worthwhile to explore these same GSS panels for evidence of panel conditioning.
+However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
+
 Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.
@@
 In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
-Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike as as particularly useless.
+Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly useless.
 Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).

### 2019-12-06 old_drafts/manuscript6.Rmd
kevin's revisions

---
title: "(Almost) Nobody Changes"
subtitle: "Comparing Theoretical Models of Cultural Change Using Panel Data"
author:
  - Kevin Kiley
  - Stephen Vaisey
date: "Duke University"
output: pdf_document
header-includes:
     - \usepackage{longtable}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dagitty)
```

# Introduction

Culture is an important part of social life but cultures are continuously evolving. In 1972, for example, over 40% of US adults supported a law outlawing interracial marriage. Three decades later, this opinion had become so uncommon that the question was removed from the US General Social Survey. How does this kind of cultural change happen?

Although rarely made explicit, sociological models of cultural change generally fall into two broad classes. The first is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions. In this model, people continuously revise their views and practices based on the people, discourse, and events around them. The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. In this model, adults' views and practices are relatively robust to changes in the social environment.

Both accounts have affinities with traditions in sociological theory. The active updating model is most compatible with a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas, Gross). The settled dispositions model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu, Vaisey and Lizardo)

In this paper, we make these two core models explicit, deduce some of their empirical implications, and derive a simple strategy for comparing them using panel data. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions (like views on politics, free speech, race, and gender roles) and practices (like socializing at at bars or attending church) that are important in contemporary society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in belief or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. That is, on several items at least, it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Third, we find that changes in social behavior (e.g., church attendance, party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology). This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

These findings are generally more consistent with the settled dispositions model than with the active updating model. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change. We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.

In addition to their theoretical significance, our findings also have practical methodological implications. We argue that short duration panel studies of adults will be of limited utility unless they have a clear focus on the sorts of things that are likely to change in persistent ways over the study period. Otherwise, researchers will waste resources measuring "change" that turns out to be noise. In general terms, then, we suggest that social scientists would do well to conduct more panel studies of children and adolescents while using cross-sectional designs to study adults.

# Theoretical Background

How does cultural change happen? Let us make the question concrete by imagining a person who answers the same question each year for several years. The question could be anything, but assume it is this GSS question: "Please tell me whether you strongly agree, agree, disagree, or strongly disagree [with this statement]: 'a working mother can establish just as warm and secure a relationship with her children as a mother who does not work.'"

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all. We will explain this in greater detail below. For now, consider the following two models:

\begin{equation}
  \label{eq: AR}
  y_{it} = y_{it-1} + \nu_{it}
\end{equation}
\begin{equation}
  \label{eq: FE}
  y_{it} = \mu_i + \nu_{it}
\end{equation}

These models may seem similar at first glance, but they have very different implications. Equation 1 represents an _active updating_ model and Equation 2 represents a _settled dispositions_ model, as we explain below. Figure 1 (XXX) shows these models in graphical form, which helps highlight their differences. In the next two sections, we consider each of these models and briefly discuss their connections to sociological theory. In the section after that, we discuss the conditions under which each model is more likely to apply.

## Active Updating Model

Equation 1 represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's (2009: 367) pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities (Gross, DeGloma). As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.

The AUM makes no assumption about the distribution of $\nu_{it}$. Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$. This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.

## Settled Dispositions Model

Equation 2 represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($\mu_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are temporary influences only.

The SDM is most consistent with the Bourdieusian model of action that emphasizes "the past conditions of production" (XXX:XXX). In other words, updating happened, but it happened in the past, prior to the time of the study. In this sense, $\mu_i$ reflects the "habitus." We will say more about this below.

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within individuals these shifts will be erased over time as individuals return to their baselines.

Metaphorically, one can think of these models as travelers. In the active updating model, the traveler wakes up in the morning, walks in a random direction for a while, goes to sleep, and then wakes up the next day to pick up the journey where she left off. In the settled dispositions model, the traveler also travels in a random direction each day but wakes up to find herself in the same bed every morning.

The models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline. In the SDM, changes tend to revert because each person returns to their existing baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_. We consider the implications of these predictions below.

## Combining the Models

Although our presentation above highlights differences between these models, they are not necessarily contradictory. There are at least two ways to understand how both models could apply without contradiction.

### Age and Cohort Formation

The easiest way to reconcile the AUM and SDM is to posit that they operate sequentially over a person's life. To continue the metaphor above, one can imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition). This is, in fact, the logic that underlies the notion of a "cohort effect." The idea of a cohort effect is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life (Elder, Bartels and Jackman). Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative (Vaisey and Lizardo). To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

This discussion of cohorts brings us close to the vast literature on age-period-cohort decomposition models, which is focused on statistical techniques to partition variance among sources of social change in repeated cross-sectional data (e.g., Yang and Land). Although this work is important, it is focused on modeling changes in population aggregates whereas we are concerned with modeling cultural change from the point of view of the same actor moving through time.

The two are related and have implications for each other, of course. Again, the idea of a cohort requires that some people (i.e., younger people) are influenced by the environment to make persistent change in ways that others (i.e., older people) are not. This implies a transition, perhaps a gradual one, from active updating to a settled disposition. A further connection between the two is that the world implied by the SDM can only change in the aggregate due to cohort replacement. That is, if persisting change is extremely rare among adults, the only way for persisting change to occur in the population is for older cohorts to die and be replaced with younger cohorts with (perhaps) different baselines. For our purposes, the most important idea here is that "cohortization" implies that the AUM might describe the data-generating process better among younger respondents whereas the SDM might describe it better among older respondents.^[We do not have space for a full discussion here, but we regard these approaches as complementary. Repeated cross-section approaches must make strong assumptions about individual processes but are able to use many years of data. A panel approach is better able to capture micro-level processes but at the cost of using only a few years of data; this requires making strong assumptions about the generalizability of the process across historical time.]

### Different Items, Different Processes

Another way to reconcile the AUM and the SDM is to consider that they apply to different sorts of questions a researcher might ask a respondent. We can illustrate this idea with (non-cultural) examples we understand well and then consider how less obvious cultural processes might be similar or different.

First, if we repeatedly ask a sample of women "how many children have you ever given birth to?", we know the data is generated by a process that looks like equation 1. Leaving measurement error aside, the response is a simple combination of how many children the respondent had last time we asked ($y_{it-1}$) and whether or not she has had any since ($\nu_{it}$).

Second, if we ask repeatedly ask a sample of adults "how many full siblings did you have at age 16?", we know the true answer cannot change and therefore equation 2 is the best model. This is because $\mu_i$ refers to a fixed quantity and cannot change over time. Responses could still vary due to measurement error (e.g., misunderstanding the question) meaning that the variance of $\nu_{it}$ need not be 0. But the underlying reality is, by definition, a settled matter.

Finally, there are responses that have no stable component but are entirely functions of local circumstances. If we repeatedly ask a sample of people in the same city "did it rain yesterday?", their responses would not be described better by Equation 1 or Equation 2. In this case, everything that matters is in the contemporary circumstances ($\nu_{it}$) and thus the terms that distinguish equation 1 from equation 2 ($y_{it-1}$ or $\mu_i$) are ignorable. (These types of items are not important for adjudicating our research questions but we still need to be aware of this possibility.)

What we want to determine, then, is how much a particular item resembles these different processes. Consider repeated requests for (dis)agreement with the prompt "a working mother can establish just as warm and secure a relationship with her children as a mother who does not work." Does the aggregate pattern of responses look more like an active updating process where a person might be changing her mind? Or does it look more like asking a person to report repeatedly on a settled matter?

When it comes to beliefs and behaviors, which items will demonstrate persistence likely depends on a number of factors. [REWRITE SHORT -- 2 paragraphs max]

## Toward Theoretical Synthesis

The active updating model and the settled dispositions model are different and they do have affinities with different theoretical traditions in sociology. Indeed, the difference in emphasis between environmental cues, creative meaning-making, and public discourse on the one hand and settled dispositions, "habitus," and cohort replacement on the other has been a basic theoretical tension in cultural sociology for more than a decade (Vaisey, Lizardo, Vaisey and Lizardo, etc.).

One model may, in fact, be a better starting place than the other. Previous work using cross-sectional data suggests that settled dispositions are hugely important (Vaisey and Lizardo). Despite these findings, the evidence is still mixed. Therefore our goal here is neither to declare victory for one of the theoretical perspective nor to simply say that both "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types processes are at work. We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

# Research Questions and Expectations

With these considerations in mind, we ask the following questions:

First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects (Vaisey and Lizardo). This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ As we just mentioned, in a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18). However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.

Third, _are there systematic differences in item content between questions that are better described by each model?_ Items differ from each other in many ways and it would be challenging to come up with clear predictions for all possible types of items. In general, however, we expect (following Vaisey and Lizardo) that core beliefs about gender, family, race, and trust should be more consistent with settled dispositions because their formation should be already be complete for adults. Items that require public commitments such as changing political parties, socializing, or religious attendance and affiliation should be open to persistent change because there are social mechanisms (not just cognitive ones) to _maintain_ the change. [ONE MORE SENTENCE? MAKE SURE THE SENTENCES IN "DIFF ITEMS, DIFF PROCESSES" PRESAGE THIS DISCUSSION.]

# Analytic Strategy

To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2008-2010). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below.

## Statistical Model

### Basic Models

Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:

\begin{equation}
  \label{eq: AUM2}
  E(y_{i3}) = y_{i2}
\end{equation}

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response. If change is persisting (i.e., if our metaphorical traveler wakes up in the same bed she went to sleep in last night), then our best guess is that she will be close to where we saw her last.

The SDM makes the following prediction:

\begin{equation}
  \label{eq: SDM2}
  E(y_{i3}) = \mu_i
\end{equation}

Since the best estimate of $\mu_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction like this:

\begin{equation}
  \label{eq: SDM2B}
  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
\end{equation}

That is, the SDM predicts that the average response is the best predictor of the next response. If change is non-persisting (i.e., if our metaphorical traveler ends up back in her own bed every morning), then taking the average of the last two places we saw her will be our best guess about the location of her home base.

### Combined Model

Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor as well, and its is treated as equally predictive as $y_{i2}$. Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i1}$ carries any additional predictive power for $y_{i3}$ once we control for $y_{i2}$. If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe likely came from a settled dispositions model. However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$, we begin to observe evidence that at least some individuals in the population engage in active updating. We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:

\begin{equation}
  \label{eq:MODEL}
  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
\end{equation}

Rather than estimate separate parameters for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is some transformation of the mean of previous waves, consistent with Equation \eqref{eq: SDM2B}. In the active updating model is present in some respondents, then wave 2 will be more predictive than wave 1, and $\phi$ will increase converge with Equation \eqref{eq: AUM2} in certain circumstances.

### Limitations of the Method

Three challenges limit our ability to evaluate the relative presence of settled dispositions and active updating models in the data set using our approach. The first is that our model is designed to allocate variance explained to each wave, rather than assign probabilities to each data-generating process. Because of this, few large persisting changes can inflate the $phi$ estimate even if most individuals make small non-persisting changes. If we were willing to assume that persisting and non-persisting changes were drawn from the same distribution, we could make inferences about the relative proportion of people coming from each data generating process, but we have no reason to believe this assumption.

The second challenge is measurement error, which is by definition non-persisting change. As a result, estimates of $\phi$ will be biased toward .5 in the presence of measurement error. There is evidence that many of the items explored in our analysis are measured with significant error (Alwin, Hout and Hastings). On the other hand, previous studies of measurement error often conflate measurement error and non-persisting belief change, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them. Because of this issue of measurement error, it is unlikely that $\phi$ and $\beta$ will be one for any item.

The third challenge of our analysis is that we focus exclusively on predicting wave 3. If for some reason individuals have a high likelihood of changing between waves 2 and 3, then our ability to predict responses at wave 3 with either wave will be limited and $\beta$ will be low. Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes between waves 2 and 3. If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model.

In addition to these two limitations, there are two forms of change that our model is not well designed to account for. The first is overall mean change. Since our model includes an intercept, changes that shift all responses toward one end of the scale are not accounted for in our $\phi$ estimates. The second is change in the variance of responses. If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will not enter into $\phi$.

Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error. Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model. We can only detect whether there is evidence of persisting belief changes, and therefore whether there is any evidence that active updating is present.

## Analysis Steps

Our analysis proceeds in three steps. We first evaluate the evidence in favor of the Active Updating model. To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using \eqref{eq:MODEL} with a free estimate of $\phi$ to a model that constrains $\phi= .5$. If the model with the constraint if preferred, then there is little evidence that respondents engage in an active updating process with respect to that item during the window we study.

Next, for variables that do show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is disproportionately concentrated among younger respondents. To test this, we re-estimate our original model and include a dummy term indicating whether respondents fall below an age cutoff. Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45. We test a range of cutoffs to ensure robustness of the overall pattern to specific ages. If the cutoff is too restrictive, there might not be enough cases to draw valid inferences. If there are strong differential effects of age in a narrow age range, a large cutoff might dilute the effects of age on active updating.

Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for AUM across variables.

## Item Selection

We seek to test our model on as broad a range of items tapping attitudes, beliefs, and social behaviors as possible. To do so, we estimate the model in equation \ref{eq:MODEL} on data from the three three-wave panel data sets of the General Social Survey, conducted from 2006 through 2014.

We sought measures of attitudes, beliefs, self-assessments, self-perceptions, and behaviors that were asked in three waves. These questions tended to come from the "core" of the GSS, a set of questions asked in each wave. Rotating topical modules asked during the panels were only asked in select waves or were not asked consistently to the same people over time, leaving too few cases with complete data to analyze.

Since our theoretical framework focuses on attitudes, beliefs, and social behaviors, we excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the questioner. We also exclude questions that ask about an individuals' childhood. While testing these questions with our method is possible, they are beyond the scope of our theoretical framework. To organize our presentation of such a large number of variables, we follow Hout and Hastings (XXXX) and group questions into 15 categories based on their subject material. Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutional leaders, questions about public spending, and questions about social life.

We also follow Hout and Hastings (XXXX) in recreating some commonly used scales designed to capture attitudes about gender roles, access to abortion, and social trust. This includes Rossi's six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances (abany). We use Smith's (1997) scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are. We use four questions to create a scale of gender role attitudes (Cotter, Hermsen, and Vanneman 2011).

Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergymen. We combine four parallel questions about how frequently individuals socialize to create a "social life" scale. We combine four questions about support for suicide under different circumstances. We also created a scale of support for police use of violence against criminal suspects by averaging five binary questions about the conditions under which individuals support police use of violence.

In total, we test the model on 184 GSS items, including the composite scales.[^coarsened]

[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options. These results are reported in an appendix.

## Results

Our model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1. If responses are generated through a Settled Dispositions Model, then $\phi$ will be .5. As the evidence of active updating increases, $\phi$ will increase toward 1. Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3. This is a highly unlikely scenario.

## Evidence for Active Updating

To evaluate the evidence in favor of the Active Updating model, we compare for all 184 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$. If the model with the constraint if preferred, then there is little evidence that respondents engage in an active updating process with respect to that item.

Figure \ref{fig:phi_hist} plots the distribution of $\phi$ estimates for the 184 items evaluated in this analysis and the probabilities that the model without the constraint fits the data better.

\begin{figure}[t]
\includegraphics{change_figures/aum_prob_dist.pdf}
\centering
\caption{Distribution of probabilities that items show evidence of Active Updating.}
\label{fig:phi_hist}
\end{figure}

The right side of figure \ref{fig:phi_hist} shows that while the majority of items prefer the free parameter, 72 items (about 40 percent of those examined) prefer the constraint, meaning these items show no evidence of active updating during the time frame examined here. While individuals might shift their response to these items in any particular wave because of measurement error or real ideological change, they will tend to revert to their previously held position over time. On the left side of the figure, we see that the majority of $\phi$ estimates fall between .5 and .6, meaning that wave 2 is only a slightly better predictor than wave 1 for most items. This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.

The group of items preferring the constraint includes a broad swath of items across a number of subject areas. It includes about half of the items about civil liberties, half of the items about confidence in institutional leaders, half of all questions about race and immigration, half the items about gender and family, and half the items about sexuality and abortion. On the other hand, most items about religion, suicide, politics and government, and public spending show evidence of active updating.

Figure \ref{fig:phibeta_scatter} plots the $\phi$ and $\beta$ estimates for items colored by whether they prefer the constraint or not, with a few items labeled. Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power. Included in the first group, where waves 1 and 2 are both predictive, are items asking whether a person hunts (labeled in Figure \ref{fig:phibeta_scatter}), views about the morality of sexual relations among teenagers and extramarital sex, several questions tapping gender roles and gender discrimination, items capturing stereotypes about race, questions about civil liberties, and most questions and scales about abortion.

\begin{figure}[t]
\includegraphics{change_figures/phibeta_scatter.pdf}
\centering
\caption{Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint.}
\label{fig:phibeta_scatter}
\end{figure}

The second group of items, those for which neither wave 1 nor wave 2 are particularly predictive and as a result prefer the constraint of $\phi= .5$, comprises a set of questions about racial stereotypes, including whether blacks and whites are intelligent. This is consistent with previous work finding a high level of measurement error for these items (Hout and Hastings). This group also includes items strongly shaped by the Great Recession, such as whether individuals believe they will lose their jobs.

Items showing evidence for active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6. A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.

## Age Heterogeneity

While it's impossible to say what proportion of people are following each data-generating process for each item, it is possible within an item to compare whether evidence of active updating is disproportionately concentrated in younger respondents.

Of the 112 items that showed evidence for active updating, 23 of these showed differential effects of age for more than 50 percent of cutoff ages. To get a sense of the magnitude of difference between older and younger individuals, Figure \ref{fig:age_group_comparison} plots the estimates of these 23 items for individuals above or equal to or below 30.

\begin{figure}[t]
\includegraphics{change_figures/age_differences_phi.pdf}
\centering
\caption{Comparison of phi estimates for individuals over and equal to to less than 30 years old.}
\label{fig:age_group_comparison}
\end{figure}

The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents. These items include views on affirmative action, women in the workforce, and politics, several civil liberties items, general views of whether people can be trusted, and views on whether doctors should let terminal patients die. These items tend to be in categories where a large proportion showed no evidence of active updating, which suggests an overall trend of these view being formed earlier in life (perhaps prior to inclusion in the GSS) and remaining relatively stable for the rest of life.

For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address pass discrimination, all evidence of active updating disappears for individuals over 30. This suggests that early adulthood is a time when these opinions are still malleable, but that over time they harden. For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.

Eight of these items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed less evidence of active updating. These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, views on whether aging parents should live with their children, and views on whether police should be allowed to use force against suspects who are verbally abusive toward them. Some of these items, such as views on whether aging parents should live with their children might be things young people are not forced to really consider until later in life and as a result do not form clear opinions until that exposure.

The remaining 89 items explored in this analysis (just under half of all items) show evidence for active updating but do not show consistent age heterogeneity. However, this does not mean they show strong evidence of an active updating model. We now turn to exploring the extent of evidence for active updating for items in the data set.

## Item heterogeneity

Which items show the strongest evidence for active updating? Figure \ref{fig:phi_relig} through \ref{fig:phi_racgen} plot $\phi$ estimates for all items included in the analysis, grouped by subject material. We constrain items that showed no evidence of active updating to $phi = .5$ in these plots. For ease of interpretation, we omit from these plots one item that showed a statistically significant negative $\phi$ estimate: whether respondents can imagine situations in which they would approve of a policeman striking an adult male citizen.

\begin{figure}[!ht]
\includegraphics{change_figures/relig_phi_estimates.pdf}
\centering
\caption{Phi estimates for items about religious activity and beliefs, social life, subjective SES, and suicide.}
\label{fig:phi_relig}
\end{figure}

\begin{figure}[!ht]
\includegraphics{change_figures/poli_phi_estimates.pdf}
\centering
\caption{Phi estimates for items about guns, law, crime and policing; politics and government; and public spending.}
\label{fig:phi_poli}
\end{figure}

\begin{figure}[!ht]
\includegraphics{change_figures/civlib_phi_estimates.pdf}
\centering
\caption{Phi estimates for items about civil liberties, confidence in leadership, health, morale, and social trust.}
\label{fig:phi_civlib}
\end{figure}

\begin{figure}[!ht]
\includegraphics{change_figures/racgen_phi_estimates.pdf}
\centering
\caption{Phi estimates for items about race, gender, sex, sexuality, and abortion.}
\label{fig:phi_racgen}
\end{figure}

The overwhelming takeaway from these figures is that even for those items where there is some evidence, the majority of change in these items is non-persisting, either because there is significant measurement error in individuals' responses or because individuals make non-persisting changes at higher rates than they make persisting change. The majority of items that show evidence of active updating have $\phi$ estimates less than .6, with only five items exceeding .7. Items that show the highest rate of active updating include items with clearly identified external referents that change (e.g., confidence in the executive branch of the federal government), items that were strongly influenced by the Great Recession of 2008 (a person's assessment of whether their financial situation has changed and ability to get a new job), and facts such as whether someone owns a gun, which are easily recalled and tend to have low levels of measurement error.

Because change can take on a variety of forms, and because the estimate of $\phi$ varies depending on the direction and relative variance of real change and measurement error, these $\phi$ estimates do not translate into a clearly understood quantity of change. For reference, however, in simulated models where persisting and non-persisting change are drawn from the same distribution with no measurement error, a $\phi$ of .6 equates to about twice as many people making non-persisting changes as persisting changes.

It's impossible to talk about all results in these tables, but we highlight some notable results and trends. Starting with Figure \ref{fig:phi_relig}, religious activity is the only category where all items show evidence of an active updating process, and these items appear to have stronger evidence than other kinds of questions, with $\phi > .6$ for three of the four items. Some items that we might assume would have high evidence of persistence do not. Whether a person had a religious turning point or was "born again" show no evidence of persistence, a surprising finding given that these changes should be non-reversible.

Almost all items dealing with politics, government, and public spending show evidence for an active updating model, with $\phi$ estimates close to .6. Within items dealing with politics, which political party an individual identifies with shows more evidence than other questions for active updating, while general political ideology and views on the government's role show less evidence.

Within the civil liberties category, persisting change does not appear to be concentrated among either a type of freedom (library, speech, college) or a subject of that freedom (atheist, militarist), as types and subjects see both persisting and non-persisting change. The notable exception here is items about freedom for homosexuals, all of which show evidence of persisting change. Similarly, all but one of the scales show evidence of persisting change, suggesting that removing measurement error increases our ability to detect the true form of change in the data. This lack of a clear pattern for most civil liberties questions raises questions about how and why people update their views on civil liberties.

Within the "confidence in leadership" category, low-visibility institutions with harder-to-identify leadership, such as the scientific community, education, organized labor, and medicine, tend to have no evidence for active updating, while high-visibility institutional leaders such as organized religion, the financial system, and the executive branch of the federal government show evidence of active updating.

As noted previously, gender, race, and sexuality topics in Figure \ref{fig:phi_racgen} tend to have a higher proportion of settled items compared to other categories. Even for those items that do show evidence of persistence, the $\phi$ estimates tend to be smaller (less than 0.6), with the exception of items such as support for same-sex marriage and support for abortion in the case of rape. Within items about sex and sexuality, all but two items on abortion appear to be settled.

As noted previously, low measurement error might be a reasonable assumption for some items. Previous studies using different approaches to measuring the reliability of survey reports suggest that some items captured in our study, such as whether a person owns a gun, are measured with a high degree of reliability (> .9) (Hout and Hastings). For other items, such as confidence in the leadership of major companies ($\phi = .58$), reliability might be as low as .5. There is very little correlation ($\rho = .165$) between our $\phi$ estimates and reliability estimates.

While we cannot control for measurement error in our analysis, we can take steps to mitigate its impact. Appendix A presents results comparing items with more than three scale points to coarsened versions of these question with either two or three scale points. If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree). If scales included a clear midpoint, we included that and coarsened responses on either side. For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median.

Generally speaking, most items with no evidence of active updating continue to have no evidence of active updating when coarsened. For items that showed evidence for active updating, the coarsened estimates tended to be very similar to the uncoarsened estimates, suggesting that persisting and non-persisting changes happen about as often between ends of these scales as they do within ends of these scales.

There are a couple notable departures from this general pattern. Several political questions -- general political views, views on whether the government should reduce inequality, and views on whether the government should help blacks -- show decreased evidence of persisting change when coarsened. This suggests that changes around the midpoint of the scale tend to be measurement error, a finding that's consistent with previous work suggesting that individuals without settled political views tend to choose points around the middle of the response scale (Converse).

## Discussion

\pagebreak

# Appendix A: Coarsened variable estimates

Phi estimates for coarsened items are presented in Figures \ref{fig:coarse_relig} through \ref{fig:coarse_racgen}. If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.

\begin{figure}[t]
\includegraphics{change_figures/coarse_relig.pdf}
\centering
\caption{Phi estimates for coarsened questions about religion, subjective socioeconomic status, and morale.}
\label{fig:coarse_relig}
\end{figure}

\begin{figure}[t]
\includegraphics{change_figures/coarse_poli.pdf}
\centering
\caption{Phi estimates for coarsened questions about politics, government, sex and sexuality.}
\label{fig:coarse_poli}
\end{figure}

\begin{figure}[t]
\includegraphics{change_figures/coarse_racgen.pdf}
\centering
\caption{Phi estimates for coarsened questions about race and gender.}
\label{fig:coarse_racgen}
\end{figure}

### 2019-12-06 old_drafts/manuscript7.Rmd
kevin's revisions

--- 2019-11-14 manuscript8.Rmd
@@
 bibliography: panel_change_bib.bib
-abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change.
-One approach theorizes that people actively update their beliefs in the face of new information.
-The other argues that, following early socialization experiences, individuals' dispositions are stable.
-We formalize these two models, elaborate empirical implications of each, and derive a simple model for comparing their prevalence using panel data.
-We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey.
-The pattern of results is more consistent with the settled dispositions model than the active updating model, and most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes in belief.
-When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes.
-We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.
+abstract: Models of population-wide cultural change tend to invoke one of two broad
+  models of individual change. One approach theorizes that people actively update
+  their beliefs in the face of new information. The other argues that, following early
+  socialization experiences, individuals' dispositions are stable. We formalize these
+  two models, elaborate empirical implications of each, and derive a simple model
+  for comparing their prevalence using panel data. We test this model on more than
+  184 attitude and behavior items from the 2006-14 rotating panels of the General
+  Social Survey. The pattern of results is more consistent with the settled dispositions
+  model than the active updating model, and most observed change in the GSS appears
+  to be short-term attitude change or measurement error rather than persisting changes
+  in belief. When persistent change occurs, it is somewhat more likely to occur in
+  younger people than older people and more common for public behaviors and beliefs about 
+  high-profile issues than private attitudes. We argue that there is a place
+  for both models in our theory of cultural evolution but that we need more evidence
+  on the circumstances under which each is more likely to apply.
 ---
@@
 library(ggrepel)
-library(here)
-'%!in%' <- function(x,y)!('%in%'(x,y))
-
 ```
@@
 
-Attempts to account for opinion change at the societal level have produced an array of often conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent, real attitudes.
-Within sociology, a number of models suggest that individuals lack the cognitive tools to maintain consistent beliefs on social and political issues.
-As a result, they construct responses in the interview setting, drawing on ideas they hear from opinion leaders and changing their attitudes when elite discourse changes [@zaller; @converse; @perrinmcfadden].
-Cohort replacement theories of attitude change assume that individuals hold opinions, that people are unwilling to alter them in the face of societal change, and that public opinion changes in response to generational turnover [@mannheim; @ryder].
-Another broad array of models argue that individuals and the public at large hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
-xi; @achen1992].
-More recently, sociologists have suggested that individuals attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].
-
-Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change that have their roots in theories of cultural sociology.
-The first of these individual-level models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation.
-This model is rooted in a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001).
+Sociological analyses of cultural change tend to focus on population-wide changes in attitudes, beliefs, or practices that, while rarely made explicit, invoke one of two broad models of individual cultural change.
+The first is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions.
+In this model, people continuously revise their views and practices based on the people, discourse, and events around them.
 The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life.
-This model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another.
-Separate from this distinction is a second dimension of attitude formation that reflects degree to which individuals can consistently report the same opinion over time.
-Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller].
-Attitudes can also be in a process of active updating but relatively easy to report, such as when individuals change their registered partisan affiliation.
-
-In this paper, we make these two dimensions of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for comparing the prevalence of active updating using panel data.
+In this model, adults' views and practices are relatively robust to changes in the social environment.
+
+Both accounts have affinities with traditions in sociological theory.
+The active updating model is most compatible with a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001).
+The settled dispositions model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
+
+In this paper, we make these two core models explicit, deduce some of their empirical implications, and derive a simple strategy for comparing their prevalence using panel data.
 We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
-By focusing on the pattern of attitude change at the individual level, we can elucidate the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change apply.
 Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
@@
 Our analysis yields several results.
-First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change.
+First, we find that the majority of what appears to be individual-level change in belief or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change.
 Simply put, there is little evidence that U.S.
 adults are changing their beliefs or practices in lasting ways over this period of study.
-Second, settled opinions vary in how consistently individuals report the same answer.
-Consistent with theories that argue that individuals lack clear opinions, some survey items appear elicit inconsistent responses and a few appear to elicit random responses.
-At the same time, individuals are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
-third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
-On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life.
-Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and individuals are more likely to report these attitudes consistently.
+Second, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
+That is, on several items at least, it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life.
+Third, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology).
 This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.
 
-These findings generally are more consistent with the settled dispositions model than with the active updating model and support theories claiming that cultural change tends to come through generational turnover rather than persuasion.
-However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change and the life course.
-The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions.
+These findings are generally more consistent with the settled dispositions model than with the active updating model.
+However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change.
 We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.
 
+In addition to their theoretical significance, our findings also have practical methodological implications.
+We argue that short-duration panel studies of adults will be of limited utility for studies of cultural change unless they have a clear focus on the sorts of things that are likely to change in persistent ways over the study period.
+Otherwise, researchers will waste resources measuring "change" that turns out to be noise.
+In general terms, then, we suggest that social scientists would do well to conduct more panel studies of children and adolescents while using cross-sectional designs to study adults.
+
 # Theoretical Background
-
-## Belief Formation in Cultural Sociology
 
@@
 Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state.
-This formal property, and this active updating model generally, is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings].
-Concerns about panel conditioning, in which a person's participation in a survey influences her subsequent answer choices, either by changing her non-survey behavior or by altering how they respond to the survey without changing behavior, also imply some sort of active updating on the part of the respondent [@warrenhalpernmanners].
+This formal property, and this active updating model generally, is often assumed to underlie data generation in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings].
 
 More informally, this AUM posits an agent that is updating her views in the face of social experience.
-There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
+There are formal Bayesian ways of modeling updating, but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
 Following Gross's [-@gross09, pp.
@@
 The models thus differ fundamentally on their emphasis on the character of personal change.
-In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time.
+In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline.
 In the SDM, changes tend to revert because each person returns to their existing baseline.
@@
 
-## Consistency in Response
-Neither model makes any assumptions about the magnitude of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
-This means that we must consider separately the overall amount of change in responses from wave to wave and the persistence of this change.
-Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development.
-However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make much greater inference about the applicability of different theories to these items.
-
 ## Combining the Models
 
-These two dimensions of attitude change give us a two-by-two matrix of patterns for attitude change.
-The first dimension reflects what we call active updating, or the idea that more recent responses are better predictors of current responses that older responses.
-In other words, active updating means that when changes happen, they are more likely to persist than revert.
-We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses.
-Individuals might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their older positions.
-The second dimension reflects what we call stability, or how consistent individuals are in reporting their beliefs once we account for the amount of active updating.
-
-With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them.
-Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix.
-We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes.
-Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.
-
-\begin{center}
-\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
-\hline
- & Low Active Updating & High Active Updating \\ 
-\hline
-High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\
-\hline
-Low Consistency &  \\ 
-\hline
-\end{tabular}
-\end{center}
-
-### Age and Attitude Formation
-
-In broad terms, cohort replacement theories of public opinion change suggest that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
-Implied in these models is the notion that individuals have a period of "cohortization" in which they form beliefs.
-
-To continue the metaphor above, cohort replacement theories imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition).
-The concept of a "cohort effect" is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman].
+Although our presentation above highlights differences between these models, they are not necessarily contradictory.
+There are at least two ways to understand how both models could apply without contradiction.
+
+### Age and Cohort Formation
+
+The easiest way to reconcile the AUM and SDM is to posit that they operate sequentially over a person's life.
+A range of sociological theories provide conflicting hypotheses and evidence for how susceptible individuals are to attitude and behavior change at different life stages.
+Most of these hypotheses suggest that early adulthood is a time of heightened susceptibility attitude change.
+Two hypotheses -- the "increasing persistence" hypothesis [@glenn; @inglehartbaker] and the "impressionable years" hypothesis [@visserkrosnick] -- suggest that susceptibility to attitude change is highest early in life and decreases (gradually or quickly, respectively) as individuals age.
+The "life stages" hypothesis [@visserkrosnick; @danigelis] posits that both early (<40) and late (60+) life are times of more attitude change but that mid-life is characterized by limited attitude change.
+
+To continue the metaphor above, one can imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition).
+This is, in fact, the logic that underlies the notion of a "cohort effect." The idea of a cohort effect is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman].
 Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo].
@@
 
-There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick].
-Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].
-
-Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender.
-Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort” change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts.
-Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].
-
-As Table ## suggests, items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults in the population, but high active updating for young people.
-Because these approaches generally assume that beliefs are prominent in the minds of respondents, the implication is that these beliefs are reported with high consistency in adulthood.
-At younger ages, however, we expect more volatility as attitude development occurs.
-We note, however, that there is some evidence that volatility of settled beliefs increases as individuals change as well [@krosnickalwin], but we do not expect this to be the general pattern.
-
-### Inconsistent Opinions
-
-Not all cultural change occurs through cohort replacement processes, and there is abundant evidence of large-scale public opinion shifts that occur in a span of time too short to be driven entirely by cohort replacement.
-This suggests that, at times, even middle-aged individuals update their opinions in persisting ways.
-
-In contrast to cohort replacement theories, a range of theories posit that, outside of a well informed elite, individuals lack consistent opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse].
-For example, when asked whether the government should spend more on health care, forming an opinion requires an individual to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result.
-Most individuals do not have this information readily stored in their minds.
-
-Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment.
-This explains why individuals are susceptible to question-order, priming, and recency effects [@zaller].
-However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues.
-Shifts in public opinion can also come about when elites shift their opinions on issues, providing new signals to the general public.
-
-In general, then, we expect low-profile and complex public policy issues to show a low rate of active updating and a low rate of consistency.
-By definition, most questions of public policy should fall into this group, since only a few issues can sustain a relatively high level of public attention.
-The small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these.
-However, we do spotlight one issue that saw significant change in elite opinion.
-In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of rights for homosexuals.
-Leaders of the Democratic Party including President Barack Obama came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
-For this issue, then, we might expect that public opinion shifts notably.
-This should manifest as a high degree of active updating for the whole population, not just young people.
-
-Outside of a lack of information on public policy issues, there are two other reasons why individuals might demonstrate low consistency and low active updating.
-First, individuals might be unwilling to report socially undesirable beliefs if they hold them, and as a result obfuscate their true position.
-Similarly, some issues might not require a lot of information to comprehend, but might be so far from the public consciousness as to be nonsensical to respondents.
-We expect both of these conditions to produce a similar pattern of low active updating and low consistency.
-
-### Updating Opinions
-
-The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, individuals update their opinions across the life course in response to gathering new information and interacting with changing social environments.
-
-Some of these approaches, common in political science, suggest that individuals consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman].
-These models do not tend to make distinctions about what kinds of attitudes shift over time and what do not.
-However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.
-
-Other theories suggest that only some beliefs are open to updating during adulthood.
-A number of theories of attitude development posit that individuals hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein].
-These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams.
-In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find little support for Lakoff's model and significant support for ideological identification as this "core" belief in the political space.
-If these  individuals have the ability to regularly assess new information against these core beliefs, they should also be able to recognize their position in a multiple-choice survey questionnaire and report them with limited error [@vaisey09].
-In other words, we expect that core beliefs such as ideological identification should have high consistency and low active updating, and conversely that items that show these traits to have the potential to be core beliefs.
-
-In a similar vein of research focused on political polarization, Baldassari and Gelman [@baldassarigelman] find little increase in the correlation between specific policy positions over time, suggesting limited active updating of these items.
-What they do find is an increased alignment between partisan identification and policy positions.
-They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
-442].
-This suggests that among items tapping issues of politics, partisan affiliation should demonstrate the highest level of active updating, and that it should be reported with a high degree of consistency.
-
-### Public vs. Private Culture
-
-A final dimension that likely influences the consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
-The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp.
-91].
-This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances, which might lead to a low degree of consistency and active updating.
-However, when public culture provides a strong "scaffold" for individuals through clear signals of how their identities and social locations should influence how they should respond to a survey question, it becomes easier for them to maintain consistency [@lizardostrand].
-
-We have touched on this idea already with our discussion of the idea that high-profile elite opinion change might drive opinion change in the broader public, or by noting that high-profile issues might be easier to report consistently because individuals receive signals that support that position.
-Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
-Any item where the burden of maintaining consistency is externalized and not in an individual's head might demonstrate both higher rates of active updating and greater consistency.
-
-### Toward Theoretical Synthesis
-
-The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating.
-As such, this disagreement forms the primary structure of our analysis.
-A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development.
-We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating, but we should still observe high consistency and low active updating.
-
-If what we find is a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
-Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
+This discussion of cohorts brings us close to the vast literature on age-period-cohort decomposition models, which is focused on statistical techniques to partition variance among sources of social change in repeated cross-sectional data [@yangland].
+Although this work is important, it is focused on modeling changes in population aggregates whereas we are concerned with modeling cultural change from the point of view of the same actor moving through time.
+
+The two are related and have implications for each other, of course.
+Again, the idea of a cohort requires that some people (i.e., younger people) are influenced by the environment to make persistent change in ways that others (i.e., older people) are not.
+This implies a transition, perhaps a gradual one, from active updating to a settled disposition.
+A further connection between the two is that the world implied by the SDM can only change in the aggregate due to cohort replacement.
+That is, if persisting change is extremely rare among adults, the only way for persisting change to occur in the population is for older cohorts to die and be replaced with younger cohorts with (perhaps) different baselines.
+For our purposes, the most important idea here is that "cohortization" implies that the AUM might describe the data-generating process better among younger respondents whereas the SDM might describe it better among older respondents.^[We do not have space for a full discussion here, but we regard these approaches as complementary.
+Repeated cross-section approaches must make strong assumptions about individual processes but are able to use many years of data.
+A panel approach is better able to capture micro-level processes but at the cost of using only a few years of data; this requires making strong assumptions about the generalizability of the process across historical time.]
+
+### Different Items, Different Processes
+
+Another way to reconcile the AUM and the SDM is to consider that they apply to different sorts of questions a researcher might ask a respondent, or that some attitudes and behaviors are stable while others are constantly updating.
+We can illustrate this idea with (non-cultural) examples we understand well and then consider how less obvious cultural processes might be similar or different.
+
+First, if we repeatedly ask a sample of women "how many children have you ever given birth to?", we know the data is generated by a process that looks like Equation (1).
+Leaving measurement error aside, the response is a simple combination of how many children the respondent had last time we asked ($y_{it-1}$) and whether or not she has had any since ($\nu_{it}$).
+
+Second, if we repeatedly ask a sample of adults "how many full siblings did you have at age 16?", we know the true answer cannot change and therefore Equation (2) is the best model.
+This is because $U_i$ refers to a fixed quantity and cannot change over time.
+Responses could still vary due to measurement error (e.g., misunderstanding the question) meaning that the variance of $\nu_{it}$ need not be 0.
+But the underlying reality is, by definition, a settled matter.
+
+Finally, there are responses that have no stable component but are entirely functions of local circumstances.
+If we repeatedly ask a sample of people in the same city "did it rain yesterday?", their responses would not be described better by Equation (1) or Equation (2).
+In this case, everything that matters is in the contemporary circumstances ($\nu_{it}$) and thus the terms that distinguish Equation (1) from Equation (2) ($y_{it-1}$ or $U_i$) are ignorable.
+These types of items are not important for adjudicating our research questions, but we still need to be aware of this possibility.
+
+What we want to determine, then, is how much a particular item resembles these different processes.
+Consider repeated requests for (dis)agreement with the prompt "a working mother can establish just as warm and secure a relationship with her children as a mother who does not work." Does the aggregate pattern of responses look more like an active updating process where a person might be changing her mind?
+Or does it look more like asking a person to report repeatedly on a settled matter?
+
+When it comes to beliefs and behaviors, which items will demonstrate persistence likely depends on a number of factors.
+Vaisey and Lizardo [-@vaiseylizardo] found evidence that beliefs about gender, race, and other "core" opinions are more driven by cohort effects and are therefore less likely to be actively updated by adults.
+Using repeated cross-sectional data, Danigelis, Hardy, and Cutler [-@danigelis] find that "intracohort" change (a good indicator that some cohort members are changing their minds) is larger for young individuals (18 to 39 years old) than for older individuals (over 60 years old) for attitudes about economic gender equality and structural reasons for racial inequality.
+For attitudes pertaining to civil liberties, political gender equality, and privacy, cohorts tended to show similar rates of intracohort change at both young and old ages during the time frame they studied.
+
+Previous studies have also found that political attitudes are more volatile during early and late adulthood but stabilize during middle age [@krosnickalwin].
+While volatility of responses increased as individuals aged, this is mostly due to increases in measurement error, rather than persisting change.
+This generally suggests that political views will show decreasing evidence of persistent change as age increases.
+
+"Public culture," such as participating in social activities, may show similar rates of persistence over time because social mechanisms (like joining a group) can help maintain changes in ways that intrapsychic forces cannot.
+Finally, items that capture issues that are highly salient in the media during the study period may be more likely to show active updating because of heightened exposure.
+Sears and Valentino [-@searsvalentino] find that during political campaigns adolescents tended to adopt new beliefs about issues that were salient during the campaign but not those issues peripheral to it.
+Given these different possibilities, we do not begin this study with clear predictions on all issues, but general expectations that items pertaining to private cultural beliefs will show less persistent change and that items related to public cultural expressions will show more persistent change.
+Because of the large, high-profile discussion on gay marriage and gay rights during this period, however, we do expect to see active updating on related items.
+
+## Toward Theoretical Synthesis
+
+The active updating model and the settled dispositions model are different and they do have affinities with different theoretical traditions in sociology.
+Indeed, the difference in emphasis between environmental cues, creative meaning-making, and public discourse on the one hand and settled dispositions, "habitus," and cohort replacement on the other has been a basic theoretical tension in cultural sociology for more than a decade [@vaisey09; @lizardo17; @vaiseylizardo].
+
+One model may, in fact, be a better starting place than the other.
+Previous work using cross-sectional data suggests that settled dispositions are hugely important [@vaiseylizardo].
+Despite these findings, the evidence is still mixed [@danigelis].
+Therefore our goal here is neither to declare victory for one of the theoretical perspective nor to simply say that both "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types processes are at work.
 We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
@@
 
-Third, _are there systematic differences in item content between questions that fall in different parts of our matrix?_ The preceding sections have made some specific predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs.
-As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.
+Third, _are there systematic differences in item content between questions that are better described by each model?_ Items differ from each other in many ways and it would be challenging to come up with clear predictions for all possible types of items.
+In general, however, we expect (following Vaisey and Lizardo) that core beliefs about gender, family, race, and trust should be more consistent with settled dispositions because their formation should be already be complete for adults.
+Items that require public commitments such as changing political parties, socializing, or religious attendance and affiliation should be open to persistent change because there are social mechanisms (not just cognitive ones) to _maintain_ the change.
 
@@
 We discuss item selection below.
-While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.
 
@@
 
-Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time.
-We first consider measuring the amount of active updating.
-Our two different models make different predictions in the three-wave panel context.
-The AUM makes the following prediction for wave 3:
+Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:
 
@@
 That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response.
-If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.
+If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses willl provide no additional predictive power.
 
@@
 
-Rather than estimate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
+Rather than estimate separate $\beta$ parameters for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
 If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B).
 If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.
-
-At the same time, our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating in responses.
-We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
-If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the population average, and $\beta = 0$.
-If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
 
@@
 The second challenge is measurement error, which is a form non-persisting change.
-In the attitude change literature, measurement error means different things to different authors.
-For some, it represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
-For other theories, measurement error reflects individuals' inability to accurately report their response.
-It is also possible that measurement error simply reflects errors of selection and interpretation (misunderstanding the question or incorrect coding).
-
-Because measurement error "looks the same" as non-persisting real change (and because the latter is sometimes interpreted as the former), estimates of $\phi$ will be biased toward .5.
+Because measurement error "looks the same" as non-persisting real change, estimates of $\phi$ will be biased toward .5.
 There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
@@
 The first is change in the overall population.
-Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates.
+Since our model includes an intercept, changes that shift all responses toward one end of the scale are aborbed into that term and not accounted for in our $\phi$ estimates.
 The second is change in the variance of responses.
@@
 First, we evaluate the overall evidence in support of the active updating model.
-To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
-We calculate the posterior probability that the model with the free parameter fits the data better.
-If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.
+To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
+We calculate the posterior probability that the model with the free parameter fits better.
+If the model with the constraint is preferred, then both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.
 
@@
 
-Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for AUM across variables as suggested by existing theories.
+Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for AUM across variables.
 Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be largely inductive.
@@
 ```{r}
-load(here("data", "results_df.Rdata"))
-load(here("data", "label_vars.Rdata"))
-load(here("data", "aum_vars.Rdata"))
-
-```
-
-Our model generates two parameter estimates of interest for each GSS item in our analysis: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
-High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
-$\phi$ is our measure of active updating and captures the relative proportion of wave 3 explained by wave 2.
+load("~/panel_change/data/results_df.Rdata")
+load("~/panel_change/data/label_vars.Rdata")
+load("~/panel_change/data/aum_vars.Rdata")
+```
+
+Our model generates two parameter estimates of interest for each GSS item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2.
 If responses are generated through a settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
@@
 This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.
-
-To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates.
-Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2, and only 147 (57 percent) of those individuals maintained the same response at wave 3.
-Under the settled disposition model, we would expect that about 50 percent or 129 individuals would maintain the same response, so we really only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting attitude change.
-In other words, even for the items that do show strong evidence of active updating, the overall rate of attitude change in the population is small.
-
-The right side of Figure \@ref(fig:phihist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
-That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they will tend to revert to their previously held position.
-This group includes many items addressing abortion, civil liberties, confidence in institutions, and views on race and gender.
-
-We will discuss in more detail below how different items perform.
+The right side of Figure \@ref(fig:phi_hist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
+That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they will tend to revert to their previously held position over time.
+
+Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items colored by whether they prefer the constraint or not.
+Although we discuss the pattern of items below, we label a few items that stand out.
+Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power.
+
+```{r phibetascatter, message=FALSE, fig.width=6.8, fig.height = 6.0, fig.cap="Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint."}
+res.df %>%
+  filter(grepl("_c", var)==FALSE) %>%
+  mutate(aum = ifelse(var %in% aum_vars, "\nPersistent\nChange\n", "\nNo Persistent\nChange\n")) %>%
+  ggplot(aes(x = p, y = b1)) + 
+  geom_point(aes(fill = aum, shape = aum)) + 
+  scale_shape_manual(values = c(21, 22)) + 
+  scale_fill_manual(values = c("gray", "black")) + 
+  geom_text_repel(data = label.vars, aes(label = var.name), size = 3) + 
+  theme_minimal() +
+  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
+  theme(axis.title=element_text(size=9))
+```
+
+\begin{figure}[t]
+\includegraphics{figures/phibeta_scatter.pdf}
+\centering
+\caption{Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint.}
+\label{fig:phibeta_scatter}
+\end{figure}
+
+Items showing evidence for active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
+A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
+
+We will discuss how different items perform below.
 To answer our first question, however, we need only focus on the overall distribution.
@@
 
-Of the 111 items that showed evidence for active updating in the last section, 23 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
-Figure \@ref(fig:agegroupcomparison) plots the estimates of these 23 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
+Of the 111 items that showed evidence for active updating in the last section, 23 showed differential effects of age for more than 50 percent of cutoff ages we tested.
+To get a sense of the magnitude of difference between older and younger individuals, Figure \@ref(fig:agegroupcomparison) plots the estimates of these 23 items for individuals above 30 and individuals equal to or below 30.
 
@@
 
-load(here("data", "age_results.Rdata"))
-load(here("data", "good_vars.Rdata"))
+load("~/panel_change/data/age_results.Rdata")
+load("~/panel_change/data/good_vars.Rdata")
 
@@
 The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents.
-These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die.
+These items include views on affirmative action, women in the workforce, and politics, several civil liberties items, general views of whether people can be trusted, and views on whether doctors should let terminal patients die.
 These items tend to be in categories where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable for the rest of life.
@@
 These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, views on whether aging parents should live with their children, and views on whether police should be allowed to use force against suspects who are verbally abusive toward them.
-Some of these items, such as views on whether aging parents should live with their children, might be things young people are not forced to consider until later in life and as a result do not form clear opinions until that exposure.
+Some of these items, such as views on whether aging parents should live with their children might be things young people are not forced to really consider until later in life and as a result do not form clear opinions until that exposure.
 This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is relatively unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previous theorizing has accounted for.
@@
 
-While there is a high proportion (40 percent) of items that show no evidence of active updating, and those that do show evidence show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
-Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
-
-Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items colored by whether they preferred the $\phi = .5$ constraint or not.
-Although we discuss the pattern of items below, we label a few items that stand out.
-Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.
-
-```{r phibetascatter, message=FALSE, fig.width=6.8, fig.height = 6.0, fig.cap="Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint."}
-res.df %>%
-  filter(grepl("_c", var)==FALSE) %>%
-  mutate(aum = ifelse(var %in% aum_vars, "\nPersistent\nChange\n", "\nNo Persistent\nChange\n")) %>%
-  ggplot(aes(x = p, y = b1)) + 
-  geom_point(aes(fill = aum, shape = aum)) + 
-  scale_shape_manual(values = c(21, 22)) + 
-  scale_fill_manual(values = c("gray", "black")) + 
-  geom_text_repel(data = label.vars, aes(label = var.name), size = 3) + 
-  theme_minimal() +
-  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
-  theme(axis.title=element_text(size=9))
-```
-
-\begin{figure}[t]
-\includegraphics{figures/phibeta_scatter.pdf}
-\centering
-\caption{Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint.}
-\label{fig:phibeta_scatter}
-\end{figure}
-
-Items showing evidence for active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
-A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
-
 Which items show the strongest evidence for active updating?
 There is no way we can possibly discuss all 184 items in detail without the discussion becoming tedious.
-Although we include $\phi$ estimates for all items in Appendix B, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
+Although we include all individual results in Appendix B, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
 We constrain items that showed no evidence of active updating to $\phi = .5$.
@@
 ```{r summary, message=FALSE, fig.width=6.8, fig.height=9.0, fig.cap="Summary of phi estimates for all items, by topical group."}
-load(here("data", "summary_plot_data.Rdata"))
-
+load("~/panel_change/data/summary_plot_data.Rdata")
 plot.data %>%
@@
 Consistent with the findings and theory of Vaisey and Lizardo [-@vaiseylizardo], we see that over half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
-These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
 Views on these "core" issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood (and thus becomes eligible to participate in the GSS).
@@
 
-Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them, tapping the idea that public culture is more persistent than personal culture.
-If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
-This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, as well as how often individuals engage in "religious activities", which are more nebulous questions that display less active updating.
-Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
-Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
-
-Other high-$\phi$ items have a _changing referent_.
+Some of the high-$\phi$ items have a _changing referent_.
 That is, although the item wording is the same, the object to which the question refers may change between survey waves.
@@
 
+Some of the high-$\phi$ items rely on _external mechanisms_ that likely help maintain them.
+If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
+Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
+Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
+
 Perhaps the most striking pattern in our findings is about questions about _gay rights_.
@@
 
+The categories that are consistently higher than .5 are public spending and politics and government, areas where political parties make efforts to keep issues in the public eye and encourage people to update their views.
+Even so, most of these values are still low in absolute terms.
+
 Finally, questions about religious activity, identity and (to a lesser extent) religious beliefs generally show some evidence for active updating.
@@
 
-### Political Beliefs
-
-Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, we examine these beliefs in greater detail than other items.
-Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for most of the questions in our data set tapping questions of political identity, the role of government, and specific policies, broken into four general areas.
-We break these items up into categories and remove a few items solely for ease of viewing.
-
-```{r polvariables, message=FALSE, fig.width=7, fig.height=8, fig.cap="Phi and beta estimates for 43 political attitudes and identities."}
-left_join(res.df, good_vars) %>%
-  filter(small.cat %in% c("Politics & government", "Public spending",
-                          "Sex, sexuality & abortion") |
-           var %in% c("cappun", "divlaw", "gunlaw", "grass", "courts",
-                      "letin1a")) %>%
-  mutate(small.cat = ifelse(small.cat %in% c("Race & immigration", 
-                                             "Guns, laws, crime & police")
-                            | var %in% c("prayer", "uswary"), 
-         "Misc. policy" , small.cat)) %>%
-  filter(var %!in% c("abscale6", "absingle", "teensex", "xmarsex",
-                     "homosex", "premarsx")) %>%
-  mutate(var.name = gsub("Government", "Gov't", var.name),
-         var.name = gsub("Abortion", "Abort", var.name),
-         small.cat = ifelse(small.cat == "Sex, sexuality & abortion",
-                            "Abortion & sex", small.cat)) %>%
-  mutate(small.cat = factor(small.cat, c("Politics & government", "Public spending",
-                                        "Abortion & sex", "Misc. policy"))) %>%
-  ggplot(aes(x = p, y = b1)) + 
-  geom_vline(xintercept = .5, linetype = 2, color = "gray") + 
-  geom_point(aes(fill = small.cat), shape = 21) + 
-  geom_text_repel(aes(label = var.name), size = 3) + 
-  theme_bw() + 
-  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
-  theme(axis.title=element_text(size=9)) +
-  facet_wrap(~small.cat) + 
-  theme(legend.position = "none")
-
-```
-
-There are several notable features of the figure.
-First, political party affiliation is a clear outlier in the plot, with much greater consistency and much greater active updating than other items.
-In contrast, questions asking about specific public spending priorities tend to have very low consistency, at least compared to other political items.
-Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government –  display higher consistency than specific policy questions but weak evidence of active updating.
-It is important to note that for political views, however, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
-
-While individuals tend to be inconsistent in their responses to spending priorities, views on some other specific policy questions with both high and low salience are reported quite consistently.
-Views on abortion, especially when considered as a scale, are quite consistently reported.
-As with all other scales, aggregating the composite items increases consistency but does not affect the amount of active updating.
-As discussed previously, views on gay marriage display high levels of active updating.
-In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating.
-This is notable, as public opinion and policies regarding both have shifted considerably in recent years, with many people assuming change in both is being driven by the same underlying process.
-Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but change in views on marijuana use have been driven primarily by cohort replacement.
+These patterns should be taken with a grain of salt since we began the study with only general notions of which type of items should show evidence for active updating.
+A story can always be told about any set of findings.
+We are encouraged, however, with the degree of fit between the findings here and our general expectations.
 
@@
 
-Third, _are there systematic differences in item content between questions that are better described by each model?_ With 184 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
-We find patterns of responses that provide a range of support for diverse theories of attitude formation.
-At the same time, our results are consistent with some general expectations about patterns in updating.
+Third, _are there systematic differences in item content between questions that are better described by each model?_ Although we began with only very general expectations about patterns in updating, our results were consistent with those expectations.
 Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
 Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18.
-The overall pattern of results, while diverse and at times hard to reconcile, does shed light on theoretical debates, to which we now turn.
-
-## Implications for Cultural Sociology
 
@@
 
-In the domain of cultural theory, our findings support the view that most cultural change happens slowly, through the mechanism of cohort succession.
-Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled during adulthood.
-The settled nature of these beliefs is generally coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report these beliefs wave after wave.
-In contrast, a range of views about race were so inconsistent that it would be difficult to call them either settled or updating.
-Even for those items that did display strong evidence for active updating (e.g.
-"whites rich or poor" and "whites work hard") were still extremely hard to predict wave after wave.
-
-In contrast to other beliefs, however, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging some people to update their views who might not otherwise do so.
-By definition, salience is a limited resource, meaning only few beliefs and behaviors could change at this rate at a time.
-The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
+In the domain of theory, our findings support the view that most cultural change happens slowly, through the mechanism of cohort succession.
+The pattern of findings for gay rights show, however, that a high degree of public salience and social movements can accelerate change by encouraging some people to update their views who might not otherwise do so.
+But the baseline process appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
 This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.
 
-While the dominant pattern is stability during adulthood, the results suggest greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories.
+The results also suggest much greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories.
 The most prominent and well supported theories of attitude change suggest a peak of susceptibility to attitude change early in adulthood and either a rapid or gradual decline in attitude change with age [@alwinkrosnick; @visserkrosnick].
@@
 
-## Implications for Political Sociology
-
-We said at the outset that these two models of individual culture underlie significant debates about public opinion formation, the formation of individual political beliefs over the life course, and the ability of individuals to hold and express clear opinions over time.
-What do our findings suggest for these debates?
-
-A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
-These respondents might express different positions from wave-to-wave, but in guessing what an individual will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
-While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
-With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still quite limited.
-Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.
-
-These results are consistent with theories that posit that, at least in the current era, individuals organize their beliefs around ideological identity [@boutylinevaisey; @converse], rather than some other organizing heuristic such as partisan identification, moral views, religion, or particular policy positions.
-It is also consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
-Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time.
-This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.
-
-The fact that there is not significant active updating for most political opinions should not be taken to imply that individuals are consistent in their positions.
-Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions.
-This is clearest when it comes to questions that deal with spending priorities.
-Prior responses to questions about government spending on foreign aid, science, drugs, child care, crime prevention, mass transit, national parks, and the problems of cities are bad predictors of what individuals will say later on.
-
-That being said, individuals are not wholly without consistent opinions.
-On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next.
-While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.
-
-The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents hold a general political identity and a few clear views on issues like abortion, generally attempt to align their identification to their views, and respond to elite opinion change when it provides clear signals.
-It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates.
-Individuals tend to have more-or-less settled views, especially on core questions such as ideology, and report (with a high degree of inconsistency) more specific policy positions based on these.
-
-## Methodological Implications
-
-Methodologically, our results highlight the challenges of evaluating population-wide attitude change using short-term panel studies.
+Methodologically, our results highlight the challanges of evaluating population-wide attitude change using short-term panel studies.
 The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change.
@@
 
-Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of younger individuals to understand how these attitudes are formed.
+Because many attitudes, including views on abortion, race, gender roles, social trust, and insitutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of younger individuals to understand how these attitudes are formed.
 Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful.
-In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
-Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike as as particularly useless.
-Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
+In a similar vein, it does not seem wortwhile to ask certain GSS questions repeatedly, and those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
 Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production" that give rise to persistent beliefs in adulthood.
@@
 ```{r loadfiguredata}
-load(here("data", "figure_data.Rdata"))
+load( "~/panel_change/data/figure_data.Rdata")
 ```
@@
 ```{r loadcoarse}
-load(here("data", "coarse_vars.Rdata"))
-
+load("~/panel_change/data/coarse_vars.Rdata")
 ```

### 2019-12-06 old_drafts/manuscript8.Rmd
kevin's revisions

--- 2019-11-15 manuscript8.Rmd
@@
 This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???).
+Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another.
 Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time.
@@
 With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them.
-Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix.
+Table 1 sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix.
 We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes.
@@
 
-\begin{center}
-\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
-\hline
- & Low Active Updating & High Active Updating \\ 
-\hline
-High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\
-\hline
-Low Consistency &  \\ 
-\hline
-\end{tabular}
-\end{center}
+![Table 1: Predicted change patterns for theoretical models](figures/theory_table.png)
 
@@
 
-In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
-Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.
+In broad terms, cohort replacement theories of public opinion change posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder; @vaiseylizardo].
+Implied in these models is the notion that people have a period of "cohortization" in which they form these initial beliefs.
 
@@
 There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick].
-Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].
+Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older people might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].
 
@@
 Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts.
-Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].
+Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than people changing their beliefs throughout their lives [@voaschaves].
 
@@
 
+### Panel Conditioning Bias
+
+Another possibility is that it is not everyday experience that drives changes and stability in beliefs, rather that participation in the survey itself produces change (or stability).
+This is commonly referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
+While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
+One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystalize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to "game" the survey to get through it faster.
+If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as individuals change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.
+
+We do not view this kind of panel conditioning as a problem for our theoretical models.
+If people change their attitudes or behavior or how they respond to survey questions as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
+This might lead to a higher estimate of active updating than we might observe in the population, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update their beliefs over time.
+
+A second form of panel conditioning bias would posit that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time.
+In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change.
+This would be problematic for our study, as it would under-estimate the amount of real change in the population.
+However, if this is the case, we should observe no active updating in responses and high levels of consistency in response, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.
+
 ### Public vs. Private Culture
@@
 
-That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response.
+That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0).
 If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.
@@
 If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
+
+### Comparison to Other Approaches
+
+Our model is not the first approach to measuring stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
+For example, Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS survey responses over time.
+This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
+These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier.
+While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount of "structural" or "non-structural" change is consistent across waves.
+This approach combines persisting and non-persisting change into two similar but distinct kinds of change "structural" and "non-structural," both of which can be persisting and non-persisting.
+This distinction, while important for some theoretical questions, is not our focus.
+
+A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief.
+This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
+Since we at times invoke both these models, we include a handful of composite scales (discussed below) of related items.
+If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in persistence.
+
+Finally, a nubmer of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
+These tools are well designed to address the questions they set out to.
+However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
 
@@
 
+We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation.
+One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.
+
+We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
+We see too much inconsistency in responses to most questions for models of commitment bias to be applicable.
+Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
+499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p.
+500], or questions that “induce respondents to provide socially nonnormative or stigmatized responses” [@warrenhalpernmanners: p.
+501], tend to show low active updating and low consistency.
+
+We cannot and would not want to rule out the possibility that panel conditioning is taking place in the GSS, and we believe it would be worthwhile to explore these same GSS panels for evidence of panel conditioning.
+However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
+
 Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.

### 2019-12-06 manuscript9.Rmd
minor edits

@@
     fig_caption: yes
-    reference_docx: word-styles-reference-01.docx
+    reference_docx: word-styles-reference-02.docx
   pdf_document:
@@
 As such, we designate a second term, _consistency_, to reflect this second dimension of change.
+_Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.
 
@@
 442].
-This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate a highest level.
-
-These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate a cascade of changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs.
+This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate the highest level.
+
+These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs.
 In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating.
@@
 This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these.
-However, we do spotlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population.
+However, we do highlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population.
 In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights.
 Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
-For this issue, we predict active updating throughout the population, not just young people.
+For this issue, we expect to find evidence of active updating throughout the population, not just among young people.
 
@@
 
-First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
+First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
 This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.
@@
 We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
-If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the population average, and $\beta = 0$.
+If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$.
 If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
@@
 Finally, a number of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
-These tools are well designed to address the questions they set out to.
+These tools are well designed to address the questions they set out to answer.
 However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
@@
 Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3.
-Therefore, we only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting change.
+Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change.
 In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.
 
-The right side of Figure \@ref(fig:phihist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
+The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
 That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
@@
 
-### Political Beliefs
+## Political Beliefs
 
@@
 
-First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
+First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
 Around 40 percent of all items show no evidence for updating.
-For the items that do show evidence of active updating, the overall rate of change in the population is low.
+For the items that do show evidence of active updating, the overall rate of change in the population is likely low.
 Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were.
@@
 
-The overall pattern of results, while diverse and at times hard to reconcile, does shed light on theoretical debates, to which we now turn.
+The overall pattern of results, while diverse and at times hard to reconcile, does shed light on several theoretical debates. We now consider these implications.
 
@@
 What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area?
-
 In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly, through the mechanism of cohort succession.
@@
 For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
-Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to factors that encourage openness to change at different ages.
+Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages.
 Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes.
-Doing so will expand our understanding of the social and psychological factors that that give rise to stable or variable attitudes [@howekrosnick].
-Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is a strong start [@eatonetal; @vissermirabile].
+Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick].
+Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].
 
@@
 We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
-We see too much inconsistency in responses to most questions for models of commitment bias to be applicable.
+We see too much inconsistency in responses to most questions for models of commitment bias to be a major explanatory factor.
 Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
@@
 
-We cannot and would not want to rule out the possibility that panel conditioning is taking place in the GSS, and we believe it would be worthwhile to explore these same GSS panels for evidence of panel conditioning.
+We cannot (and would not want to) rule out the possibility that panel conditioning is taking place in the GSS.
+We believe it is worthwhile to explore these same GSS panels for evidence of panel conditioning.
 However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.

### 2019-12-09 manuscript9.Rmd

@@
 abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change.
-One approach theorizes that people actively update their beliefs in the face of new information.
+One approach theorizes that people actively update their beliefs and behaviors in the face of new information.
 The other argues that, following early socialization experiences, dispositions are stable.
@@
 This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-
-In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data.
+These two models represent different approaches to understanding how people come to hold different forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo2016].
+
+In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data.
 In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors.
 We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
-By classifying the pattern of change at the individual level, we can clarify when different accounts of aggregate change are more likely to apply.
+By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply.
 Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
@@
 This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].
+In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.
 
@@
 
-The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007].
+The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise].
 The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
@@
 Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
-While Warren and Halpern-Manners [@-warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
+While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
 One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
@@
 The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
-This pattern is strongly consistent with Zaller's [@-zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
+This pattern is strongly consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
 
@@
 For the items that do show evidence of active updating, the overall rate of change in the population is likely low.
-Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were.
+Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to a stable baseline.
 What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
@@
 
+The overall pattern of results suggests that one reason attitudes are largely stable is because most issues simply do not reach the kind of salience necessary to shift people's opinions.
 In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
@@
 For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
+This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level.
 Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages.
@@
 
-Our work cannot speak to the strength of ideological commitments or whether individuals have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions.
+The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on most specific policy questions.
 The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
@@
 On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next.
-While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.
+While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [-@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.
 

### 2020-02-10 manuscript9.Rmd
conditional accept revisions

@@
 We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data.
-We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey.
+We test this model on 183 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey.
 Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model.
@@
 
-Attempts to account for opinion changes in society have produced conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes.
+Attempts to account for opinion changes in society have produced conflicting theories about the process of opinion formation and the ability of individuals to maintain consistent attitudes.
 Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues.
 As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden].
-By contrast, cohort replacement theories posit that people do hold opinions, are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder].
+By contrast, cohort replacement theories posit that people do hold opinions and are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder].
 Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
@@
 This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-These two models represent different approaches to understanding how people come to hold different forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo2016].
+These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].
 
@@
 In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors.
-We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS).
+We then apply this method to 183 items from the 2006-2014 General Social Surveys (GSS).
 By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply.
@@
 Simply put, there is little evidence that large numbers of U.S.
-adults are changing their beliefs or practices in lasting ways over this period of study.
+adults changed their beliefs or practices in lasting ways over this period of study.
 Second, settled opinions vary in how consistently individuals report the same answer.
@@
 On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life.
-Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently.
+Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to persist than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes and behaviors consistently.
 This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.
@@
 However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course.
-The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to specific policy positions.
+The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to their general political beliefs.
 We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
@@
 To make things as explicit as possible, we can write the data-generating process formally.
-Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all.
+Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions.
 We will explain this in greater detail below.
@@
 
-```{r dag, message=FALSE, fig.width=6.8, fig.cap="Different Models of Response Over Time"}
+```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
 library(ggdag)
@@
 
-More informally, this AUM posits an agent that is updating her views in the face of social experience.
-There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
+More informally, this AUM posits a person who is updating her views in the face of social experience.
+There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
 Following Gross's [-@gross09, pp.
@@
 
-The concept of active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
+Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
 When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model.
@@
 
-A number of theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the population is becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
+Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
 In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.
@@
 The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
-These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965).
-Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.
+These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder].
+Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.
 
@@
 The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have (Zaller 1992; Converse 1964), and suggest that peoples’ responses at any particular time are a random deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
-As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if these wave-to-wave differences are quite large.
+These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse], and suggest that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
+As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.
 
@@
 That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time.
-The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that these shifts will be erased over time as people return to their baselines.
+The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.
 
@@
 
-The models thus differ fundamentally on their emphasis on the character of personal change.
+The two general models thus differ fundamentally on their emphasis on the character of personal change.
 In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time.
-In the SDM, changes tend to revert because each person returns to their existing baseline.
+In the SDM, changes tend to revert because each person returns to their baseline.
 We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.
-We consider the implications of these predictions below.
 
@@
 This disagreement thus forms the primary structure of our analysis.
-However, it is possible that some items display active updating while others display settled dispositions.
+However, it is possible that some items display active updating while others do not.
 There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.
@@
 _Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman].
-To put it differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
+To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
 
@@
 
-_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein].
-Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe].
-In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint.
-Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time.
+_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein].
+This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe].
+In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space.
+Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time.
 They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
 442].
-This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate the highest level.
-
-These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs.
-In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating.
-Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions.
-Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.
+
+The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition.
+Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency.
+On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman].
+In other words, liberal Republicans should become liberal Democrats, not conservative Republicans.
+This means that partisan identification should display high levels of updating.
+Again, this does not mean that many people change, only that changes tend to persist.
+
+Alongside ideological identificaton, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman].
+For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy.
+Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently.
+There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.
+
+Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs.
+However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse].
+However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.
 
@@
 Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
-The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp.
+The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p.
 91].
@@
 Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse].
-This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these.
-However, we do highlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population.
-In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights.
+This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time.
+However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population.
+In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights.
 Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
@@
 
-Items that tap some public dimension of behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
+Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
 Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
@@
 
-First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
+First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
 This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.
 
-Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ As we just mentioned, in a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18).
+Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18).
 However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.
@@
 
-To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model.
+To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model.
 This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012).
@@
 
-That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0).
+That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2.
 If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.
@@
 
-Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction like this:
+Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:
 
@@
 If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model.
-However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, we begin to observe evidence that at least some individuals in the population engage in active updating.
+However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating.
 We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:
@@
 
-Rather than estimate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest for each item in our analysis: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
+Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
 If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B).
@@
 
-At the same time, our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating in responses.
+Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating.
 We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
@@
 Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
-Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS survey responses over time.
+Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS responses.
 This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
@@
 While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves.
-This approach also combines persisting and non-persisting change into two similar but distinct kinds of change "structural" and "non-structural," both of which can be persisting and non-persisting.
+This approach also combines persisting and non-persisting change into two similar but distinct kinds of change: "structural" and "non-structural," both of which can be persisting and non-persisting.
 This distinction, while important for some theoretical questions, is not our focus.
@@
 This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
-Since we at times invoke both these models, we include a handful of composite scales (discussed below) of related items.
-If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in persistence.
+Since we at times invoke both these models, we include in our analysis several composite scales of related items.
+If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in their levels of active updating.
 
@@
 One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
-If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.
+If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 would become a better predictor of wave 3.
 
@@
 If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
-This might lead to a higher estimate of active updating than we might observe in the population, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update their beliefs over time.
+This might lead to a higher estimate of active updating than we might observe in a population that did not take the survey, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update beliefs over time.
 
@@
 
-Finally, a number of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
+Finally, a number of approaches exist for evaluating theoretical process of belief formation and change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
 These tools are well designed to address the questions they set out to answer.
@@
 
-Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach.
-These challenges put some limitations on the conclusions we can reach.
-
-The first is that our model is designed to allocate variance explained to each of the prior predictors rather than to assign probabilities to each data-generating process.
+Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach and therefore limit the conclusions we can reach.
+The first is that our model is designed to allocate variance explained to each of the prior waves rather than to assign probabilities to each data-generating process.
 Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.
@@
 The second is measurement error, which is a form of non-persisting change.
-In the attitude change literature, measurement error means different things to different authors.
-For some, it represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
+For some researchers, measurement error represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
 For other theories, measurement error reflects individuals' inability to accurately report their response.
@@
 
-Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), estimates of $\phi$ will be biased toward .5, since wave 1 would be informative for all individuals who do change.
+Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), $\phi$ estimates will be biased toward .5, since responses with error would be random departures from baseline.
 There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
-On the other hand, previous studies of measurement reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
+On the other hand, previous studies of reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
 Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
@@
 
-Finally, with these findings in hand, we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
-Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be largely inductive.
+Finally we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
+Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be inductive.
 
@@
 
-We want to test our model on as broad a range of items tapping attitudes, beliefs, and social behaviors as possible.
-We sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
+To test our model on as broad a range of items as possible, we sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
 We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
-We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on their subject material.
-Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about public spending, and questions about social life.
+We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on subject material.
+Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about government spending, and questions about social life.
 
 We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust.
-This includes Rossi's six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
+This includes a six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
 We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are.
@@
 We combine four questions about support for suicide under different circumstances.
-We also created a scale of support for police use of violence against criminal suspects by averaging five binary questions about the conditions under which individuals support police use of violence.
-
-In total, we test the model on 184 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves.
+We also created a scale of support for police use of violence against criminal suspects by combining five binary questions about the conditions under which individuals support police use of violence.
+
+In total, we test the model on 183 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves.
 Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.
@@
 
-```{r}
+```{r, message = FALSE, warn = FALSE}
 load(here("data", "results_df.Rdata"))
-load(here("data", "label_vars.Rdata"))
-load(here("data", "aum_vars.Rdata"))
+load(here("data", "attitude_vars.Rdata"))
+
+label_vars <- res.df %>%
+  filter(var %in% c("intlblks", "confed", "selfhunt", "socbar")) %>%
+  left_join(attitude_vars)
+
+#create a list of variables that display updating
+aum_vars <- res.df %>% filter(aum_pref == 1)
+aum_vars <- aum_vars$var
 
@@
 High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
-$\phi$ is our measure of active updating and captures the relative proportion of wave 3 variance predicted by wave 2.
-If responses are generated through a settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
+Our measure of active updating, $\phi$, captures the relative proportion of wave 3 variance predicted by wave 2.
+If responses are generated through a true settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
 As the evidence of active updating increases, $\phi$ will increase toward 1.
@@
 
-To evaluate the evidence in favor of the active updating model, we compare for all 184 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
+To evaluate the evidence in favor of the active updating model, we compare for all 183 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
 If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.
 
-Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 184 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.
+Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 183 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.
 
@@
   gather(key = "key", value = "value", -var) %>%
-  mutate(key = recode(key, "p"="Phi estimates", "prob_aum"="Probability of some active updating")) %>%
+  mutate(key = recode(key, "p"="Phi estimates (Updating)", "prob_aum"="Probability of some active updating")) %>%
   ggplot(aes(x = value, y=..count../sum(..count..))) +
-  geom_histogram(fill = "gray", color = "black", bins = 10) + 
+  geom_histogram(fill = "gray", color = "black", bins = 15) + 
   geom_vline(xintercept = .5, color = "black", linetype = 2) + 
@@
   facet_wrap(~key, scales = "free_x") +
+  expand_limits(x = 1) +
   labs(x = "", y = "Proportion of items") + 
@@
 ```
-
-\begin{figure}[t]
-\includegraphics{figures/aum_prob_dist.pdf}
-\centering
-\caption{Distribution of phi estimates and probabilities that items show evidence of active updating.}
-\label{fig:phi_hist}
-\end{figure}
 
@@
 Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change.
+
+Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list.
 In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.
 
-The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
+The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 75 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
 That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
@@
 Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent change.
-We can only be really confident in detecting persistent change among a small minority of items, perhaps 1 in 5.
+We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5.
 This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.
@@
 
-Of the 111 items that showed evidence for active updating in the last section, 23 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
-Figure \@ref(fig:agegroupcomparison) plots the estimates of these 23 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
-
-```{r agegroupcomparison, message=FALSE, fig.width=6.8, fig.height=8, fig.cap="Comparison of phi estimates for individuals over and equal to to less than 30 years old."}
+Of the 108 items that showed evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
+Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
+
+```{r agegroupcomparison, message=FALSE, fig.width=6.8, fig.height=8, fig.cap="Comparison of phi estimates for individuals over and equal to or less than 30 years old, with 95 percent confidence intervals."}
 
 load(here("data", "age_results.Rdata"))
-load(here("data", "good_vars.Rdata"))
 
 bind_rows(age.results) %>%
-  mutate(ci.lower = estimate - 1.645 * std.error,
-         ci.upper = estimate + 1.645 * std.error) %>%
-  left_join(good_vars) %>%
+  mutate(ci.lower = estimate - 1.96 * std.error,
+         ci.upper = estimate + 1.96 * std.error) %>%
+  left_join(attitude_vars) %>%
   mutate(age = recode(age, "old"="Over 30", "young"="30 & under")) %>%
@@
   scale_color_manual(values = c("black", "gray")) + 
-  labs(x = "", y = "Phi estimate", color = "") +
+  labs(x = "", y = "Phi estimate (Updating)", color = "") +
   theme(axis.text=element_text(size=7),
@@
 
-\begin{figure}[t]
-\includegraphics{figures/age_differences_phi.pdf}
-\centering
-\caption{Comparison of $\phi$ estimates for individuals over and equal to to less than 30 years old.}
-\label{fig:age_group_comparison}
-\end{figure}
-
 The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents.
 These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die.
-These items tend to be in categories where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable for the rest of life.
-
-For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address pass discrimination, all evidence of active updating disappears for people over 30.
+These items tend to be in subject areas where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable over time.
+
+For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address past discrimination, all evidence of active updating disappears for people over 30.
 This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but beliefs quickly harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
 For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
-This pattern is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
-
-Eight of these items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
+This is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
+
+Eight items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
 These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, and views on whether aging parents should live with their children.
-Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on early in life.
-This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
-
-The remaining 89 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
+Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on while young.
+This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is somewhat unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
+
+The remaining 86 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
 However, this does not mean these show strong evidence of an active updating model.
@@
 
-While there is a high proportion (40 percent) of items that show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
+While 40 percent of items show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
 Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
@@
   scale_fill_manual(values = c("gray", "black")) + 
-  geom_text_repel(data = label.vars, aes(label = var.name), size = 3) + 
+  geom_text_repel(data = label_vars, aes(label = var.name), size = 3) + 
   theme_minimal() +
-  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
+  labs(x = "Phi estimate (Updating)", y = "Beta estimate (Consistency)", fill = "", shape = "") + 
   theme(axis.title=element_text(size=9))
@@
 
-\begin{figure}[t]
-\includegraphics{figures/phibeta_scatter.pdf}
-\centering
-\caption{Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint.}
-\label{fig:phibeta_scatter}
-\end{figure}
-
-Items showing evidence for active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
+Items showing evidence of active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
 A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
@@
 Which items show the strongest evidence for active updating?
-There is no way we can possibly discuss all 184 items in detail without the discussion becoming tedious.
+There is no way we can discuss all 183 items in detail without the discussion becoming tedious.
 Although we include $\phi$ estimates for all items in Appendix A, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
@@
 
-```{r summary, message=FALSE, fig.width=6.8, fig.height=9.0, fig.cap="Summary of phi estimates for all items, by topical group."}
+```{r summary, message=FALSE, fig.width=6.8, fig.height=9.0, fig.cap="Summary of phi estimates for all items, by topical group.
+Numbers in parentheses indicate the number of items in each topical group."}
 load(here("data", "summary_plot_data.Rdata"))
 
+cat_counts <- attitude_vars %>% group_by(small.cat) %>% summarise(n = n())
+
 plot.data %>%
+  left_join(cat_counts) %>%
+  mutate(small.cat = paste(small.cat, " (", n, ")", sep = "")) %>%
   ggplot(aes(x = reorder(small.cat, median))) + 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-
-```
-
-\begin{figure}[!ht]
-\includegraphics{figures/summary_plot.pdf}
-\centering
-\caption{$\phi$ estimates for all items by topical group.}
-\label{fig:summary}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+
+```
 
@@
 The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low in absolute terms.
-Only two groups of items have median $\phi$ values above 0.6: public spending and religious activity.
+Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity.
 In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]
@@
 
-Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that over half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
+Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that more than half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
 These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
-Views on these "core" issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood (and thus becomes eligible to participate in the GSS).
+Views on these issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood.
 This means that, for most of these items, social change must occur through cohort succession rather than through individual change.
@@
 There are several exceptions to this general pattern, however, even in categories with otherwise low $\phi$ values.
-The items with the largest values generally have one or more attributes in common.
+Items with the largest values generally have one or more attributes in common.
 We consider these attributes to give some general impressions of the pattern.
 
-Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them, perhaps making changes "stickier" than forms of personal culture.
+Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them.
 If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
@@
 Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
-Questions about religious activity, identity and (to a lesser extent) religious beliefs generally show some evidence for active updating.
-In some cases, this fits with the general pattern of external mechanisms (e.g., attendance and identification are largely public acts).
 
@@
 Likewise, all public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
-The change of administration and changing federal spending policies likely affected all spending items.
+The change of administration and changing federal spending policies likely affected these items.
 The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession).
@@
 The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
-This pattern is strongly consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
+This pattern is consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
 
@@
 
-```{r polvariables, message=FALSE, fig.width=7, fig.height=8, fig.cap="Phi and beta estimates for 43 political attitudes and identities."}
-left_join(res.df, good_vars) %>%
+```{r polvariables, message=FALSE, fig.width=7, fig.height=8, fig.cap="Phi (updating) and beta (consistency) estimates for 43 political attitudes and identities."}
+left_join(res.df, attitude_vars) %>%
   filter(small.cat %in% c("Politics & government", "Public spending",
@@
   mutate(small.cat = ifelse(small.cat %in% c("Race & immigration", 
-                                             "Guns, laws, crime & police")
-                            | var %in% c("prayer", "uswary"), 
+                                             "Guns, laws, crime & police") | 
+                              var %in% c("prayer", "uswary"), 
          "Misc. policy" , small.cat)) %>%
@@
   theme_bw() + 
-  labs(x = "Phi estimate", y = "Beta estimate", fill = "", shape = "") + 
+  labs(x = "Phi estimate (Updating)", y = "Beta estimate (Consistency)", fill = "", shape = "") + 
   theme(axis.title=element_text(size=9)) +
@@
 In contrast, questions asking about specific public spending priorities tend to have low consistency compared to other items.
-Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government –  display higher consistency than specific policy questions but weak evidence of active updating.
+Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government ("Gov't do more or less", "Gov't reduce inequality") – display higher consistency than specific policy questions but weak evidence of active updating.
 It is important to restate that for political views, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
@@
 In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating.
-This is notable, as public opinion and policies regarding both have shifted considerably in recent years, with many people assuming change in both is being driven by the same underlying process [@schnabelsevell].
+This is notable, as public opinion and policies regarding both have shifted considerably in recent years.
+Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell].
 Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana use have been driven primarily by cohort replacement.
@@
 Around 40 percent of all items show no evidence for updating.
-For the items that do show evidence of active updating, the overall rate of change in the population is likely low.
-Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to a stable baseline.
+For the items that do show evidence of active updating, the overall rate of persisting change in the population is likely low.
+For most items measured by the GSS, less than one percent of the population appears to make any persisting change in their views in a two-year period, and most changes appear to be short term or random deviations.
 What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
@@
 This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014).
-Because the GSS begins with 18-year-olds, we may lack power to detect updating that occurs even earlier, so this is probably an underestimate.
+Because the youngest respondents in the GSS are 18 years old, we may lack the ability to detect updating that occurs even earlier, so this is probably an underestimate.
 This may be the case for the roughly 70 items that show no evidence of persistent change over time.
@@
 
-Third, _are there systematic differences in item content between questions that are better described by each model?_ With 184 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
-We find patterns of responses that provide a range of support for diverse theories of attitude formation.
-At the same time, our results are consistent with some general expectations about patterns in updating.
+Third, _are there systematic differences in item content between questions that are better described by each model?_ With 183 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
+We find patterns of responses that provide a range of support for diverse theories of attitude formation, which we discuss below.
+At the same time, our results are consistent with some general expectations about updating.
 Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
@@
 
-What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area?
-In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly, through the mechanism of cohort succession.
+In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession.
 Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood.
@@
 In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating.
-Even for those items that did display strong evidence for active updating (e.g.
+Even those items that did display strong evidence for active updating (e.g.
 "whites rich or poor" and "whites work hard") were still hard to predict wave after wave.
 
-The overall pattern of results suggests that one reason attitudes are largely stable is because most issues simply do not reach the kind of salience necessary to shift people's opinions.
+Our results suggest that one reason attitudes are largely stable is because most issues simply do not reach the level of salience necessary to shift opinions.
 In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
@@
 This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level.
-Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages.
+Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors that encourage openness to change at different ages.
 Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes.
@@
 A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
-These respondents might express different positions from wave to wave, but in guessing what an individual will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
-While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
-With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still quite limited.
+Respondents might express different positions from wave to wave, but in guessing what a person will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
+While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals in our data changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
+With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still limited.
 Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.
 
-These results are consistent with theories that posit that, at least in the current era, individuals organize their beliefs around ideological identity [@boutylinevaisey; @converse], rather than some other organizing heuristic such as partisan identification, moral views, religion, or particular policy positions.
-It is also consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
+These results are consistent with theories that posit that, at least in the current era, ideological identification (view of oneself as liberal or conservative), rather than some other belief such as partisan identification, moral views, religion, or particular policy positions, is the "central" political belief [@boutylinevaisey; @converse].
+
+It is important not to overstate the role of this belief in forming other beliefs.
+While our findings suggest that individuals appear to bring their partisan identification in line with their ideological identification, there is limited evidence of adjusting other beliefs.
+This pattern is most consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
+
 Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time.
 This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.
-
 Similarly, our findings are not consistent with the idea that individuals adopt a partisan affiliation based on their social groupings and subsequently adjust their ideological commitments to conform to that [@greenetal].
-For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much persistent change as partisan identification, which is simply not true in our data.
-
-The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on most specific policy questions.
+For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much updating as partisan identification, which is not true in our data.
+
+The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile items such as partisan identity, abortion, and gay marriage, individuals appear to lack clear opinions on most specific policy questions.
 The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
@@
 
-That said, respondents are not wholly without consistent opinions.
-On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next.
-While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [-@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.
-
-The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents hold a general political identity and a few clear views on issues like abortion, generally attempt to align their identification to their views, and respond to elite opinion change when it provides clear signals.
+The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents, especially those over 30 years old, hold a general political identity and a few clear views on issues like abortion, attempt to align their partisan identification to their views, and respond to elite opinion change when it provides clear signals.
 It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates.
-Individuals tend to have more-or-less settled views, especially on core questions such as ideology, and report (with a high degree of inconsistency) more specific policy positions based on these.
 
@@
 The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change.
-It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that change is too rare in a sample of adults to measure accurately on the vast majority of items.
+It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that substantive change is too rare in a sample of adults to measure accurately on the vast majority of items.
 This strongly argues against using two-wave panels to measure attitude change, which do not allow researchers to separate persisting from non-persisting changes.
@@
 While we generally assume that lasting changes in attitudes are more likely to influence behavior, this is not necessarily true.
-Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying this kind of change is difficult.
-Even for items related to abortion, where there is almost no evidence of persistent change and the overall picture is one of stability of belief, individuals do bounce around from year-to-year.
+Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying it is difficult.
 
@@
 We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
-We see too much inconsistency in responses to most questions for models of commitment bias to be a major explanatory factor.
+We see too much inconsistency in responses for commitment bias to be a major explanatory factor.
 Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
@@
 In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
-Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly useless.
+Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly problematic.
 Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
+
+Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults.
 Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production" that give rise to persistent beliefs in adulthood.
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-
-```
-
-\begin{figure}[!ht]
-\includegraphics{figures/relig_phi_estimates.pdf}
-\centering
-\caption{$\phi$ estimates for items about religious activity and beliefs, social life, subjective SES, and suicide.}
-\label{fig:phi_relig}
-\end{figure}
-
-```{r poliphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about guns, law, crime and policing; politics and government; and public spending."}
+  labs(x = "", y = "Phi estimate (Updating)")
+
+```
+
+```{r poliphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about guns, law, crime and policing; politics and government; and public spending.
+The item 'Police can hit citizens', which has phi = .43, has been removed for ease of viewing"}
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-
-```
-
-\begin{figure}[!ht]
-\includegraphics{figures/poli_phi_estimates.pdf}
-\centering
-\caption{$\phi$ estimates for items about guns, law, crime and policing; politics and government; and public spending.}
-\label{fig:phi_poli}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+
+```
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-```
-
-\begin{figure}[!ht]
-\includegraphics{figures/civlib_phi_estimates.pdf}
-\centering
-\caption{$\phi$ estimates for items about civil liberties, confidence in leadership, health, morale, and social trust.}
-\label{fig:phi_civlib}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+```
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-
-```
-
-\begin{figure}[!ht]
-\includegraphics{figures/racgen_phi_estimates.pdf}
-\centering
-\caption{$\phi$ estimates for items about race, gender, sex, sexuality, and abortion.}
-\label{fig:phi_racgen}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+
+```
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-
-```
-
-\begin{figure}[t]
-\includegraphics{figures/coarse_relig.pdf}
-\centering
-\caption{$\phi$ estimates for coarsened questions about religion, subjective socioeconomic status, and morale.}
-\label{fig:coarse_relig}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+
+```
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-```
-
-\begin{figure}[t]
-\includegraphics{figures/coarse_poli.pdf}
-\centering
-\caption{$\phi$ estimates for coarsened questions about politics, government, sex and sexuality.}
-\label{fig:coarse_poli}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+```
 
@@
   expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate")
-```
-
-\begin{figure}[t]
-\includegraphics{figures/coarse_racgen.pdf}
-\centering
-\caption{$\phi$ estimates for coarsened questions about race and gender.}
-\label{fig:coarse_racgen}
-\end{figure}
+  labs(x = "", y = "Phi estimate (Updating)")
+```
 

### 2020-02-10 code/clean06.R

@@ -1,0 +1,216 @@
+
+#### Good GSS cleaning file
+
+#Load in the wide version of the 2006-2010 GSS panels
+gss06 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel06.dta")
+
+#convert the GSS from wide to long format
+gss.long06 <- gss06 %>% 
+  mutate(idnum = id_1) %>%
+  gather(v, value, abany_1:intrace_3) %>% 
+  separate(v, c("var", "wave")) %>% 
+  mutate(var = tolower(var)) %>%
+  arrange(idnum) %>% 
+  spread(var, value) %>%
+  mutate(idnum = paste("06-", idnum, sep = "")) 
+
+#recode some variables:
+gss.long06 <- gss.long06 %>%
+  dplyr::mutate(
+    #recode variables where "stay the same" is 3, making that 2
+    finalter = recode(finalter, "1"=1, "2"=3, "3"=2),
+    courts = recode(courts, "1"=1, "2"=3, "3"=2),
+    divlaw = recode(divlaw, "1"=1, "2"=3, "3"=2),
+    trust = recode(trust, "1"=1, "2"=3, "3"=2),
+    fair = recode(fair, "1"=1, "2"=3, "3"=2),
+    helpful = recode(helpful, "1"=1, "2"=3, "3"=2),
+    aged = recode(aged, "1"=1, "2"=3, "3"=2),
+    fear = ifelse(fear == 2, 0, fear),
+    #Reliten as coded goes "strong", "not very strong" and "somewhat strong"
+    #so recoded to more logical "strong", "somewhat strong", "not very strong"
+    reliten = recode(reliten, "1"=1, "2"=3, "3"=2, "4"=4),
+    #recode some reported values as missing
+    racopen = ifelse(racopen == 3, NA, racopen),
+    partyid = ifelse(partyid == 7, 3, partyid),
+    bible = ifelse(bible == 4, NA, bible),
+    #create some dummies from multinomial variables
+    neverrelact = ifelse(relactiv == 1, 1, 0),
+    relactiv = ifelse(relactiv == 1, 0, relactiv - 1),
+    suicide1 = ifelse(suicide1 == 2, 0, 1),
+    suicide2 = ifelse(suicide2 == 2, 0, 1),
+    suicide3 = ifelse(suicide3 == 2, 0, 1),
+    suicide4 = ifelse(suicide4 == 2, 0, 1),
[…]

### 2020-02-10 code/clean08.R

@@ -1,0 +1,213 @@
+
+#### Good GSS cleaning file
+gss08 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel08.dta")
+
+
+#convert the GSS from wide to long format
+gss.long08 <- gss08 %>% select(-vetyear2) %>%
+  mutate(idnum = id_1) %>%
+  gather(v, value, abany_1:letin1a_3) %>% 
+  separate(v, c("var", "wave")) %>% 
+  mutate(var = tolower(var)) %>%
+  arrange(idnum) %>% 
+  spread(var, value) %>%
+  mutate(idnum = paste("08-", idnum, sep = "")) 
+
+#recode some variables:
+gss.long08 <- gss.long08 %>%
+  dplyr::mutate(
+    #recode variables where "stay the same" is 3, making that 2
+    finalter = recode(finalter, "1"=1, "2"=3, "3"=2),
+    courts = recode(courts, "1"=1, "2"=3, "3"=2),
+    divlaw = recode(divlaw, "1"=1, "2"=3, "3"=2),
+    trust = recode(trust, "1"=1, "2"=3, "3"=2),
+    fair = recode(fair, "1"=1, "2"=3, "3"=2),
+    helpful = recode(helpful, "1"=1, "2"=3, "3"=2),
+    aged = recode(aged, "1"=1, "2"=3, "3"=2),
+    fear = ifelse(fear == 2, 0, fear),
+    #Reliten as coded goes "strong", "not very strong" and "somewhat strong"
+    #so recoded to more logical "strong", "somewhat strong", "not very strong"
+    reliten = recode(reliten, "1"=1, "2"=3, "3"=2, "4"=4),
+    #recode some reported values as missing
+    racopen = ifelse(racopen == 3, NA, racopen),
+    partyid = ifelse(partyid == 7, 3, partyid),
+    bible = ifelse(bible == 4, NA, bible),
+    #create some dummies from multinomial variables
+    neverrelact = ifelse(relactiv == 1, 1, 0),
+    relactiv = ifelse(relactiv == 1, 0, relactiv - 1),
+    suicide1 = ifelse(suicide1 == 2, 0, 1),
+    suicide2 = ifelse(suicide2 == 2, 0, 1),
+    suicide3 = ifelse(suicide3 == 2, 0, 1),
+    suicide4 = ifelse(suicide4 == 2, 0, 1),
+    socbar = 8-socbar,
[…]

### 2020-02-10 code/clean10.R

@@ -1,0 +1,518 @@
+
+#### Good GSS cleaning file
+
+gss10 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel10.dta")
+
+#convert the GSS from wide to long format
+gss.long10 <- gss10 %>% 
+  mutate(idnum = id_1) %>%
+  gather(v, value, abany_1:letin1a_3) %>% 
+  separate(v, c("var", "wave")) %>% 
+  mutate(var = tolower(var)) %>%
+  arrange(idnum) %>% 
+  spread(var, value) %>%
+  mutate(idnum = paste("10-", idnum, sep = "")) 
+
+#recode some variables:
+gss.long10 <- gss.long10 %>%
+  dplyr::mutate(
+    #recode variables where "stay the same" is 3, making that 2
+    finalter = recode(finalter, "1"=1, "2"=3, "3"=2),
+    courts = recode(courts, "1"=1, "2"=3, "3"=2),
+    divlaw = recode(divlaw, "1"=1, "2"=3, "3"=2),
+    trust = recode(trust, "1"=1, "2"=3, "3"=2),
+    fair = recode(fair, "1"=1, "2"=3, "3"=2),
+    helpful = recode(helpful, "1"=1, "2"=3, "3"=2),
+    aged = recode(aged, "1"=1, "2"=3, "3"=2),
+    fear = ifelse(fear == 2, 0, fear),
+    #Reliten as coded goes "strong", "not very strong" and "somewhat strong"
+    #so recoded to more logical "strong", "somewhat strong", "not very strong"
+    reliten = recode(reliten, "1"=1, "2"=3, "3"=2, "4"=4),
+    #recode some reported values as missing
+    racopen = ifelse(racopen == 3, NA, racopen),
+    partyid = ifelse(partyid == 7, 3, partyid),
+    bible = ifelse(bible == 4, NA, bible),
+    #create some dummies from multinomial variables
+    neverrelact = ifelse(relactiv == 1, 1, 0),
+    relactiv = ifelse(relactiv == 1, 0, relactiv - 1),
+    suicide1 = ifelse(suicide1 == 2, 0, 1),
+    suicide2 = ifelse(suicide2 == 2, 0, 1),
+    suicide3 = ifelse(suicide3 == 2, 0, 1),
+    suicide4 = ifelse(suicide4 == 2, 0, 1),
+    socbar = 8-socbar,
[…]

### 2020-02-10 code/phi_beta_model.R

@@ -1,0 +1,272 @@
+
+### What is this file?
+#This code uses the model outlined in Kiley and Vaisey (2020) to estimate
+#the presence of persistent change in the GSS panels.
+
+#### Load in a long version of the GSS panels.
+load("~/Desktop/panel_change/data/gsslong.Rdata")
+load("~/Desktop/panel_change/data/attitude_vars.Rdata")
+
+#Make a list of variables in the gss
+gss_vars <- attitude_vars$var
+
+#Function that creates a subset of the panel that is just a variable of 
+#interest, taken as a character vector, and omits missing cases
+cleanGSS <- function(var) {
+  data <- gss.long %>%
+    select(idnum, ds, minage, wtpannr123, wave, var) %>%
+    mutate(wave = paste("w", wave, sep = "")) %>%
+    spread(wave, var) %>%
+    na.omit() %>% filter(!is.infinite(minage)) %>%
+    mutate(var = var)
+  return(data)
+}
+
+## Loop that estimates the non-linear model for all variables in the 
+## list. Stores those in a list named "results"
+results <- vector(mode = "list", length = length(gss_vars))
+for (i in 1:length(gss_vars)) {
+  
+  var <- gss_vars[i]
+  
+  #extract focal variable
+  all.data <-  cleanGSS(var) %>%
+    mutate(y1 = w1, y2 = w2, y3 = w3) %>%
+    select(-c(w1, w2, w3)) 
+
+  #Unconstrained model
+  nls1 <- nls(y3 ~ a + b*p*y2 + b*(1-p)*y1,
+              data = all.data,
+              start = list(a = 0, b = 1, p = 0),
+              weights = wtpannr123)
+  bic1 <- BIC(nls1)
+  
+  #Model that constrains phi to .5
+  nls2 <- nls(y3 ~ a + b*.5*y2 + b*(.5)*y1,
+              data = all.data,
+              start = list(a = 0, b = 1),
+              weights = wtpannr123)
+  bic2 <- BIC(nls2)
+
+  #store results in list
+  results[[i]] <- data.frame(var = var, bic1 = bic1, bic2 = bic2,
+                             p = coef(nls1)[3], 
+                             b1 = coef(nls1)[2], b2 = coef(nls2)[2])
+  
+  #print to spot errors
+  print(paste(var, i, sep = " "))
+  
+  #clean out before the next iteration
[…]

### 2020-02-12 manuscript9.Rmd
final edits

@@
 
-On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this does not need to be the case.
+On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case.
 The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse], and suggest that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
+These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse].
+Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
 As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.
@@
 
-Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot.
+Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot.
 Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
@@
 Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list.
-In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.
+In other words, even for items that show strong evidence of some active updating, the overall amount of attitude change in the population is likely small.
 
@@
 To answer our first question, however, we need only focus on the overall distribution.
-Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1, meaning that for almost all items, measurement error or non-persistent change tends to be much more common than persistent change.
+Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1.
+This means that for almost all items, measurement error or non-persistent change tend to be much more common than persistent change.
 We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5.
@@
 
-Our second research question asks whether there is evidence that younger respondents are updating their views more than older respondents.
+Our second research question asks whether there is evidence that younger respondents update their views more than older respondents.
 Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
 
-Of the 108 items that showed evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
+Of the 108 items that showed any evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
 Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
@@
 There is a lot to process, even in this summary figure.
-The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low in absolute terms.
+The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low.
 Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity.
@@
 Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell].
-Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana use have been driven primarily by cohort replacement.
+Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.
 
@@
 In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession.
-Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood.
+Most beliefs about gender roles, sexual morality, and abortion appear to be settled by early adulthood.
 The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time.
@@
 Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults.
-Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production" that give rise to persistent beliefs in adulthood.
+Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production"—childhood and adolescence—that give rise to persistent beliefs in adulthood.
 

### 2020-02-14 manuscript9.Rmd
final submission revisions

@@
 This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].
+These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-declarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].
 
@@
 
-```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
+[Figure \@ref(fig:dag) about here]
+
+```{r}
 library(ggdag)
 library(dagitty)
+```
+
+## Active Updating Model
+
+Equation \@ref(eq:AR) represents the active updating model.
+The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$).
+There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.
+Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state.
+This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].
+In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.
+
+More informally, this AUM posits a person who is updating her views in the face of social experience.
+There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
+Following Gross's [-@gross09, pp.
+367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."
+
+Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
+When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model.
+When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time.
+Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.
+
+Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities [@gross09; @degloma].
+As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.
+
+The AUM makes no assumption about the distribution of $\nu_{it}$.
+Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$.
+This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.
+
+Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
+In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.
+
+## Settled Dispositions Model
+
+Equation \@ref(eq:FE) represents the settled dispositions model.
+Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$).
+Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$.
+As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses.
+Such considerations are thus temporary influences only.
+At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp.
+66ff].
+In other words, updating happened, but it happened in the past, prior to the time of the study.
+In this sense, $U_i$ reflects the "habitus."
+
+The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise].
+The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
+These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder].
+Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.
+
+On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case.
+The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
+These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse].
+Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
+As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.
+
+Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time.
+That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time.
+The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.
+
+Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level.
+For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported.
+Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].
+
+The two general models thus differ fundamentally on their emphasis on the character of personal change.
+In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time.
+In the SDM, changes tend to revert because each person returns to their baseline.
+We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.
+
+## Consistency in Response
+
+Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
+This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change.
+Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development.
+However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.
+As such, we designate a second term, _consistency_, to reflect this second dimension of change.
+_Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.
+
+## Toward Theoretical Synthesis
+
+We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating.
+Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.
+
+The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern.
+This disagreement thus forms the primary structure of our analysis.
+However, it is possible that some items display active updating while others do not.
+There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.
+
+_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman].
+To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
+
+Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].
+
+_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein].
+This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe].
+In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space.
+Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time.
+They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
+442].
+
+The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition.
+Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency.
+On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman].
+In other words, liberal Republicans should become liberal Democrats, not conservative Republicans.
+This means that partisan identification should display high levels of updating.
+Again, this does not mean that many people change, only that changes tend to persist.
+
+Alongside ideological identification, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman].
+For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy.
+Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently.
+There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.
+
+Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs.
+However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse].
+However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.
+
+_Public vs.
+Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
+The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p.
+91].
+This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances.
+However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].
+
+Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse].
+This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time.
+However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population.
+In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights.
+Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
+For this issue, we expect to find evidence of active updating throughout the population, not just among young people.
+
+Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot.
+Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
+
+As we noted previously, we do not expect that all items will align with just one model.
+If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
+Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
+We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
+
+# Research Questions and Expectations
+
+With these considerations in mind, we ask the following questions:
+
+First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
+This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.
+
+Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18).
+However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.
+
+Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors.
+As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.
+
+# Analytic Strategy
+
+To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model.
+This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012).
+Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model.
+We discuss item selection below.
+While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.
+
+## Statistical Model
+
+### Basic Models
+
+Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time.
+We first consider measuring the amount of active updating.
+Our two different models make different predictions in the three-wave panel context.
+The AUM makes the following prediction for wave 3:
+
+\begin{equation}
+  E(y_{i3}) = y_{i2}
+  (\#eq:AUM2)
+\end{equation}
+
+That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2.
+If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.
+
+The SDM makes the following prediction:
+
+\begin{equation}
+  E(y_{i3}) = U_i
+  (\#eq:SDM2)
+\end{equation}
+
+Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:
+
+\begin{equation}
+  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
+  (\#eq:SDM2B)
+\end{equation}
+
+That is, the SDM predicts that the average response of previous waves is the best predictor of the next response.
+If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.
+
+### Combined Model
+
+Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor.
+If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition.
+Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$.
+If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model.
+However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating.
+We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:
+
+\begin{equation}
+  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
+  (\#eq:MODEL)
+\end{equation}
+
+Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
+If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B).
+If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.
+
+Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating.
+We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
+If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$.
+If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
+
+### Comparison to Other Approaches
+
+Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
+Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS responses.
+This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
+These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier.
+While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves.
+This approach also combines persisting and non-persisting change into two similar but distinct kinds of change: "structural" and "non-structural," both of which can be persisting and non-persisting.
+This distinction, while important for some theoretical questions, is not our focus.
+
+A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief.
+This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
+Since we at times invoke both these models, we include in our analysis several composite scales of related items.
+If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in their levels of active updating.
+
+Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
+While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
+One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
+If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 would become a better predictor of wave 3.
+
+We do not view this as a problem for our theoretical models.
+If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
+This might lead to a higher estimate of active updating than we might observe in a population that did not take the survey, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update beliefs over time.
+
+A second form of panel conditioning posits that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time, even if they actually change.
+In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change.
+This would be problematic for our study, as it would under-estimate the amount of real change in the population.
+However, if this is the case, we should observe no active updating in responses and high levels of consistency, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.
+
+Finally, a number of approaches exist for evaluating theoretical process of belief formation and change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
+These tools are well designed to address the questions they set out to answer.
+However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
+
+### Limitations of the Method
+
+Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach and therefore limit the conclusions we can reach.
+The first is that our model is designed to allocate variance explained to each of the prior waves rather than to assign probabilities to each data-generating process.
+Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.
+
+The second is measurement error, which is a form of non-persisting change.
+For some researchers, measurement error represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
+For other theories, measurement error reflects individuals' inability to accurately report their response.
+It is also possible that measurement error reflects errors of selection and interpretation, such as misunderstanding the question or incorrect coding.
+
+Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), $\phi$ estimates will be biased toward .5, since responses with error would be random departures from baseline.
+There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
+On the other hand, previous studies of reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
+Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
+
+The third challenge of our analysis is that we focus on predicting wave 3.
+If individuals have a high likelihood of changing between waves 2 and 3, our ability to predict responses at wave 3 will be limited and $\beta$ will be low.
+Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3.
+If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error, and it may not be reasonable to consider this sort of change "persistent."
+
+In addition to these three challenges, there are two forms of change that our model is not well designed to account for.
+The first is a uni-directional shock to the population.
+Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates.
+The second is change in the variance of responses.
+If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will be absorbed into $\beta$ but not enter into $\phi$.
+
+Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error.
+Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model.
+We can only detect whether there is any evidence of persisting belief changes, and therefore whether there is any evidence that active updating is taking place in the population.
+
+## Analysis Steps
+
+Our analysis proceeds in three steps to answer our three research questions.
+First, we evaluate the overall evidence in support of the active updating model.
+To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
+We calculate the posterior probability that the model with the free parameter fits the data better.
+If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.[^link]
+
+[^link]: All data used in this analysis is publicly available through the General Social Survey at the National Opinion Research Center at http://gss.norc.org/.
+R code for how we cleaned the GSS panels, all analyses performed in this paper, and all figures created in this paper can be found at http://github.com/krkiley/panel_change.
+
+Second, for variables that show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is concentrated among younger respondents.
+To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff.
+Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45.
+We test a range of cutoffs to ensure robustness of the overall pattern to specific ages.
+
+Finally we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
+Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be inductive.
+
+## Item Selection
+
+To test our model on as broad a range of items as possible, we sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
+We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
+We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on subject material.
+Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about government spending, and questions about social life.
+
+We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust.
+This includes a six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
+We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are.
+We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman].
+Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy.
+We combine four parallel questions about how frequently individuals socialize to create a "social life" scale.
+We combine four questions about support for suicide under different circumstances.
+We also created a scale of support for police use of violence against criminal suspects by combining five binary questions about the conditions under which individuals support police use of violence.
+
+In total, we test the model on 183 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves.
+Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.
+
+[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options.
+These results are reported in Appendix B.
+
+# Results
+
+```{r, message = FALSE, warn = FALSE}
+load(here("data", "results_df.Rdata"))
+load(here("data", "attitude_vars.Rdata"))
+
+label_vars <- res.df %>%
+  filter(var %in% c("intlblks", "confed", "selfhunt", "socbar")) %>%
+  left_join(attitude_vars)
+
+#create a list of variables that display updating
+aum_vars <- res.df %>% filter(aum_pref == 1)
+aum_vars <- aum_vars$var
+
+```
+
+Our model estimates two parameters of interest for each GSS item: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
+High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
+Our measure of active updating, $\phi$, captures the relative proportion of wave 3 variance predicted by wave 2.
+If responses are generated through a true settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
+As the evidence of active updating increases, $\phi$ will increase toward 1.
+Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3.
+
+## Evidence for Active Updating
+
+To evaluate the evidence in favor of the active updating model, we compare for all 183 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
+If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.
+
+Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 183 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.
+
+[Figure \@ref(fig:phihist) about here.]
+
+On the left side of the figure, we see that the majority of $\phi$ estimates fall between .5 and .6, meaning that wave 2 is only a slightly better predictor than wave 1 for most items.
+This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.
+
+To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates.
+Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2.
+Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3.
+Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3.
+Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change.
+
+Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list.
+In other words, even for items that show strong evidence of some active updating, the overall amount of attitude change in the population is likely small.
+
+The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 75 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
+That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
+This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.
+
+We will discuss in more detail below how different items perform.
+To answer our first question, however, we need only focus on the overall distribution.
+Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1.
+This means that for almost all items, measurement error or non-persistent change tend to be much more common than persistent change.
+We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5.
+This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.
+
+## Age Heterogeneity
+
+Our second research question asks whether there is evidence that younger respondents update their views more than older respondents.
+Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
+
+Of the 108 items that showed any evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
+Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
+
+[Figure \@ref(fig:agegroupcomparison) about here.]
+
+The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents.
+These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die.
+These items tend to be in subject areas where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable over time.
+
+For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address past discrimination, all evidence of active updating disappears for people over 30.
+This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but beliefs quickly harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
+For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
+This is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
+
+Eight items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
+These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, and views on whether aging parents should live with their children.
+Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on while young.
+This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is somewhat unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
+
+The remaining 86 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
+However, this does not mean these show strong evidence of an active updating model.
+These items may simply be susceptible to updating for a small proportion of the population.
+
+## Item heterogeneity
+
+While 40 percent of items show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
+Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
+
+Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items shaded by whether they preferred the $\phi = .5$ constraint or not and label a few items that stand out.
+Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.
+
+[Figure \@ref(fig:phibetascatter) about here.]
+
+Items showing evidence of active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
+A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
+
+Which items show the strongest evidence for active updating?
+There is no way we can discuss all 183 items in detail without the discussion becoming tedious.
+Although we include $\phi$ estimates for all items in Appendix A, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
+We constrain items that showed no evidence of active updating to $\phi = .5$.
+In addition to showing the median and interquartile range of each distribution, the figure also highlights the item in each group that shows the greatest degree of evidence for active updating.
+
+[Figure \@ref(fig:summary) about here.]
+
+There is a lot to process, even in this summary figure.
+The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low.
+Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity.
+In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]
+
+[^error]: While we cannot control for measurement error in our analysis, we can take steps to mitigate its impact.
+Appendix B presents results comparing items with more than three scale points to coarsened versions of these question with either two or three scale points.
+As noted previously, low measurement error might be a reasonable assumption for some items.
+Previous studies using different approaches to measuring the reliability of survey reports suggest that some items captured in our study, such as whether a person owns a gun, are measured with a high degree of reliability ($> 0.9$) [@houthastings].
+For other items, such as confidence in the leadership of major companies ($\phi = .58$), reliability might be as low as 0.5.
+There is very little correlation ($\rho = .165$) between our $\phi$ estimates and reliability estimates.
+
+Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that more than half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
+These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
+Views on these issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood.
+This means that, for most of these items, social change must occur through cohort succession rather than through individual change.
+
+There are several exceptions to this general pattern, however, even in categories with otherwise low $\phi$ values.
+Items with the largest values generally have one or more attributes in common.
+We consider these attributes to give some general impressions of the pattern.
+
+Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them.
+If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
+This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, which are more nebulous questions that display less active updating.
+Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
+Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
+
+Other high-$\phi$ items have a _changing referent_.
+That is, although the item wording is the same, the object to which the question refers may change between survey waves.
+The most obvious example of this is the item about confidence in the executive branch of the federal government (which has the highest $\phi$ value of all items in the analysis).
+The president changed between the 2008 and 2010 waves of the GSS, meaning that the question no longer referred to the same administration.
+If we generate estimates for this item for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$).
+In the 2006-10 panel, waves 1 and 2 have almost no predictive power ($\beta = .10$).
+For the 2010-14 panel, which takes place entirely during the Obama administration, $\phi$ moves much closer to .5 and predictive power increases ($\phi = .57$, $\beta = .65$).
+
+Likewise, all public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
+The change of administration and changing federal spending policies likely affected these items.
+The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession).
+If the environment is changing we should see exactly this sort of pattern.
+
+Perhaps the most striking pattern in our findings pertains to questions about _gay rights_.
+There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating.
+Questions about civil liberties for gays and about gay marriage are the highest in their categories.
+The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
+This pattern is consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
+
+A few items lack external mechanisms, represent forms of private culture, and did not achieve salience during this period, but still display active updating, including some views on abortion and the morality of different forms of sex.
+We address these items in greater detail in the discussion section.
+
+## Political Beliefs
+
+Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, and because many of these beliefs display evidence of active updating, we examine these beliefs in greater detail than other items.
+Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for questions about political identity, the role of government, and specific policies.
+We break these items up into categories and remove a few items for ease of viewing.
+
+[Figure \@ref(fig:polvariables) about here.]
+
+There are several notable features of the figure.
+First, political party affiliation is a clear outlier, with greater consistency and greater active updating than other items.
+In contrast, questions asking about specific public spending priorities tend to have low consistency compared to other items.
+Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government ("Gov't do more or less", "Gov't reduce inequality") – display higher consistency than specific policy questions but weak evidence of active updating.
+It is important to restate that for political views, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
+
+While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently.
+Views on abortion, especially when considered as a scale, are consistently reported.
+As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change.
+As discussed previously, views on gay marriage display high levels of active updating.
+In contrast, the item asking about support for legalizing marijuana use is reported with consistency but weak active updating.
+This is notable, as public opinion and policies regarding both have shifted considerably in recent years.
+Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell].
+Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.
+
+# Discussion and Conclusions
+
+This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making.
+These two models of individual change are implicit in analyses of belief change at the society level, but they are rarely compared empirically.
+We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models.
+We now revisit our research questions to summarize what we have learned.
+
+First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
+Around 40 percent of all items show no evidence for updating.
+For the items that do show evidence of active updating, the overall rate of persisting change in the population is likely low.
+For most items measured by the GSS, less than one percent of the population appears to make any persisting change in their views in a two-year period, and most changes appear to be short term or random deviations.
+What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
+Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is.
+But the average level of updating for most views appears to be low.
+
+Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ For a limited subset of items, there seems to be evidence that younger respondents are updating their views whereas older respondents are not.
+This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014).
+Because the youngest respondents in the GSS are 18 years old, we may lack the ability to detect updating that occurs even earlier, so this is probably an underestimate.
+This may be the case for the roughly 70 items that show no evidence of persistent change over time.
+Overall, however, there is significant heterogeneity in the relationship between age and attitude and behavior change.
+
+Third, _are there systematic differences in item content between questions that are better described by each model?_ With 183 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
+We find patterns of responses that provide a range of support for diverse theories of attitude formation, which we discuss below.
+At the same time, our results are consistent with some general expectations about updating.
+Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
+Items assumed to be "peripheral" beliefs, such as specific policy questions, also showed evidence for updating and low consistency.
+Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18.
+
+There are a handful of items that demonstrate active updating and enough consistency to merit further consideration, but are not easily explained by theoretical mechanisms outlined so far.
+These include questions about abortion in the case of rape or poverty, the morality of premarital and teen sex, the ideal number of children in a family, and whether children should obey adults or think for themselves.
+These items did not achieve high salience during the time of the study and there were not, to the best of our knowledge, clear changes in elite opinion on these issues.
+This suggests additional mechanisms that can drive persisting change at the individual level, such as exposure to these ideas through mass media or by directly interacting with people in these groups.
+In the case of questions about children and family structures, it might be the experience of having children that leads to changes in these attitudes.
+It is important to note that even though these items demonstrate active updating, it is a small proportion of the population that are changing their views during the study window.
+
+The overall pattern of results, while diverse and at times hard to reconcile, does shed light on several theoretical debates. We now consider these implications.
+
+## Implications for Cultural Sociology
+
+In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession.
+Most beliefs about gender roles, sexual morality, and abortion appear to be settled by early adulthood.
+The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time.
+In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating.
+Even those items that did display strong evidence for active updating (e.g.
+"whites rich or poor" and "whites work hard") were still hard to predict wave after wave.
+
+Our results suggest that one reason attitudes are largely stable is because most issues simply do not reach the level of salience necessary to shift opinions.
+In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
+By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period.
+The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
+This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.
+
+While the dominant pattern is stability during adulthood, the results suggest greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories.
+The most prominent and well supported theories of attitude change suggest a peak of susceptibility to attitude change early in adulthood and either a rapid or gradual decline in attitude change with age [@alwinkrosnick; @visserkrosnick].
+An additional view, the "life stages" hypothesis, suggests a similar pattern with a surge in attitude change late in life [@visserkrosnick].
+While this age-related decline appears to be true for many political attitudes, the pattern is not nearly as consistent for other kinds of attitudes.
+Some items, such as views on the Bible, suggest equal openness to attitude change as individuals age.
+Others, such as views on most abortion questions, suggest that early adult socialization is so strong that lifetime opinions are settled by the time most people reach even 18 years old.
+
+For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
+This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level.
+Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors that encourage openness to change at different ages.
+Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes.
+Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick].
+Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].
+
+## Implications for Political Sociology
+
+A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
+Respondents might express different positions from wave to wave, but in guessing what a person will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
+While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals in our data changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
+With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still limited.
+Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.
+
+These results are consistent with theories that posit that, at least in the current era, ideological identification (view of oneself as liberal or conservative), rather than some other belief such as partisan identification, moral views, religion, or particular policy positions, is the "central" political belief [@boutylinevaisey; @converse].
+
+It is important not to overstate the role of this belief in forming other beliefs.
+While our findings suggest that individuals appear to bring their partisan identification in line with their ideological identification, there is limited evidence of adjusting other beliefs.
+This pattern is most consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
+
+Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time.
+This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.
+Similarly, our findings are not consistent with the idea that individuals adopt a partisan affiliation based on their social groupings and subsequently adjust their ideological commitments to conform to that [@greenetal].
+For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much updating as partisan identification, which is not true in our data.
+
+The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile items such as partisan identity, abortion, and gay marriage, individuals appear to lack clear opinions on most specific policy questions.
+The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
+What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."
+
+The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents, especially those over 30 years old, hold a general political identity and a few clear views on issues like abortion, attempt to align their partisan identification to their views, and respond to elite opinion change when it provides clear signals.
+It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates.
+
+## Methodological Implications
+
+Methodologically, our results highlight the challenges of evaluating population-wide attitude change using short-term panel studies.
+The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change.
+It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that substantive change is too rare in a sample of adults to measure accurately on the vast majority of items.
+This strongly argues against using two-wave panels to measure attitude change, which do not allow researchers to separate persisting from non-persisting changes.
+
+The fact that persistent change is practically nonexistent for many items bolsters the case for using repeated survey responses to measure the reliability of survey items [@alwin; @houthastings], since it is often a valid assumption that the underlying view is unchanging.
+At the same time, our results call for greater focus on methodological tools that can separate short-term attitude change from measurement error.
+While we generally assume that lasting changes in attitudes are more likely to influence behavior, this is not necessarily true.
+Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying it is difficult.
+
+We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation.
+One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.
+
+We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
+We see too much inconsistency in responses for commitment bias to be a major explanatory factor.
+Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
+499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p.
+500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p.
+501], tend to show low active updating and low consistency.
+
+We cannot (and would not want to) rule out the possibility that panel conditioning is taking place in the GSS.
+We believe it is worthwhile to explore these same GSS panels for evidence of panel conditioning.
+However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
+
+Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.
+Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful.
+In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
+Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly problematic.
+Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
+
+Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults.
+Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production"—childhood and adolescence—that give rise to persistent beliefs in adulthood.
+
+\pagebreak
+
+# Appendix A: $\phi$ values for all variables
+
+Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) plot $\phi$ estimates for all items included in the analysis, grouped by subject material.
+
+[Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) about here.]
+
+```{r loadfiguredata}
+load(here("data", "figure_data.Rdata"))
+```
+
+# Appendix B: Coarsened variable estimates
+
+```{r loadcoarse}
+load(here("data", "coarse_vars.Rdata"))
+
+```
+
+Estiates of $\phi$ for coarsened items are presented in Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen).
+If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree).
+If scales included a clear midpoint, we included that and coarsened responses on either side.
+For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median.
+If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.
+
+[Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen) about here.]
+
+Generally speaking, most items with no evidence of active updating continue to have no evidence of active updating when coarsened.
+For items that showed evidence for active updating, the coarsened estimates tended to be very similar to the uncoarsened estimates, suggesting that persisting and non-persisting changes happen about as often between ends of these scales as they do within ends of these scales.
+
+There are a couple notable departures from this general pattern.
+Several political questions -- general political views, views on whether the government should reduce inequality, and views on whether the government should help blacks -- show decreased evidence of persisting change when coarsened.
+This suggests that changes around the midpoint of the scale tend to be measurement error, a finding that is consistent with previous work suggesting that individuals without settled political views tend to choose points around the middle of the response scale [@converse].
+
+\pagebreak
+
+# References
+
+```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
 aum <- 
@@
 gridExtra::grid.arrange(aum_dag, sdm_dag, ncol = 2)
-
-```
-
-## Active Updating Model
-
-Equation \@ref(eq:AR) represents the active updating model.
-The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$).
-There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.
-Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state.
-This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].
-In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.
-
-More informally, this AUM posits a person who is updating her views in the face of social experience.
-There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
-Following Gross's [-@gross09, pp.
-367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."
-
-Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
-When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model.
-When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time.
-Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.
-
-Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities [@gross09; @degloma].
-As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.
-
-The AUM makes no assumption about the distribution of $\nu_{it}$.
-Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$.
-This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.
-
-Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
-In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.
-
-## Settled Dispositions Model
-
-Equation \@ref(eq:FE) represents the settled dispositions model.
-Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$).
-Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$.
-As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses.
-Such considerations are thus temporary influences only.
-At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp.
-66ff].
-In other words, updating happened, but it happened in the past, prior to the time of the study.
-In this sense, $U_i$ reflects the "habitus."
-
-The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise].
-The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
-These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder].
-Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.
-
-On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case.
-The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse].
-Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
-As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.
-
-Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time.
-That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time.
-The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.
-
-Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level.
-For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported.
-Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].
-
-The two general models thus differ fundamentally on their emphasis on the character of personal change.
-In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time.
-In the SDM, changes tend to revert because each person returns to their baseline.
-We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.
-
-## Consistency in Response
-
-Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
-This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change.
-Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development.
-However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.
-As such, we designate a second term, _consistency_, to reflect this second dimension of change.
-_Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.
-
-## Toward Theoretical Synthesis
-
-We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating.
-Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.
-
-The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern.
-This disagreement thus forms the primary structure of our analysis.
-However, it is possible that some items display active updating while others do not.
-There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.
-
-_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman].
-To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
-
-Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].
-
-_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein].
-This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe].
-In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space.
-Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time.
-They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
-442].
-
-The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition.
-Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency.
-On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman].
-In other words, liberal Republicans should become liberal Democrats, not conservative Republicans.
-This means that partisan identification should display high levels of updating.
-Again, this does not mean that many people change, only that changes tend to persist.
-
-Alongside ideological identificaton, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman].
-For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy.
-Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently.
-There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.
-
-Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs.
-However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse].
-However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.
-
-_Public vs.
-Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
-The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p.
-91].
-This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances.
-However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].
-
-Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse].
-This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time.
-However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population.
-In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights.
-Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
-For this issue, we expect to find evidence of active updating throughout the population, not just among young people.
-
-Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot.
-Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
-
-As we noted previously, we do not expect that all items will align with just one model.
-If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
-Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
-We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
-
-# Research Questions and Expectations
-
-With these considerations in mind, we ask the following questions:
-
-First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
-This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.
-
-Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18).
-However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.
-
-Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors.
-As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.
-
-# Analytic Strategy
-
-To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model.
-This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012).
-Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model.
-We discuss item selection below.
-While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.
-
-## Statistical Model
-
-### Basic Models
-
-Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time.
-We first consider measuring the amount of active updating.
-Our two different models make different predictions in the three-wave panel context.
-The AUM makes the following prediction for wave 3:
-
-\begin{equation}
-  E(y_{i3}) = y_{i2}
-  (\#eq:AUM2)
-\end{equation}
-
-That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2.
-If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.
-
-The SDM makes the following prediction:
-
-\begin{equation}
-  E(y_{i3}) = U_i
-  (\#eq:SDM2)
-\end{equation}
-
-Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:
-
-\begin{equation}
-  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
-  (\#eq:SDM2B)
-\end{equation}
-
-That is, the SDM predicts that the average response of previous waves is the best predictor of the next response.
-If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.
-
-### Combined Model
-
-Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor.
-If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition.
-Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$.
-If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model.
-However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating.
-We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:
-
-\begin{equation}
-  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
-  (\#eq:MODEL)
-\end{equation}
-
-Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
-If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B).
-If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.
-
-Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating.
-We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
-If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$.
-If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
-
-### Comparison to Other Approaches
-
-Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
-Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS responses.
-This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
-These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier.
-While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves.
-This approach also combines persisting and non-persisting change into two similar but distinct kinds of change: "structural" and "non-structural," both of which can be persisting and non-persisting.
-This distinction, while important for some theoretical questions, is not our focus.
-
-A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief.
-This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
-Since we at times invoke both these models, we include in our analysis several composite scales of related items.
-If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in their levels of active updating.
-
-Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
-While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
-One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
-If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 would become a better predictor of wave 3.
-
-We do not view this as a problem for our theoretical models.
-If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
-This might lead to a higher estimate of active updating than we might observe in a population that did not take the survey, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update beliefs over time.
-
-A second form of panel conditioning posits that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time, even if they actually change.
-In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change.
-This would be problematic for our study, as it would under-estimate the amount of real change in the population.
-However, if this is the case, we should observe no active updating in responses and high levels of consistency, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.
-
-Finally, a number of approaches exist for evaluating theoretical process of belief formation and change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
-These tools are well designed to address the questions they set out to answer.
-However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
-
-### Limitations of the Method
-
-Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach and therefore limit the conclusions we can reach.
-The first is that our model is designed to allocate variance explained to each of the prior waves rather than to assign probabilities to each data-generating process.
-Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.
-
-The second is measurement error, which is a form of non-persisting change.
-For some researchers, measurement error represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
-For other theories, measurement error reflects individuals' inability to accurately report their response.
-It is also possible that measurement error reflects errors of selection and interpretation, such as misunderstanding the question or incorrect coding.
-
-Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), $\phi$ estimates will be biased toward .5, since responses with error would be random departures from baseline.
-There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
-On the other hand, previous studies of reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
-Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
-
-The third challenge of our analysis is that we focus on predicting wave 3.
-If individuals have a high likelihood of changing between waves 2 and 3, our ability to predict responses at wave 3 will be limited and $\beta$ will be low.
-Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3.
-If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error, and it may not be reasonable to consider this sort of change "persistent."
-
-In addition to these three challenges, there are two forms of change that our model is not well designed to account for.
-The first is a uni-directional shock to the population.
-Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates.
-The second is change in the variance of responses.
-If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will be absorbed into $\beta$ but not enter into $\phi$.
-
-Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error.
-Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model.
-We can only detect whether there is any evidence of persisting belief changes, and therefore whether there is any evidence that active updating is taking place in the population.
-
-## Analysis Steps
-
-Our analysis proceeds in three steps to answer our three research questions.
-First, we evaluate the overall evidence in support of the active updating model.
-To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
-We calculate the posterior probability that the model with the free parameter fits the data better.
-If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.
-
-Second, for variables that show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is concentrated among younger respondents.
-To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff.
-Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45.
-We test a range of cutoffs to ensure robustness of the overall pattern to specific ages.
-
-Finally we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
-Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be inductive.
-
-## Item Selection
-
-To test our model on as broad a range of items as possible, we sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
-We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
-We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on subject material.
-Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about government spending, and questions about social life.
-
-We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust.
-This includes a six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
-We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are.
-We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman].
-Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy.
-We combine four parallel questions about how frequently individuals socialize to create a "social life" scale.
-We combine four questions about support for suicide under different circumstances.
-We also created a scale of support for police use of violence against criminal suspects by combining five binary questions about the conditions under which individuals support police use of violence.
-
-In total, we test the model on 183 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves.
-Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.
-
-[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options.
-These results are reported in Appendix B.
-
-# Results
-
-```{r, message = FALSE, warn = FALSE}
-load(here("data", "results_df.Rdata"))
-load(here("data", "attitude_vars.Rdata"))
-
-label_vars <- res.df %>%
-  filter(var %in% c("intlblks", "confed", "selfhunt", "socbar")) %>%
-  left_join(attitude_vars)
-
-#create a list of variables that display updating
-aum_vars <- res.df %>% filter(aum_pref == 1)
-aum_vars <- aum_vars$var
-
-```
-
-Our model estimates two parameters of interest for each GSS item: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
-High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
-Our measure of active updating, $\phi$, captures the relative proportion of wave 3 variance predicted by wave 2.
-If responses are generated through a true settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
-As the evidence of active updating increases, $\phi$ will increase toward 1.
-Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3.
-
-## Evidence for Active Updating
-
-To evaluate the evidence in favor of the active updating model, we compare for all 183 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
-If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.
-
-Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 183 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.
+```
 
@@
 ```
-
-On the left side of the figure, we see that the majority of $\phi$ estimates fall between .5 and .6, meaning that wave 2 is only a slightly better predictor than wave 1 for most items.
-This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.
-
-To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates.
-Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2.
-Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3.
-Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3.
-Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change.
-
-Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list.
-In other words, even for items that show strong evidence of some active updating, the overall amount of attitude change in the population is likely small.
-
-The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 75 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
-That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
-This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.
-
-We will discuss in more detail below how different items perform.
-To answer our first question, however, we need only focus on the overall distribution.
-Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1.
-This means that for almost all items, measurement error or non-persistent change tend to be much more common than persistent change.
-We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5.
-This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.
-
-## Age Heterogeneity
-
-Our second research question asks whether there is evidence that younger respondents update their views more than older respondents.
-Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
-
-Of the 108 items that showed any evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
-Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
 
@@
 
-The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents.
-These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die.
-These items tend to be in subject areas where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable over time.
-
-For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address past discrimination, all evidence of active updating disappears for people over 30.
-This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but beliefs quickly harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
-For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
-This is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
-
-Eight items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
-These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, and views on whether aging parents should live with their children.
-Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on while young.
-This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is somewhat unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
-
-The remaining 86 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
-However, this does not mean these show strong evidence of an active updating model.
-These items may simply be susceptible to updating for a small proportion of the population.
-
-## Item heterogeneity
-
-While 40 percent of items show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
-Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
-
-Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items shaded by whether they preferred the $\phi = .5$ constraint or not and label a few items that stand out.
-Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.
-
 ```{r phibetascatter, message=FALSE, fig.width=6.8, fig.height = 6.0, fig.cap="Distribution of beta and phi estimates for GSS items, by whether model prefers phi = .5 constraint."}
@@
 ```
-
-Items showing evidence of active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
-A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
-
-Which items show the strongest evidence for active updating?
-There is no way we can discuss all 183 items in detail without the discussion becoming tedious.
-Although we include $\phi$ estimates for all items in Appendix A, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
-We constrain items that showed no evidence of active updating to $\phi = .5$.
-In addition to showing the median and interquartile range of each distribution, the figure also highlights the item in each group that shows the greatest degree of evidence for active updating.
 
@@
 ```
-
-There is a lot to process, even in this summary figure.
-The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low.
-Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity.
-In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]
-
-[^error]: While we cannot control for measurement error in our analysis, we can take steps to mitigate its impact.
-Appendix B presents results comparing items with more than three scale points to coarsened versions of these question with either two or three scale points.
-As noted previously, low measurement error might be a reasonable assumption for some items.
-Previous studies using different approaches to measuring the reliability of survey reports suggest that some items captured in our study, such as whether a person owns a gun, are measured with a high degree of reliability ($> 0.9$) [@houthastings].
-For other items, such as confidence in the leadership of major companies ($\phi = .58$), reliability might be as low as 0.5.
-There is very little correlation ($\rho = .165$) between our $\phi$ estimates and reliability estimates.
-
-Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that more than half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
-These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
-Views on these issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood.
-This means that, for most of these items, social change must occur through cohort succession rather than through individual change.
-
-There are several exceptions to this general pattern, however, even in categories with otherwise low $\phi$ values.
-Items with the largest values generally have one or more attributes in common.
-We consider these attributes to give some general impressions of the pattern.
-
-Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them.
-If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
-This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, which are more nebulous questions that display less active updating.
-Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
-Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
-
-Other high-$\phi$ items have a _changing referent_.
-That is, although the item wording is the same, the object to which the question refers may change between survey waves.
-The most obvious example of this is the item about confidence in the executive branch of the federal government (which has the highest $\phi$ value of all items in the analysis).
-The president changed between the 2008 and 2010 waves of the GSS, meaning that the question no longer referred to the same administration.
-If we generate estimates for this item for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$).
-In the 2006-10 panel, waves 1 and 2 have almost no predictive power ($\beta = .10$).
-For the 2010-14 panel, which takes place entirely during the Obama administration, $\phi$ moves much closer to .5 and predictive power increases ($\phi = .57$, $\beta = .65$).
-
-Likewise, all public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
-The change of administration and changing federal spending policies likely affected these items.
-The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession).
-If the environment is changing we should see exactly this sort of pattern.
-
-Perhaps the most striking pattern in our findings pertains to questions about _gay rights_.
-There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating.
-Questions about civil liberties for gays and about gay marriage are the highest in their categories.
-The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
-This pattern is consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
-
-A few items lack external mechanisms, represent forms of private culture, and did not achieve salience during this period, but still display active updating, including some views on abortion and the morality of different forms of sex.
-We address these items in greater detail in the discussion section.
-
-## Political Beliefs
-
-Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, and because many of these beliefs display evidence of active updating, we examine these beliefs in greater detail than other items.
-Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for questions about political identity, the role of government, and specific policies.
-We break these items up into categories and remove a few items for ease of viewing.
 
@@
 
-There are several notable features of the figure.
-First, political party affiliation is a clear outlier, with greater consistency and greater active updating than other items.
-In contrast, questions asking about specific public spending priorities tend to have low consistency compared to other items.
-Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government ("Gov't do more or less", "Gov't reduce inequality") – display higher consistency than specific policy questions but weak evidence of active updating.
-It is important to restate that for political views, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
-
-While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently.
-Views on abortion, especially when considered as a scale, are consistently reported.
-As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change.
-As discussed previously, views on gay marriage display high levels of active updating.
-In contrast, the item asking about support for legalizing recreational marijuana use is reported with consistency but weak active updating.
-This is notable, as public opinion and policies regarding both have shifted considerably in recent years.
-Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell].
-Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.
-
-# Discussion and Conclusions
-
-This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making.
-These two models of individual change are implicit in analyses of belief change at the society level, but they are rarely compared empirically.
-We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models.
-We now revisit our research questions to summarize what we have learned.
-
-First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
-Around 40 percent of all items show no evidence for updating.
-For the items that do show evidence of active updating, the overall rate of persisting change in the population is likely low.
-For most items measured by the GSS, less than one percent of the population appears to make any persisting change in their views in a two-year period, and most changes appear to be short term or random deviations.
-What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
-Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is.
-But the average level of updating for most views appears to be low.
-
-Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ For a limited subset of items, there seems to be evidence that younger respondents are updating their views whereas older respondents are not.
-This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014).
-Because the youngest respondents in the GSS are 18 years old, we may lack the ability to detect updating that occurs even earlier, so this is probably an underestimate.
-This may be the case for the roughly 70 items that show no evidence of persistent change over time.
-Overall, however, there is significant heterogeneity in the relationship between age and attitude and behavior change.
-
-Third, _are there systematic differences in item content between questions that are better described by each model?_ With 183 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
-We find patterns of responses that provide a range of support for diverse theories of attitude formation, which we discuss below.
-At the same time, our results are consistent with some general expectations about updating.
-Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
-Items assumed to be "peripheral" beliefs, such as specific policy questions, also showed evidence for updating and low consistency.
-Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18.
-
-There are a handful of items that demonstrate active updating and enough consistency to merit further consideration, but are not easily explained by theoretical mechanisms outlined so far.
-These include questions about abortion in the case of rape or poverty, the morality of premarital and teen sex, the ideal number of children in a family, and whether children should obey adults or think for themselves.
-These items did not achieve high salience during the time of the study and there were not, to the best of our knowledge, clear changes in elite opinion on these issues.
-This suggests additional mechanisms that can drive persisting change at the individual level, such as exposure to these ideas through mass media or by directly interacting with people in these groups.
-In the case of questions about children and family structures, it might be the experience of having children that leads to changes in these attitudes.
-It is important to note that even though these items demonstrate active updating, it is a small proportion of the population that are changing their views during the study window.
-
-The overall pattern of results, while diverse and at times hard to reconcile, does shed light on several theoretical debates. We now consider these implications.
-
-## Implications for Cultural Sociology
-
-In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession.
-Most beliefs about gender roles, sexual morality, and abortion appear to be settled by early adulthood.
-The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time.
-In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating.
-Even those items that did display strong evidence for active updating (e.g.
-"whites rich or poor" and "whites work hard") were still hard to predict wave after wave.
-
-Our results suggest that one reason attitudes are largely stable is because most issues simply do not reach the level of salience necessary to shift opinions.
-In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
-By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period.
-The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
-This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.
-
-While the dominant pattern is stability during adulthood, the results suggest greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories.
-The most prominent and well supported theories of attitude change suggest a peak of susceptibility to attitude change early in adulthood and either a rapid or gradual decline in attitude change with age [@alwinkrosnick; @visserkrosnick].
-An additional view, the "life stages" hypothesis, suggests a similar pattern with a surge in attitude change late in life [@visserkrosnick].
-While this age-related decline appears to be true for many political attitudes, the pattern is not nearly as consistent for other kinds of attitudes.
-Some items, such as views on the Bible, suggest equal openness to attitude change as individuals age.
-Others, such as views on most abortion questions, suggest that early adult socialization is so strong that lifetime opinions are settled by the time most people reach even 18 years old.
-
-For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
-This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level.
-Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors that encourage openness to change at different ages.
-Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes.
-Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick].
-Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].
-
-## Implications for Political Sociology
-
-A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
-Respondents might express different positions from wave to wave, but in guessing what a person will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
-While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals in our data changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
-With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still limited.
-Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.
-
-These results are consistent with theories that posit that, at least in the current era, ideological identification (view of oneself as liberal or conservative), rather than some other belief such as partisan identification, moral views, religion, or particular policy positions, is the "central" political belief [@boutylinevaisey; @converse].
-
-It is important not to overstate the role of this belief in forming other beliefs.
-While our findings suggest that individuals appear to bring their partisan identification in line with their ideological identification, there is limited evidence of adjusting other beliefs.
-This pattern is most consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
-
-Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time.
-This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.
-Similarly, our findings are not consistent with the idea that individuals adopt a partisan affiliation based on their social groupings and subsequently adjust their ideological commitments to conform to that [@greenetal].
-For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much updating as partisan identification, which is not true in our data.
-
-The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile items such as partisan identity, abortion, and gay marriage, individuals appear to lack clear opinions on most specific policy questions.
-The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
-What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."
-
-The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents, especially those over 30 years old, hold a general political identity and a few clear views on issues like abortion, attempt to align their partisan identification to their views, and respond to elite opinion change when it provides clear signals.
-It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates.
-
-## Methodological Implications
-
-Methodologically, our results highlight the challenges of evaluating population-wide attitude change using short-term panel studies.
-The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change.
-It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that substantive change is too rare in a sample of adults to measure accurately on the vast majority of items.
-This strongly argues against using two-wave panels to measure attitude change, which do not allow researchers to separate persisting from non-persisting changes.
-
-The fact that persistent change is practically nonexistent for many items bolsters the case for using repeated survey responses to measure the reliability of survey items [@alwin; @houthastings], since it is often a valid assumption that the underlying view is unchanging.
-At the same time, our results call for greater focus on methodological tools that can separate short-term attitude change from measurement error.
-While we generally assume that lasting changes in attitudes are more likely to influence behavior, this is not necessarily true.
-Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying it is difficult.
-
-We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation.
-One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.
-
-We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
-We see too much inconsistency in responses for commitment bias to be a major explanatory factor.
-Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
-499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p.
-500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p.
-501], tend to show low active updating and low consistency.
-
-We cannot (and would not want to) rule out the possibility that panel conditioning is taking place in the GSS.
-We believe it is worthwhile to explore these same GSS panels for evidence of panel conditioning.
-However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
-
-Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.
-Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful.
-In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
-Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly problematic.
-Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
-
-Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults.
-Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production"—childhood and adolescence—that give rise to persistent beliefs in adulthood.
-
-\pagebreak
-
-# Appendix A: $\phi$ values for all variables
-
-Figures \ref{fig:phi_relig} through \ref{fig:phi_racgen} plot $\phi$ estimates for all items included in the analysis, grouped by subject material.
-
-```{r loadfiguredata}
-load(here("data", "figure_data.Rdata"))
-```
-
 ```{r religphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about religious activity and beliefs, social life, subjective SES, and suicide."}
@@
 ```
-
-# Appendix B: Coarsened variable estimates
-
-```{r loadcoarse}
-load(here("data", "coarse_vars.Rdata"))
-
-```
-
-$\phi$ estimates for coarsened items are presented in Figures \ref{fig:coarse_relig} through \ref{fig:coarse_racgen}.
-If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree).
-If scales included a clear midpoint, we included that and coarsened responses on either side.
-For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median.
-If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.
 
@@
 ```
-
-Generally speaking, most items with no evidence of active updating continue to have no evidence of active updating when coarsened.
-For items that showed evidence for active updating, the coarsened estimates tended to be very similar to the uncoarsened estimates, suggesting that persisting and non-persisting changes happen about as often between ends of these scales as they do within ends of these scales.
-
-There are a couple notable departures from this general pattern.
-Several political questions -- general political views, views on whether the government should reduce inequality, and views on whether the government should help blacks -- show decreased evidence of persisting change when coarsened.
-This suggests that changes around the midpoint of the scale tend to be measurement error, a finding that is consistent with previous work suggesting that individuals without settled political views tend to choose points around the middle of the response scale [@converse].
-
-\pagebreak
-
-# References

### 2020-08-27 manuscript9.Rmd
added ASR edits

@@
 
-```{r dag, message=FALSE, fig.width=6.8, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
+```{r dag, message=FALSE, fig.width=5, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
 aum <- 
@@
 coordinates( aum ) <-
-  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3),
+  list( x=c(y1 = 1, y2 = 3, y3 = 5, v1 = 1, v2 = 3, v3 = 5),
         y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )
 
-aum_dag <- aum %>% tidy_dagitty() %>% ggdag() + 
+aum_dag <- aum %>% tidy_dagitty() %>% ggdag(node_size = 8) + 
   theme_dag_blank() + 
@@
   scale_y_continuous(limits = c(.8,3.2)) +
-  theme(plot.title = element_text(hjust = 0.5)) +
-  theme(plot.margin = unit(c(1,1,.2,.2), "cm"))
+  scale_x_continuous(limits = c(.5, 5.5)) + 
+  theme(plot.title = element_text(hjust = 0.5))
 
@@
 coordinates( sdm ) <-
-  list( x=c(y1 = 1, y2 = 2, y3 = 3, v1 = 1, v2 = 2, v3 = 3, U = 2),
+  list( x=c(y1 = 1, y2 = 3, y3 = 5, v1 = 1, v2 = 3, v3 = 5, U = 3),
         y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1, U = 3) )
 
-sdm_dag <- sdm %>% tidy_dagitty() %>% ggdag() + 
+sdm_dag <- sdm %>% tidy_dagitty() %>% ggdag(node_size = 8) + 
   theme_dag_blank() + 
@@
   scale_y_continuous(limits = c(.8,3.2)) +
-  theme(plot.title = element_text(hjust = 0.5)) +
-  theme(plot.margin = unit(c(1,.2,.2,1), "cm"))
-
-gridExtra::grid.arrange(aum_dag, sdm_dag, ncol = 2)
+    scale_x_continuous(limits = c(.5, 5.5)) + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+gridExtra::grid.arrange(aum_dag, sdm_dag, nrow = 1)
 ```
@@
   filter(grepl("_c", var)==FALSE) %>%
-  mutate(aum = ifelse(var %in% aum_vars, "\nPersistent\nChange\n", "\nNo Persistent\nChange\n")) %>%
+  mutate(aum = ifelse(var %in% aum_vars, "\nPersistent\nChange\n", "\nNo\nPersistent\nChange\n")) %>%
   ggplot(aes(x = p, y = b1)) + 
-  geom_point(aes(fill = aum, shape = aum)) + 
+  geom_point(aes(fill = aum, shape = aum), alpha = .7) + 
   scale_shape_manual(values = c(21, 22)) + 
@@
   left_join(cat_counts) %>%
+    mutate(small.cat = ifelse(small.cat == "Sex, sexuality & abortion",
+                            "Sex, sexuality &\nabortion", small.cat),
+           small.cat = ifelse(small.cat == "Guns, laws, crime & police",
+                            "Guns, laws, crime &\npolice", small.cat),
+          small.cat = ifelse(small.cat == "Confidence in leadership",
+                            "Confidence in\nleadership", small.cat),
+          small.cat = ifelse(small.cat == "Relig. identity & beliefs",
+                            "Relig. identity &\nbeliefs", small.cat),
+          var.name = ifelse(var.name == "Executive branch", "Exe. branch", 
+                                   var.name)) %>%
   mutate(small.cat = paste(small.cat, " (", n, ")", sep = "")) %>%
@@
     legend.title = element_blank(),
-    axis.text=element_text(size=9),
+    axis.text=element_text(size=7),
     legend.text=element_text(size=9)
@@
   geom_point(aes(fill = small.cat), shape = 21) + 
-  geom_text_repel(aes(label = var.name), size = 3) + 
+  geom_text_repel(aes(label = var.name), size = 1.75) + 
   theme_bw() + 
@@
     strip.text.x = element_blank(),
+    strip.text.y =element_text(size=8),
     strip.placement = "outside",
@@
     strip.text.x = element_blank(),
+        strip.text.y =element_text(size=8),
     strip.placement = "outside",
@@
     strip.text.x = element_blank(),
+            strip.text.y =element_text(size=8),
     strip.placement = "outside",
@@
     strip.text.x = element_blank(),
+                strip.text.y =element_text(size=8),
     strip.placement = "outside",

### 2021-12-15 corrected_figures.Rmd
results corrected

--- 2020-08-27 manuscript9.Rmd
@@
 ---
-title: "Stability and Change in Personal Culture"
-author:
-- Kevin Kiley
-- Stephen Vaisey
-date: "Duke University"
-output:
-  bookdown::word_document2: 
-    fig_caption: yes
-    reference_docx: word-styles-reference-02.docx
-  pdf_document:
-    number_sections: yes
-header-includes: \usepackage{longtable}
-nocite: |
-  @swidler01
-bibliography: panel_change_bib.bib
-abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change.
-One approach theorizes that people actively update their beliefs and behaviors in the face of new information.
-The other argues that, following early socialization experiences, dispositions are stable.
-We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data.
-We test this model on 183 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey.
-Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model.
-Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes.
-When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes.
-We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
+title: "Corrected Figures"
+author: "Kevin Kiley"
+date: "12/15/2021"
+output: pdf_document
 ---
@@
 ```{r setup, include=FALSE}
-# Before the first commit, I added "here", changed the load paths with here(), and added the %!in% function
-
 knitr::opts_chunk$set(echo = FALSE)
@@
 '%!in%' <- function(x,y)!('%in%'(x,y))
-
 ```
 
-# Introduction
-
-Culture is an important part of social life but cultures are continuously evolving.
-In 1972, for example, more than 40 percent of US adults supported a law outlawing interracial marriage.
-Three decades later, this opinion had become so uncommon that the question was removed from the US General Social Survey.
-How does this kind of cultural change happen?
-
-Attempts to account for opinion changes in society have produced conflicting theories about the process of opinion formation and the ability of individuals to maintain consistent attitudes.
-Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues.
-As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden].
-By contrast, cohort replacement theories posit that people do hold opinions and are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder].
-Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p.
-xi; @achen1992].
-More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].
-
-Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology.
-The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation.
-This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001).
-The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life.
-This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).
-These two models represent different approaches to understanding how people come to hold diverse forms of _personal culture_, or the declarative and non-declarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo17].
-
-In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data.
-In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors.
-We then apply this method to 183 items from the 2006-2014 General Social Surveys (GSS).
-By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply.
-Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names).
-We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S.
-society.
-
-Our analysis yields several results.
-First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change.
-Simply put, there is little evidence that large numbers of U.S.
-adults changed their beliefs or practices in lasting ways over this period of study.
-Second, settled opinions vary in how consistently individuals report the same answer.
-Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent or random responses.
-At the same time, people are consistent on both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest.
-Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents.
-On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life.
-Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to persist than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes and behaviors consistently.
-This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.
-
-Our findings offer broad support to theories claiming that cultural change comes through generational turnover rather than persuasion and social influence.
-However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge the idea that change in all attitudes follows a similar trajectory over the life course.
-The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification than to their general political beliefs.
-We argue that there is a place for both the active updating and settled dispositions models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
-
-# Theoretical Background
-
-## Belief Formation in Cultural Sociology
-
-How does cultural change happen at the individual level?
-Let us make the question concrete by imagining a person who answers the same question each year for several years.
-The question could be anything, but assume it is this GSS question: "Please tell me whether you strongly agree, agree, disagree, or strongly disagree [with this statement]: 'a working mother can establish just as warm and secure a relationship with her children as a mother who does not work.'"
-
-How does the respondent formulate a response to that question each time, year after year?
-To make things as explicit as possible, we can write the data-generating process formally.
-Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions.
-We will explain this in greater detail below.
-For now, consider the following two simple models:
-
-\begin{equation}
-  y_{it} = y_{it-1} + \nu_{it}
-  (\#eq:AR)
-\end{equation}
-
-\begin{equation}
-  y_{it} = U_i + \nu_{it}
-  (\#eq:FE)
-\end{equation}
-
-These models may seem similar at first glance, but they have different implications for the pattern of individuals' responses we would observe over time.
-Equation \@ref(eq:AR) represents an _active updating_ model and Equation \@ref(eq:FE) represents a _settled dispositions_ model, as we explain below.
-Figure \@ref(fig:dag) shows these models in graphical form, which helps highlight their differences.
-In the next two sections, we consider each of these models and briefly discuss their connections to sociological theories.
-
-[Figure \@ref(fig:dag) about here]
-
-```{r}
-library(ggdag)
-library(dagitty)
-```
-
-## Active Updating Model
-
-Equation \@ref(eq:AR) represents the active updating model.
-The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$).
-There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time.
-Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state.
-This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].
-In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.
-
-More informally, this AUM posits a person who is updating her views in the face of social experience.
-There are formal Bayesian ways of modeling updating, and this model underlies theories of rational updating in the face of new information [@achen1992; @bartelsjackman; @pageshapiro], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply.
-Following Gross's [-@gross09, pp.
-367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."
-
-Active updating lies at the heart of most theories that suggest social environments influence individual attitudes and behaviors.
-When sociologists posit that adults adopt the beliefs and behaviors of their friends [@christakisfowler; @centola; @dellapostashimacy], or that individuals adopt behaviors that they view as consistent with their cultural dispositions [@goldbergstein], they invoke this model.
-When social scientists argue that people adapt their partisan affiliations to conform to their social groupings, change their ideological commitments to conform to their partisan identifications [@greenetal], change their partisan affiliations to match their ideological commitments [@baldassarigelman; @abramowitzsaunders06], or respond to exposure to diverse opinions by becoming more extreme in their viewpoints [@bailetal], the claim is rarely that these changes are temporary shocks that will revert to a stable baseline over time.
-Instead, the claim is that the changing social environment changes dispositions, which will then remain stable until the environment changes again.
-
-Regardless of whether the exact process is rational, heuristic, discursive, embodied, or any combination of these, the key notion is that the actor continues to be influenced by the (social) environment in ways that might lead to long-term, persistent change in beliefs, practices, and identities [@gross09; @degloma].
-As the person encounters new considerations throughout her life (e.g., adding working mothers to one's social network, seeing changing media representations of working mothers), she can continue to revise her views.
-
-The AUM makes no assumption about the distribution of $\nu_{it}$.
-Specifically, it does not assume that $\nu_{it}$ has an expected value of 0, either for any time $t$ or for any person $i$.
-This leaves open the possibility of a population-wide shift in responses as many people react to the same changes in the environment.
-
-Several theories of cultural change at the societal level require evidence of active updating at the individual level, such as arguments that the individuals in the population are becoming more polarized on political issues [@dimaggio_polarization; @iyengarwestwood], arguments that individuals adjust their political preferences based on the performance of the governing party [@pageshapiro; @bartelsjackman], arguments that new ideas diffuse across the population through person-to-person contact [@centola], and arguments that changes in elite discourse drive change in mass opinion [@zaller].
-In the absence of evidence that people change their minds in persisting ways, these theories cannot be valid at the aggregate level, and we must seek other explanations for aggregate changes.
-
-## Settled Dispositions Model
-
-Equation \@ref(eq:FE) represents the settled dispositions model.
-Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$).
-Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$.
-As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses.
-Such considerations are thus temporary influences only.
-At its core, this model reflects the Bourdieusian model of action that emphasizes "the past conditions of production" [@bourdieulogic, pp.
-66ff].
-In other words, updating happened, but it happened in the past, prior to the time of the study.
-In this sense, $U_i$ reflects the "habitus."
-
-The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise].
-The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level.
-These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder].
-Vaisey and Lizardo [-@vaiseylizardo], looking at a range of opinion items in the GSS, suggest that population-wide cultural change most likely comes in the form of cohort replacement.
-
-On its face, the settled dispositions model might seem to imply that individuals hold strong, consistent opinions, but this need not be the case.
-The settled dispositions model is also consistent with theories that suggest that people lack stable opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion].
-These theories tend to focus on questions of public policy, since the act of opinion creation requires more information than individuals typically have [@zaller; @converse].
-Such accounts imply that peoples’ responses at any particular time are a deviation from a baseline, with this deviation often assumed to be random and commonly referred to as a form of “measurement error” [@pageshapiro; @converse].
-As long as these deviations are random, there is no reason to perceive these changes as a form of updating, even if wave-to-wave differences are quite large.
-
-Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time.
-That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time.
-The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within people these shifts will be erased over time as people return to their baselines.
-
-Like with the active updating model, there are theories of attitude development and change at societal level that would require evidence of the settled dispositions model at the individual level.
-For example, if people's attitudes change in a way that reflects updating, then it would be wrong to consider their deviations from their average to be "measurement error," and theories that suggest that changes in belief are not real [@pageshapiro] would be unsupported.
-Similarly, an item would have to be relatively stable for most adults for aggregate cultural change to come primarily through cohort replacement [@ryder].
-
-The two general models thus differ fundamentally on their emphasis on the character of personal change.
-In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time.
-In the SDM, changes tend to revert because each person returns to their baseline.
-We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_.
-
-## Consistency in Response
-
-Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave.
-This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change.
-Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development.
-However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.
-As such, we designate a second term, _consistency_, to reflect this second dimension of change.
-_Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.
-
-## Toward Theoretical Synthesis
-
-We do not suggest that all attitudes and behaviors captured by survey questions follow a settled dispositions model or that all display active updating.
-Instead, we suggest that by classifying the pattern of change at the individual level, we can use that as a tool for adjudicating debates of change at the aggregate level.
-
-The clearest theoretical contrast is between models that assume beliefs are settled during adulthood, such as cohort replacement theories, which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives, which would produce an active updating pattern.
-This disagreement thus forms the primary structure of our analysis.
-However, it is possible that some items display active updating while others do not.
-There are at least three ways to understand how both models could be present in the population without contradiction: age-based differences, core-periphery models, and public-private differences.
-
-_Age-based differences:_ The concept of a "cohort effect" is that the environment influences the baseline response tendency when a person is young and then stabilizes for the remainder of one's life [@elder; @bartelsjackman].
-To put it differently, cohort formation requires an early period of updating followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.
-
-Items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults but high active updating for young people, as this is the period most commonly found to be susceptible to attitude change [@glenn; @inglehartbaker; @visserkrosnick; @danigelis].
-
-_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs, which they use to adjudicate peripheral attitudes [@boutylinevaisey; @goldbergstein].
-This work has predominantly focused on political beliefs and adjudicating whether people use ideological identity, policy positions, moral beliefs, or partisanship to form their positions on other issues [@lakoff; @greenetal; @kinderkalmoe].
-In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification -- assessment of oneself as a conservative or a liberal, not as a Republican or Democrat -- as the "core" belief in the political space.
-Similarly, Baldassari and Gelman [-@baldassarigelman] find increased alignment between partisan identification and policy positions but little increase in the correlation between specific policy positions over time.
-They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp.
-442].
-
-The picture that emerges from these previous works is that ideological identity should be a core, if poorly reported, disposition.
-Changes in ideological identity should be non-persisting, occurring because people either do not fully understand these terms and report their views with error [@converse] or because people bounce around a stable mean, meaning this item should show low updating and medium to low consistency.
-On the other hand, we should see evidence that individuals adjust their partisan identification to match their ideological identification as they come to find that one party better represents their core positions [@baldassarigelman].
-In other words, liberal Republicans should become liberal Democrats, not conservative Republicans.
-This means that partisan identification should display high levels of updating.
-Again, this does not mean that many people change, only that changes tend to persist.
-
-Alongside ideological identification, moral issues and general political sentiments might serve as core organizing principles, but what these beliefs are will not be consistent across people [@baldassarigelman].
-For example, some people might hold strong, consistent views on abortion while others care about the amount of government intervention in the economy.
-Because of this heterogeneity, these beliefs should have a moderate level of consistency, since those people who do not value them should report them inconsistently.
-There should be no evidence of updating in these beliefs, as previous work suggests that people do not bring these attitudes in line with other beliefs, even if they are discordant.
-
-Peripheral beliefs that "grow" out of core beliefs, such as preferences for government spending on various priorities and specific policy positions, should show some evidence of updating, as individuals attempt to align these with their core beliefs.
-However, because of the cognitive effort required to map peripheral beliefs onto core beliefs, we generally expect the former to be inconsistently reported [@zaller; @converse].
-However, there are conditions that might decrease that cognitive effort and therefore increase the amount of updating and consistency with which peripheral issues are reported, to which we now turn.
-
-_Public vs.
-Private Culture:_ A final dimension that likely influences the degree of active updating and consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions.
-The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: p.
-91].
-This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances.
-However, when public culture provides a strong "scaffold" through clear signals of how identities and social locations should influence opinions, it becomes easier for people to maintain consistency [@lizardostrand].
-
-Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse].
-This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these, since signals are consistent over time.
-However, we do highlight one issue that saw significant change in elite opinion, which should have lead to active updating in the general population.
-In the time frame we observe (2006-2014), there was a major shift in elite opinion on gay rights.
-Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage.
-For this issue, we expect to find evidence of active updating throughout the population, not just among young people.
-
-Items that tap some public dimension of behavior and attitudes are also likely to show higher levels of active updating because social mechanisms, such as publicly joining a group, can help maintain changes in ways that intrapsychic forces cannot.
-Any item where the burden of maintaining consistency is externalized and not only “in a person’s head” might demonstrate both higher rates of active updating and greater consistency.
-
-As we noted previously, we do not expect that all items will align with just one model.
-If we find a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions.
-Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all “matter.” Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work.
-We believe achieving a better understanding of these processes will be relevant for many sub-fields of sociology as well as for other social science fields that study changes in beliefs and behaviors.
-
-# Research Questions and Expectations
-
-With these considerations in mind, we ask the following questions:
-
-First, _to what extent are patterns of personal cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo].
-This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.
-
-Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ In a sample of adults, the possibility exists that cohort formation may be complete for most beliefs and behaviors before the study period (i.e., before age 18).
-However, if some cohort formation is still occurring among younger respondents, we should see that evidence consistent with the AUM is disproportionately located among younger respondents.
-
-Third, _are there systematic differences in item content between questions that exhibit different levels of active updating and consistency?_ The preceding sections have made some predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs and behaviors.
-As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.
-
-# Analytic Strategy
-
-To investigate these ideas empirically, we examine 183 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model.
-This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012).
-Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model.
-We discuss item selection below.
-While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.
-
-## Statistical Model
-
-### Basic Models
-
-Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time.
-We first consider measuring the amount of active updating.
-Our two different models make different predictions in the three-wave panel context.
-The AUM makes the following prediction for wave 3:
-
-\begin{equation}
-  E(y_{i3}) = y_{i2}
-  (\#eq:AUM2)
-\end{equation}
-
-That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2.
-If change is persisting, then our best guess is that a person's response will be close to what they said last time, and previous responses will provide no additional predictive power.
-
-The SDM makes the following prediction:
-
-\begin{equation}
-  E(y_{i3}) = U_i
-  (\#eq:SDM2)
-\end{equation}
-
-Since the best estimate of $U_i$ is the mean of the respondent's two previous answers, we can rewrite the SDM prediction as:
-
-\begin{equation}
-  E(y_{i3}) = \frac{y_{i2} + y_{i1}}{2}
-  (\#eq:SDM2B)
-\end{equation}
-
-That is, the SDM predicts that the average response of previous waves is the best predictor of the next response.
-If change is non-persisting, then taking the average of the last two responses will be our best guess about a person's underlying position.
-
-### Combined Model
-
-Both of the models above include $y_{i2}$ as a predictor of $y_{i3}$, but only the settled dispositions model includes $y_{i1}$ as a predictor.
-If the SDM is correct, $y_{i1}$ should be just as predictive as $y_{i2}$ because both are (on average) equally informative about the respondent's stable disposition.
-Therefore, to test for evidence of active updating we use a model that evaluates whether $y_{i2}$ carries any additional predictive power over $y_{i1}$.
-If the two previous estimates are equally predictive, then we can be relatively confident that the data we observe came from a settled dispositions model.
-However, if $y_{i2}$ is a better predictor of $y_{i3}$ than $y_{i1}$ is, this provides evidence that some respondents engage in active updating.
-We use the following non-linear model to estimate the relative influence of $y_{i1}$ and $y_{i2}$:
-
-\begin{equation}
-  E(y_{i3}) = \alpha + \phi\beta y_{i2} + (1-\phi)\beta y_{i1}
-  (\#eq:MODEL)
-\end{equation}
-
-Rather than generate separate coefficient estimates for $y_{i2}$ and $y_{i1}$, this model generates two parameter estimates of interest: $\beta$, which captures how well any combination of previous waves predicts a person's response at wave 3, and $\phi$, the relative proportion of wave 3 explained by wave 2 compared to wave 1.
-If the Settled Dispositions Model is the preferred data-generating process for an item, then both $y_{i2}$ and $y_{i1}$ should be equally predictive of $y_{i3}$, and $\phi$ will equal .5, meaning the best estimate of wave 3 is a function of the mean of previous waves, consistent with Equation \@ref(eq:SDM2B).
-If the active updating model is present in at least some respondents and wave 1 provides no additional predictive power when we control for wave 2, then $\phi$ will increase toward 1 to converge with Equation \@ref(eq:AUM2) in certain circumstances.
-
-Our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating.
-We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model.
-If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$.
-If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.
-
-### Comparison to Other Approaches
-
-Our model is not the first to measure stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test.
-Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS responses.
-This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error.
-These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier.
-While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount change is consistent across waves.
-This approach also combines persisting and non-persisting change into two similar but distinct kinds of change: "structural" and "non-structural," both of which can be persisting and non-persisting.
-This distinction, while important for some theoretical questions, is not our focus.
-
-A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief.
-This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_.
-Since we at times invoke both these models, we include in our analysis several composite scales of related items.
-If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in their levels of active updating.
-
-Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal].
-While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns.
-One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster.
-If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 would become a better predictor of wave 3.
-
-We do not view this as a problem for our theoretical models.
-If people change their attitudes or behavior as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant.
-This might lead to a higher estimate of active updating than we might observe in a population that did not take the survey, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update beliefs over time.
-
-A second form of panel conditioning posits that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time, even if they actually change.
-In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change.
-This would be problematic for our study, as it would under-estimate the amount of real change in the population.
-However, if this is the case, we should observe no active updating in responses and high levels of consistency, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.
-
-Finally, a number of approaches exist for evaluating theoretical process of belief formation and change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization].
-These tools are well designed to address the questions they set out to answer.
-However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.
-
-### Limitations of the Method
-
-Three challenges limit our ability to evaluate the presence of settled dispositions and active updating models using our approach and therefore limit the conclusions we can reach.
-The first is that our model is designed to allocate variance explained to each of the prior waves rather than to assign probabilities to each data-generating process.
-Because of this, a few individuals making large persisting changes can inflate the $\phi$ estimate even if most individuals make small non-persisting changes.
-
-The second is measurement error, which is a form of non-persisting change.
-For some researchers, measurement error represents the inconsistency that results from constructing responses anew each wave [@zaller; @converse], and in that case it should not be considered "error" so much as an indicator of that process at work, since there is no "true" item to measure.
-For other theories, measurement error reflects individuals' inability to accurately report their response.
-It is also possible that measurement error reflects errors of selection and interpretation, such as misunderstanding the question or incorrect coding.
-
-Because measurement error looks the same as non-persisting real change (and because the latter is sometimes interpreted as the former), $\phi$ estimates will be biased toward .5, since responses with error would be random departures from baseline.
-There is evidence that many of the items explored in our analysis are measured with significant error [@alwin; @houthastings].
-On the other hand, previous studies of reliability tend to conflate measurement error and non-persisting real change in attitudes, meaning that while we might have good estimates for the combination of these two processes, we cannot separate them.
-Because of measurement error, it is unlikely that $\phi$ and $\beta$ will reach 1 for any item, even if the underlying process is fully based on active updating.
-
-The third challenge of our analysis is that we focus on predicting wave 3.
-If individuals have a high likelihood of changing between waves 2 and 3, our ability to predict responses at wave 3 will be limited and $\beta$ will be low.
-Our model relies on the assumption that "persisting" change is relatively rare and that individuals who change between waves 1 and 2 do not also make persisting changes in the opposite direction between waves 2 and 3.
-If the rate of active updating is so high that individuals make changes between each wave, then the model becomes indistinguishable from the settled dispositions model with high measurement error, and it may not be reasonable to consider this sort of change "persistent."
-
-In addition to these three challenges, there are two forms of change that our model is not well designed to account for.
-The first is a uni-directional shock to the population.
-Since our model includes an intercept, changes that shift all responses toward one end of the scale are absorbed into that term and not accounted for in our $\phi$ estimates.
-The second is change in the variance of responses.
-If all individuals shift outward or inward toward the population mean but maintain their relative position in the overall distribution, this change will be absorbed into $\beta$ but not enter into $\phi$.
-
-Despite these limitations, the model is capable of detecting the presence of persisting change even in the presence of high levels of measurement error.
-Because of this, it is best to think of our approach as seeking any evidence in favor of active updating, rather than allocating probabilities to each model.
-We can only detect whether there is any evidence of persisting belief changes, and therefore whether there is any evidence that active updating is taking place in the population.
-
-## Analysis Steps
-
-Our analysis proceeds in three steps to answer our three research questions.
-First, we evaluate the overall evidence in support of the active updating model.
-To do this, we compare for each item the Bayesian Information Criteria (BIC) of a model estimated using Equation \@ref(eq:MODEL) with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
-We calculate the posterior probability that the model with the free parameter fits the data better.
-If the model with the constraint is preferred, then we conclude that both wave 1 and wave 2 are equally good predictors of wave 3, meaning there is no evidence that respondents are actively updating on that item.[^link]
-
-[^link]: All data used in this analysis is publicly available through the General Social Survey at the National Opinion Research Center at http://gss.norc.org/.
-R code for how we cleaned the GSS panels, all analyses performed in this paper, and all figures created in this paper can be found at http://github.com/krkiley/panel_change.
-
-Second, for variables that show at least some evidence of active updating ($\phi > 0.5$), we ask whether the persistent change is concentrated among younger respondents.
-To test this, we re-estimate our original model and allow $\phi$ to have different values above and below a given age cutoff.
-Rather than test a single age cutoff, we again use BIC comparisons to evaluate whether including the dummy variable improves the model fit using a cutoff of every age between 20 and 45.
-We test a range of cutoffs to ensure robustness of the overall pattern to specific ages.
-
-Finally we consider whether there are any meaningful patterns in the relative distribution of evidence for active updating across variables as suggested by existing theories.
-Although, as we discussed above, previous work gives some indications about what we might expect, the approach here will necessarily be inductive.
-
-## Item Selection
-
-To test our model on as broad a range of items as possible, we sought measures of attitudes, beliefs, self-assessments, self-perceptions, and social behaviors that were asked in three waves of the GSS panels.
-We excluded from our analysis questions that focused on demographic characteristics (marital status, household size, region, gender, race, ethnicity), work activity (employment status, income, hours worked, size of workplace), objective socioeconomic status (years of education and highest degree, home ownership), and evaluations of a respondent by the interviewer.
-We follow Hout and Hastings [-@houthastings] and group questions into 15 categories based on subject material.
-Questions in the same category tend to be asked in the same block during the survey and have the same structure, such as questions about confidence in institutions, questions about government spending, and questions about social life.
-
-We also follow Hout and Hastings [-@houthastings] in recreating common scales about gender roles, access to abortion, and social trust.
-This includes a six-question scale of support for abortion and a seven-question scale which includes the question asking about abortion under any circumstances ("abany").
-We use Smith's [-@smith97] scale of "misanthropy" by combining questions about how helpful, fair, and trustworthy people are.
-We use four questions to create a scale of gender role attitudes [@cotterhermsenvanneman].
-Like Hout and Hastings, we combine civil liberties items into six scales about the freedom of atheists, communists, militarists, racists, and, in the 2010-14 panel, Muslim clergy.
-We combine four parallel questions about how frequently individuals socialize to create a "social life" scale.
-We combine four questions about support for suicide under different circumstances.
-We also created a scale of support for police use of violence against criminal suspects by combining five binary questions about the conditions under which individuals support police use of violence.
-
-In total, we test the model on 183 GSS items, including the composite scales.[^coarsened] For each question, we use all cases for which the respondent gave responses in all three waves.
-Models are estimated using weights that account for the GSS's sampling design as well as non-response adjustment.
-
-[^coarsened]: To ensure that our estimates of $\phi$ are not simply artifacts of response scale construction, we estimate the model on coarsened versions of items, generated by collapsing responses to questions with more than three response options into scales of two or three response options.
-These results are reported in Appendix B.
-
-# Results
-
 ```{r, message = FALSE, warn = FALSE}
-load(here("data", "results_df.Rdata"))
+load(here("correction", "results_df.Rdata"))
 load(here("data", "attitude_vars.Rdata"))
@@
 
-Our model estimates two parameters of interest for each GSS item: $\beta$, our measure of consistency, captures how well any combination of previous waves predicts a person's response at wave 3.
-High values of $\beta$ indicate that individuals are relatively consistent in their responses, once we control for the amount of active updating.
-Our measure of active updating, $\phi$, captures the relative proportion of wave 3 variance predicted by wave 2.
-If responses are generated through a true settled dispositions model, then $\phi$ will be .5 (i.e., both wave 1 and wave 2 are equally good predictors of wave 3).
-As the evidence of active updating increases, $\phi$ will increase toward 1.
-Both $\phi$  and $\beta$ equaling 1 would indicate that all individuals who changed between waves 1 and 2 persisted in their change, that an item was measured with no measurement error, and there was no additional change between waves 2 and 3.
-
-## Evidence for Active Updating
-
-To evaluate the evidence in favor of the active updating model, we compare for all 183 items the Bayesian Information Criteria (BIC) of a model with a free estimate of $\phi$ to a model that constrains $\phi= .5$.
-If the model with the constraint is preferred, then there is no evidence that respondents engage in an active updating process with respect to that item.
-
-Figure \@ref(fig:phihist) plots the distribution of $\phi$ estimates for the 183 items evaluated in this analysis and the posterior probabilities that the model without the constraint fits the data better, generated by comparing the BIC from models with and without the constraint.
-
-[Figure \@ref(fig:phihist) about here.]
-
-On the left side of the figure, we see that the majority of $\phi$ estimates fall between .5 and .6, meaning that wave 2 is only a slightly better predictor than wave 1 for most items.
-This suggests that if active updating is happening in these responses, it is relatively infrequent or small compared to temporary change and measurement error.
-
-To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates.
-Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2.
-Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3.
-Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3.
-Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change.
-
-Considered this way, the majority of GSS items demonstrate persisting change at a rate of less than one percent of the total sample, and none shows evidence of persisting change greater than 5 percent, with confidence in the leadership of the executive branch of the federal government, confidence in banks and the financial system, and the respondent's belief about whether they will be able to find a good job topping the list.
-In other words, even for items that show strong evidence of some active updating, the overall amount of attitude change in the population is likely small.
-
-The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 75 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period.
-That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position.
-This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.
-
-We will discuss in more detail below how different items perform.
-To answer our first question, however, we need only focus on the overall distribution.
-Forty percent of items show no evidence for active updating, and among items that do show some evidence of persistent change, very few come close to approaching 1.
-This means that for almost all items, measurement error or non-persistent change tend to be much more common than persistent change.
-We can only be really confident in detecting substantial amounts of persistent change (greater than 2 percent of the population) among a small minority of items, perhaps 1 in 5.
-This means that most of the "change" that shows up in the GSS panels reflects some combination of measurement error or non-persistent change.
-
-## Age Heterogeneity
-
-Our second research question asks whether there is evidence that younger respondents update their views more than older respondents.
-Although it's impossible to determine what proportion of people are following each data-generating process, it is possible to compare the age distribution of evidence for updating in each item.
-
-Of the 108 items that showed any evidence for active updating in the last section, 22 showed differential effects of age for more than 50 percent of cutoff ages we tested, meaning the majority did not.
-Figure \@ref(fig:agegroupcomparison) plots the estimates of these 22 items for individuals above 30 and individuals equal to or below 30 to get a sense of the magnitude of difference between older and younger individuals on these items.
-
-[Figure \@ref(fig:agegroupcomparison) about here.]
-
-The majority of items that show evidence for age concentration show that active updating is more prevalent among younger respondents than older respondents.
-These items include views on affirmative action, women in the workforce, and politics; several civil liberties items; general views of whether people can be trusted; and views on whether doctors should let terminal patients die.
-These items tend to be in subject areas where a large proportion showed no evidence of active updating, which suggests an overall trend of these views being formed earlier in life (i.e., prior to becoming eligible for the GSS at 18) and remaining relatively stable over time.
-
-For some items, such as whether individuals can be trusted, political views, whether physicians should allow terminal patients to die, and whether companies should make special efforts to hire and promote women to address past discrimination, all evidence of active updating disappears for people over 30.
-This suggests that these items follow an "impressionable years" pattern, where early adulthood is a time when these opinions are still malleable, but beliefs quickly harden into "durable dispositions" [@vaiseylizardo; @krosnickalwin; @alwinkrosnick].
-For other items, such as how important people believe it is for children to be popular and views on how much the government should spend on health care, there is still evidence of active updating in older individuals even though it is substantially less than for younger individuals.
-This is consistent with the "increasing persistence" hypothesis, where attitude change gradually becomes less likely as individuals age [@glenn; @inglehartbaker].
-
-Eight items show a negative effect of being below the age cutoff on the $\phi$ value, meaning that younger people showed _less_ evidence of active updating than older individuals.
-These items include how often individuals were active in religious activities, views on suicide in the case of bankruptcy, and views on whether aging parents should live with their children.
-Some of these items might be things people are not forced to consider until later in life and as a result do not form clear opinions on while young.
-This pattern where older individuals change their attitudes and behaviors at higher rates than younger individuals is somewhat unanticipated in the attitude change literature [@visserkrosnick; @danigelis], and suggests greater heterogeneity in the relationship between age and attitude change than previously theorized.
-
-The remaining 86 items (just under half of all items explored here) show evidence for some active updating but do not show consistent evidence for age heterogeneity, suggesting a more complicated relationship between age and attitude change than previously theorized.
-However, this does not mean these show strong evidence of an active updating model.
-These items may simply be susceptible to updating for a small proportion of the population.
-
-## Item heterogeneity
-
-While 40 percent of items show no evidence of active updating, and those that do show evidence tend to show only weak support for active updating, it is difficult with just these findings to draw any broad conclusions about how these results speak to theories of attitude development and change.
-Here we bring in our second dimension of attitude change, consistency in responses, to clarify the overall pattern.
-
-Figure \@ref(fig:phibetascatter) plots the $\phi$ and $\beta$ estimates for items shaded by whether they preferred the $\phi = .5$ constraint or not and label a few items that stand out.
-Items tend to prefer the $\phi=0.5$ constraint for a combination of two reasons: because wave 1 and wave 2 have equal predictive power ($\phi$ is close to .5) or because the measure is so unpredictable ($\beta$ is low) that neither wave 1 nor wave 2 has much predictive power, making any observed active updating close to meaningless.
-
-[Figure \@ref(fig:phibetascatter) about here.]
-
-Items showing evidence of active updating tend to have $\phi$ estimates greater than .55, and most have $\beta$ estimates greater than .6.
-A small group of variables, including confidence in the leadership of the executive branch of the federal government have low $\beta$ estimates, meaning that prediction at wave 3 is difficult, but have large $\phi$ estimates, meaning that wave 2 is still a better predictor than wave 1.
-
-Which items show the strongest evidence for active updating?
-There is no way we can discuss all 183 items in detail without the discussion becoming tedious.
-Although we include $\phi$ estimates for all items in Appendix A, Figure \@ref(fig:summary) summarizes the distributions of $\phi$ by the content of the question.
-We constrain items that showed no evidence of active updating to $\phi = .5$.
-In addition to showing the median and interquartile range of each distribution, the figure also highlights the item in each group that shows the greatest degree of evidence for active updating.
-
-[Figure \@ref(fig:summary) about here.]
-
-There is a lot to process, even in this summary figure.
-The main takeaway is that, even for items that show some evidence for active updating, the values of $\phi$ are still quite low.
-Only two groups of items have median $\phi$ values above $0.6$: public spending and religious activity.
-In general, then, it is accurate to say that most of the "change" measured by the GSS is not persistent but some combination of measurement error and short-term fluctuations.[^error]
-
-[^error]: While we cannot control for measurement error in our analysis, we can take steps to mitigate its impact.
-Appendix B presents results comparing items with more than three scale points to coarsened versions of these question with either two or three scale points.
-As noted previously, low measurement error might be a reasonable assumption for some items.
-Previous studies using different approaches to measuring the reliability of survey reports suggest that some items captured in our study, such as whether a person owns a gun, are measured with a high degree of reliability ($> 0.9$) [@houthastings].
-For other items, such as confidence in the leadership of major companies ($\phi = .58$), reliability might be as low as 0.5.
-There is very little correlation ($\rho = .165$) between our $\phi$ estimates and reliability estimates.
-
-Consistent with the findings of Vaisey and Lizardo [-@vaiseylizardo], we see that more than half of items in the gender, family, race, sex, civil liberties, and confidence in institutions groups show no evidence for updating (see Appendix B).
-These categories also contain several items that show evidence of active updating in younger cohorts, suggesting that these items became settled by the time respondents entered the GSS sample.
-Views on these issues are likely shaped by early socialization experiences and mostly settled by the time a respondent reaches adulthood.
-This means that, for most of these items, social change must occur through cohort succession rather than through individual change.
-
-There are several exceptions to this general pattern, however, even in categories with otherwise low $\phi$ values.
-Items with the largest values generally have one or more attributes in common.
-We consider these attributes to give some general impressions of the pattern.
-
-Some of the high-$\phi$ items rely on _external mechanisms_ that help maintain them.
-If a person starts going to church or starts socializing with friends at a bar, she builds social networks that make this behavior more likely to continue.
-This is clear when contrasted with how often individuals socialize with friends, relatives, or neighbors, which are more nebulous questions that display less active updating.
-Switching political parties (which involves changing public registration) is a more persistent change than changing political ideology (which can happen privately in the mind).
-Owning a gun has a high $\phi$ value because a new physical object either enters or leaves the person's possession.
-
-Other high-$\phi$ items have a _changing referent_.
-That is, although the item wording is the same, the object to which the question refers may change between survey waves.
-The most obvious example of this is the item about confidence in the executive branch of the federal government (which has the highest $\phi$ value of all items in the analysis).
-The president changed between the 2008 and 2010 waves of the GSS, meaning that the question no longer referred to the same administration.
-If we generate estimates for this item for each of the three panels (2006-10, 2008-12, and 2010-14) for the "confidence in the leadership of the executive branch of the federal government" item, it is only for the middle panel (2008-12), where the president changed between waves 1 and 2, that shows significant evidence of persistent change ($\phi = .95$, $\beta = .51$).
-In the 2006-10 panel, waves 1 and 2 have almost no predictive power ($\beta = .10$).
-For the 2010-14 panel, which takes place entirely during the Obama administration, $\phi$ moves much closer to .5 and predictive power increases ($\phi = .57$, $\beta = .65$).
-
-Likewise, all public spending items refer to whether we're spending "too much, too little, or the right amount" on different areas.
-The change of administration and changing federal spending policies likely affected these items.
-The same applies to most of the questions about subjective SES, where the respondent is asked questions about her personal financial or work situation (which changed for many Americans during the time of the study due to the Great Recession).
-If the environment is changing we should see exactly this sort of pattern.
-
-Perhaps the most striking pattern in our findings pertains to questions about _gay rights_.
-There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating.
-Questions about civil liberties for gays and about gay marriage are the highest in their categories.
-The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views.
-This pattern is consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.
-
-A few items lack external mechanisms, represent forms of private culture, and did not achieve salience during this period, but still display active updating, including some views on abortion and the morality of different forms of sex.
-We address these items in greater detail in the discussion section.
-
-## Political Beliefs
-
-Because theories of public opinion formation and the development of political beliefs form the bulk of the theoretical tension that frames this analysis, and because many of these beliefs display evidence of active updating, we examine these beliefs in greater detail than other items.
-Figure \@ref(fig:polvariables) plots the $\phi$ and $\beta$ estimates for questions about political identity, the role of government, and specific policies.
-We break these items up into categories and remove a few items for ease of viewing.
-
-[Figure \@ref(fig:polvariables) about here.]
-
-There are several notable features of the figure.
-First, political party affiliation is a clear outlier, with greater consistency and greater active updating than other items.
-In contrast, questions asking about specific public spending priorities tend to have low consistency compared to other items.
-Political views – including ideological identification (Named “Political views” in the figure) and general views of the role of government ("Gov't do more or less", "Gov't reduce inequality") – display higher consistency than specific policy questions but weak evidence of active updating.
-It is important to restate that for political views, all evidence of active updating disappears by the time individuals reach 30, and individuals become more consistent in reporting their ideological identity as they age.
-
-While people tend to be inconsistent in their responses to spending priorities, views on other policy questions with both high and low salience are reported quite consistently.
-Views on abortion, especially when considered as a scale, are consistently reported.
-As with all other scales, aggregating the composite items increases consistency but does not affect the estimate of the amount of active updating, validating our method's ability to separate persisting and non-persisting change.
-As discussed previously, views on gay marriage display high levels of active updating.
-In contrast, the item asking about support for legalizing marijuana use is reported with consistency but weak active updating.
-This is notable, as public opinion and policies regarding both have shifted considerably in recent years.
-Previous work suggests that change in both could be driven by the same underlying process [@schnabelsevell].
-Our findings suggest that, at least in recent years, many individuals have changed their views on gay marriage, but changing views on marijuana have been driven primarily by cohort replacement.
-
-# Discussion and Conclusions
-
-This study was motivated by a theoretical contrast in cultural sociology between the settled dispositions model (SDM), which emphasizes the power of the past, and the active updating model (AUM), which emphasizes contemporary meaning making.
-These two models of individual change are implicit in analyses of belief change at the society level, but they are rarely compared empirically.
-We asked three general questions in light of this distinction and developed an empirical approach to adjudicating (albeit imperfectly) between these models.
-We now revisit our research questions to summarize what we have learned.
-
-First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model.
-Around 40 percent of all items show no evidence for updating.
-For the items that do show evidence of active updating, the overall rate of persisting change in the population is likely low.
-For most items measured by the GSS, less than one percent of the population appears to make any persisting change in their views in a two-year period, and most changes appear to be short term or random deviations.
-What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago.
-Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is.
-But the average level of updating for most views appears to be low.
-
-Second, _is there evidence that younger respondents are doing more active updating than older respondents?_ For a limited subset of items, there seems to be evidence that younger respondents are updating their views whereas older respondents are not.
-This is consistent with a "cohortization" model that views young respondents as susceptible to updating shocks and older respondents as relatively insensitive to such shocks (see e.g., Bartels and Jackman 2014).
-Because the youngest respondents in the GSS are 18 years old, we may lack the ability to detect updating that occurs even earlier, so this is probably an underestimate.
-This may be the case for the roughly 70 items that show no evidence of persistent change over time.
-Overall, however, there is significant heterogeneity in the relationship between age and attitude and behavior change.
-
-Third, _are there systematic differences in item content between questions that are better described by each model?_ With 183 items tapping very different kinds of opinions and demonstrating a range of active updating and consistency, there is no one overall pattern of attitude change in the population.
-We find patterns of responses that provide a range of support for diverse theories of attitude formation, which we discuss below.
-At the same time, our results are consistent with some general expectations about updating.
-Questions with public (or otherwise changing) referents and questions tapping high-salience topics over the study period (such as gay rights) showed the most evidence for active updating.
-Items assumed to be "peripheral" beliefs, such as specific policy questions, also showed evidence for updating and low consistency.
-Most items about gender, family, race, and institutions showed the least evidence for updating, suggesting most people's views on those topics are settled by the time they turn 18.
-
-There are a handful of items that demonstrate active updating and enough consistency to merit further consideration, but are not easily explained by theoretical mechanisms outlined so far.
-These include questions about abortion in the case of rape or poverty, the morality of premarital and teen sex, the ideal number of children in a family, and whether children should obey adults or think for themselves.
-These items did not achieve high salience during the time of the study and there were not, to the best of our knowledge, clear changes in elite opinion on these issues.
-This suggests additional mechanisms that can drive persisting change at the individual level, such as exposure to these ideas through mass media or by directly interacting with people in these groups.
-In the case of questions about children and family structures, it might be the experience of having children that leads to changes in these attitudes.
-It is important to note that even though these items demonstrate active updating, it is a small proportion of the population that are changing their views during the study window.
-
-The overall pattern of results, while diverse and at times hard to reconcile, does shed light on several theoretical debates. We now consider these implications.
-
-## Implications for Cultural Sociology
-
-In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly through the mechanism of cohort succession.
-Most beliefs about gender roles, sexual morality, and abortion appear to be settled by early adulthood.
-The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time.
-In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating.
-Even those items that did display strong evidence for active updating (e.g.
-"whites rich or poor" and "whites work hard") were still hard to predict wave after wave.
-
-Our results suggest that one reason attitudes are largely stable is because most issues simply do not reach the level of salience necessary to shift opinions.
-In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views.
-By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period.
-The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views.
-This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.
-
-While the dominant pattern is stability during adulthood, the results suggest greater heterogeneity in the relationship between age and attitude change than what is emphasized in existing theories.
-The most prominent and well supported theories of attitude change suggest a peak of susceptibility to attitude change early in adulthood and either a rapid or gradual decline in attitude change with age [@alwinkrosnick; @visserkrosnick].
-An additional view, the "life stages" hypothesis, suggests a similar pattern with a surge in attitude change late in life [@visserkrosnick].
-While this age-related decline appears to be true for many political attitudes, the pattern is not nearly as consistent for other kinds of attitudes.
-Some items, such as views on the Bible, suggest equal openness to attitude change as individuals age.
-Others, such as views on most abortion questions, suggest that early adult socialization is so strong that lifetime opinions are settled by the time most people reach even 18 years old.
-
-For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change.
-This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level.
-Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors that encourage openness to change at different ages.
-Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes.
-Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick].
-Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].
-
-## Implications for Political Sociology
-
-A major takeaway of our analysis is that ideological identity (identity as a liberal or conservative, and the extremity of this identification) was in all practical terms stable for individuals over 30.
-Respondents might express different positions from wave to wave, but in guessing what a person will say in the future, we are better off guessing the mean of their previous responses than their most recent response.
-While our sample does not cover a large enough window of time or the life course to say for certain whether this represents a regular pattern (perhaps there is some period-specific reason that younger individuals in our data changed while older individuals did not), the pattern is consistent with theories and previous findings that political dispositions become settled by age 30.
-With the exception of some low-profile government spending questions, most policy questions show greater evidence of active updating than this question of ideological identification, though the overall level of active updating is still limited.
-Partisan identification, in contrast, showed the highest degree of active updating of all political questions, as well as some of the highest consistency, and this updating was active across all age ranges.
-
-These results are consistent with theories that posit that, at least in the current era, ideological identification (view of oneself as liberal or conservative), rather than some other belief such as partisan identification, moral views, religion, or particular policy positions, is the "central" political belief [@boutylinevaisey; @converse].
-
-It is important not to overstate the role of this belief in forming other beliefs.
-While our findings suggest that individuals appear to bring their partisan identification in line with their ideological identification, there is limited evidence of adjusting other beliefs.
-This pattern is most consistent with Baldassari and Gelman's [-@baldassarigelman] model of "partisans without constraint," which suggests that individuals hold a few strong beliefs and align their partisan identification with these, rather than adopting beliefs as a function of their partisan identification.
-
-Since there is no evidence of ideological change for the majority of the sample, and only very weak evidence of changes in specific policy positions, even high profile ones, the pattern of results is not consistent with the popular conception of political polarization in which individuals become more extreme in their views over time.
-This is notable given the time frame of our study, which covered the Obama administration, a time that is commonly assumed to observe a conservative shift for Republicans and a liberal shift for Democrats.
-Similarly, our findings are not consistent with the idea that individuals adopt a partisan affiliation based on their social groupings and subsequently adjust their ideological commitments to conform to that [@greenetal].
-For this to be true, ideological identification and other political beliefs would have to demonstrate at least as much updating as partisan identification, which is not true in our data.
-
-The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile items such as partisan identity, abortion, and gay marriage, individuals appear to lack clear opinions on most specific policy questions.
-The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe].
-What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."
-
-The overall picture that emerges from evaluating the active updating and stability of political items in the GSS is one in which the majority of respondents, especially those over 30 years old, hold a general political identity and a few clear views on issues like abortion, attempt to align their partisan identification to their views, and respond to elite opinion change when it provides clear signals.
-It is not a picture of a rapidly polarizing society or one wholly ignorant of public debates.
-
-## Methodological Implications
-
-Methodologically, our results highlight the challenges of evaluating population-wide attitude change using short-term panel studies.
-The evidence strongly suggests that most of what might be interpreted as "change" in the GSS panels is some combination of measurement error or non-persistent change.
-It does not matter whether measurement error or short-term change is the predominant driver behind this pattern; what matters is that substantive change is too rare in a sample of adults to measure accurately on the vast majority of items.
-This strongly argues against using two-wave panels to measure attitude change, which do not allow researchers to separate persisting from non-persisting changes.
-
-The fact that persistent change is practically nonexistent for many items bolsters the case for using repeated survey responses to measure the reliability of survey items [@alwin; @houthastings], since it is often a valid assumption that the underlying view is unchanging.
-At the same time, our results call for greater focus on methodological tools that can separate short-term attitude change from measurement error.
-While we generally assume that lasting changes in attitudes are more likely to influence behavior, this is not necessarily true.
-Short-term attitude changes might be meaningful in shaping short-term behaviors, but identifying it is difficult.
-
-We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation.
-One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.
-
-We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning.
-We see too much inconsistency in responses for commitment bias to be a major explanatory factor.
-Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p.
-499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p.
-500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p.
-501], tend to show low active updating and low consistency.
-
-We cannot (and would not want to) rule out the possibility that panel conditioning is taking place in the GSS.
-We believe it is worthwhile to explore these same GSS panels for evidence of panel conditioning.
-However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
-
-Because many attitudes, including views on abortion, race, gender roles, social trust, and institutional confidence, have mostly stabilized by the time individuals enter the GSS, our results also call for greater emphasis on surveying the attitudes of adolescents and children to understand how these attitudes are formed.
-Panel studies tracing the political socialization of adolescents are rare but could be highly fruitful.
-In a similar vein, it does not seem worthwhile to ask certain GSS questions repeatedly.
-Questions about racial stereotypes, which show almost no consistency from wave to wave but have been asked every wave since 1996, strike us as particularly problematic.
-Those that are repeated should be specifically targeted to topics that are believed to be changing broadly (e.g., politics, gay rights).
-
-Our results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults.
-Understanding the social origins of individuals' attitudes requires greater focus on the "conditions of past production"—childhood and adolescence—that give rise to persistent beliefs in adulthood.
-
-\pagebreak
-
-# Appendix A: $\phi$ values for all variables
-
-Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) plot $\phi$ estimates for all items included in the analysis, grouped by subject material.
-
-[Figures \@ref(fig:religphi) through \@ref(fig:racgenphi) about here.]
-
 ```{r loadfiguredata}
-load(here("data", "figure_data.Rdata"))
+load(here("correction", "figure_data.Rdata"))
 ```
 
-# Appendix B: Coarsened variable estimates
+```{r loadcoarse}
+#load(here("correction", "coarse_vars.Rdata"))
 
-```{r loadcoarse}
-load(here("data", "coarse_vars.Rdata"))
-
-```
-
-Estiates of $\phi$ for coarsened items are presented in Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen).
-If scales had symmetrical scales with no clear midpoint (e.g., strongly agree/agree/disagree/strongly disagree), we coarsened those scales to two points (agree/disagree).
-If scales included a clear midpoint, we included that and coarsened responses on either side.
-For items on large scales such as hours of TV watched, we split responses into greater than the median or less than or equal to the median.
-If changes between ends of each scale are more persistent than changes within ends of these scales, then $\phi$ estimates of coarsened item should be higher than estimates for items with more response options.
-
-[Figures \@ref(fig:coarserelig) through \@ref(fig:coarseracgen) about here.]
-
-Generally speaking, most items with no evidence of active updating continue to have no evidence of active updating when coarsened.
-For items that showed evidence for active updating, the coarsened estimates tended to be very similar to the uncoarsened estimates, suggesting that persisting and non-persisting changes happen about as often between ends of these scales as they do within ends of these scales.
-
-There are a couple notable departures from this general pattern.
-Several political questions -- general political views, views on whether the government should reduce inequality, and views on whether the government should help blacks -- show decreased evidence of persisting change when coarsened.
-This suggests that changes around the midpoint of the scale tend to be measurement error, a finding that is consistent with previous work suggesting that individuals without settled political views tend to choose points around the middle of the response scale [@converse].
-
-\pagebreak
-
-# References
-
-```{r dag, message=FALSE, fig.width=5, fig.cap="Graphical representation of active updating and settled dispositions models over time."}
-aum <- 
-  dagitty("dag {
-        y1 -> y2 -> y3 ;
-        v1 -> y1 ;
-        v2 -> y2 ;
-        v3 -> y3 
-        }")
-coordinates( aum ) <-
-  list( x=c(y1 = 1, y2 = 3, y3 = 5, v1 = 1, v2 = 3, v3 = 5),
-        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1) )
-
-aum_dag <- aum %>% tidy_dagitty() %>% ggdag(node_size = 8) + 
-  theme_dag_blank() + 
-  labs(title = "Active Updating Model") +
-  scale_y_continuous(limits = c(.8,3.2)) +
-  scale_x_continuous(limits = c(.5, 5.5)) + 
-  theme(plot.title = element_text(hjust = 0.5))
-
-sdm <- 
-  dagitty("dag {
-        U -> {y1 y2 y3} ;
-        v1 -> y1 ;
-        v2 -> y2 ;
-        v3 -> y3 
-        }")
-coordinates( sdm ) <-
-  list( x=c(y1 = 1, y2 = 3, y3 = 5, v1 = 1, v2 = 3, v3 = 5, U = 3),
-        y=c(y1 = 2, y2 = 2, y3 = 2, v1 = 1, v2 = 1, v3 = 1, U = 3) )
-
-sdm_dag <- sdm %>% tidy_dagitty() %>% ggdag(node_size = 8) + 
-  theme_dag_blank() + 
-  labs(title = "Settled Dispositions Model") +
-  scale_y_continuous(limits = c(.8,3.2)) +
-    scale_x_continuous(limits = c(.5, 5.5)) + 
-  theme(plot.title = element_text(hjust = 0.5))
-
-gridExtra::grid.arrange(aum_dag, sdm_dag, nrow = 1)
 ```
@@
 
-load(here("data", "age_results.Rdata"))
+load(here("correction", "age_results.Rdata"))
 
@@
 Numbers in parentheses indicate the number of items in each topical group."}
-load(here("data", "summary_plot_data.Rdata"))
+load(here("correction", "summary_plot_data.Rdata"))
 
@@
 ```
-
-```{r coarserelig, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about religion, subjective socioeconomic status, and morale." }
-coarse_vars %>%
-  filter(small.cat %in% c("Relig. Activity", "Relig. identity & beliefs",
-                          "Social life", "Subjective SES", "Suicide",
-                          "Health & morale")) %>%
-  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
-  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
-  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
-  geom_point() + 
-  coord_flip() + 
-  theme_minimal() + 
-  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
-  scale_color_manual(values = c("black", "darkgray")) + 
-  theme(
-    strip.background = element_blank(),
-    strip.text.x = element_blank(),
-    strip.text.y = element_blank(),
-    strip.placement = "outside",
-    legend.position = "none",
-    legend.title = element_blank(),
-    axis.text=element_text(size=7),
-    legend.text=element_text(size=7)
-  )  +
-  expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate (Updating)")
-
-```
-
-```{r coarsepoli, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about politics, government, sex and sexuality."}
-coarse_vars %>%
-  filter(small.cat %in% c("Guns, laws, crime & police", "Politics & government",
-                          "Public spending", "Sex, sexuality & abortion")) %>%
-  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
-  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
-  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
-  geom_point() + 
-  coord_flip() + 
-  theme_minimal() + 
-  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
-  scale_color_manual(values = c("black", "darkgray")) + 
-  theme(
-    strip.background = element_blank(),
-    strip.text.x = element_blank(),
-    strip.text.y = element_blank(),
-    strip.placement = "outside",
-    legend.position = "none",
-    legend.title = element_blank(),
-    axis.text=element_text(size=7),
-    legend.text=element_text(size=7)
-  )  +
-  expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate (Updating)")
-```
-
-```{r coarseracgen, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for coarsened questions about race and gender."}
-coarse_vars %>%
-  filter(small.cat %in% c("Gender & family", "Race & immigration")) %>%
-  ggplot(aes(x = fct_rev(var.name), y = estimate, color = as.factor(stable))) + 
-  geom_hline(yintercept = .5, linetype = 2, color = "gray") + 
-  geom_pointrange(aes(ymin = ci.lower, ymax = ci.upper)) + 
-  geom_point() + 
-  coord_flip() + 
-  theme_minimal() + 
-  facet_grid(base_var ~ ., scales = "free_y", space = "free", switch = "both") + 
-  scale_color_manual(values = c("black", "darkgray")) + 
-  theme(
-    strip.background = element_blank(),
-    strip.text.x = element_blank(),
-    strip.text.y = element_blank(),
-    strip.placement = "outside",
-    legend.position = "none",
-    legend.title = element_blank(),
-    axis.text=element_text(size=7),
-    legend.text=element_text(size=7)
-  )  +
-  expand_limits(y = .9) +
-  labs(x = "", y = "Phi estimate (Updating)")
-```

### 2021-12-15 correction.Rmd
results corrected

--- 2021-12-15 corrected_figures.Rmd
@@
 ---
-title: "Corrected Figures"
-author: "Kevin Kiley"
-date: "12/15/2021"
+title: "Correction"
+author:
+- Kevin Kiley
+- Stephen Vaisey
+date: "12/14/2021"
 output: pdf_document
+bibliography: panel_change_bib.bib
 ---
@@
 ```
+
+We recently discovered that an error in the R source code for a function we used in the original analysis for our 2020 ASR paper, "Measuring Stability and Change in Personal Culture Using Panel Data," led to erroneous results, and we would like to correct the record.
+This bug, in which survey weights were incorrectly applied to residuals twice (rather than once) in calculating the log likelihood of non-linear least squares models, produces changes in how we classify some items as showing either “settled dispositions” or “active updating” throughout the paper, but it does not affect the overall pattern of results or our substantive interpretations of those results.
+We stand by the general claims we make in the paper.
+
+In our original paper, we estimated two non-linear least squares models -- one that modeled changes within people over time as random departures from a settled baseline (a settled dispositions model) and one that modeled changes within people over time as durable updates (an active updating model) -- for 183 variables from the General Social Survey's rotating three-wave panels.
+We then compared the fit of these two models using the Bayesian Information Criterion to evaluate which model better fit the observed data.
+We found the settled dispositions model was preferred for 75 questions, meaning there was no evidence of durable change for these items.
+We also found that even for questions where there was evidence of durable change, most change was not durable.
+
+We conducted this analysis using R version 3.5.3.
+In versions of R prior to 4.0.3, there was an error in the way R calculated the log likelihood of non-linear least squares models, which resulted in an incorrect calculation of the Bayesian Information Criterion.
+This led to errors in how we judged the two models against each other.
+
+In versions of R prior to 4.0.3, the log likelihood of non-linear least squares models was calculated as:
+
+$$ - \frac{N}{2}(log(2\pi) + 1 - log(N) - \sum_{i=1}^N{log(w) + log(\sum_{i=1}^N{[w^2*(y - \hat{y})]^2})}) $$
+
+where $N$ is the number of observations, $w$ are survey weights, $y$ is the observed value of the outcome, and $\hat{y}$ is the predicted value of the outcome.
+The key issue for our analysis was that the formula applied survey weights to residuals twice ($w^2$ instead of $w$).
+The formula also did not divide the sum of the log of the weights by N.
+This latter issue produced incorrect log likelihood values, but because this term is constant across both models, it did not affect comparisons.
+
+The corrected calculation of the log likelihood for non-linear least squares models, implemented in version 4.0.3,[^patch] is:
+
+$$ - \frac{N}{2}(log(2\pi) + 1 - log(N) - \frac{\sum_{i=1}^N{log(w)}}{N} + log(\sum_{i=1}^N{[w*(y - \hat{y})]^2})) $$
+
+[^patch]: Documentation of the error can be found: https://bugs.r-project.org/show_bug.cgi?id=16100#c3
+
+Because the error lies in how survey weights were applied to calculation of the BIC, there is no clear pattern in how results changed when the correct formula was applied.
+Re-running our analyses with the corrected calculation produces slightly different results than what we presented in the published paper.
+Seventeen variables where we found a better fit for the stable disposition model prefer the active updating model under the corrected calculation.
+Fourteen variables where we found that the active updating model fit the data better prefer the settled dispositions model under the corrected calculation.
+This resulted in a net change of three more variables preferring the active updating model to the settled dispositions model (75, compared to 72 in the original analysis).
+The revised calculation also found fewer variables with a statistically significant difference between the $\phi$ estimates for people 30 and under and people over 30 (9 instead of 22), but the overall pattern of younger people showing more active updating is still consistent.
+
+The items we labeled as "settled dispositions" under the incorrect calculation that are "active updating" under the corrected calculation are: abnomore, confinan, conjudge, contv, courts, fund, hapmar, misanthropy, neverrelact, polattak, reborn, socommun, socrel, spkath, spkrac, teensex, and xmovie.
+
+The items we labeled as "active updating" under the incorrect calculation that are "settled dispositions" under the corrected calculation are: aged, colrac, divlaw, getahead, marwht, meovrwrk, pillok, polhitok, racdif2, socfrend, suicide1, trust, uswary, and workblks.
+
+While fixing this calculation changes how these variables are classified, the erroneous calculation was unrelated to our estimates of $\phi$ or $\beta$, and the overall substantive conclusions of our paper, which we reproduced with different methods elsewhere [@vaisey2021], remain the same.
+
+Corrected versions of Figures 2, 3, 4, 5, and A1-A4 are included below. Figures 1 and 6 were not affected by the change.
+
+We thank Hugo Mercier for bringing the discrepancy to our attention. We thank Sebastian Meyer for patching the function in R and for explaining the change in personal communications.
+
+## Figures
 
@@
 
-```{r polvariables, message=FALSE, fig.width=7, fig.height=8, fig.cap="Phi (updating) and beta (consistency) estimates for 43 political attitudes and identities."}
-left_join(res.df, attitude_vars) %>%
-  filter(small.cat %in% c("Politics & government", "Public spending",
-                          "Sex, sexuality & abortion") |
-           var %in% c("cappun", "divlaw", "gunlaw", "grass", "courts",
-                      "letin1a")) %>%
-  mutate(small.cat = ifelse(small.cat %in% c("Race & immigration", 
-                                             "Guns, laws, crime & police") | 
-                              var %in% c("prayer", "uswary"), 
-         "Misc. policy" , small.cat)) %>%
-  filter(var %!in% c("abscale6", "absingle", "teensex", "xmarsex",
-                     "homosex", "premarsx")) %>%
-  mutate(var.name = gsub("Government", "Gov't", var.name),
-         var.name = gsub("Abortion", "Abort", var.name),
-         small.cat = ifelse(small.cat == "Sex, sexuality & abortion",
-                            "Abortion & sex", small.cat)) %>%
-  mutate(small.cat = factor(small.cat, c("Politics & government", "Public spending",
-                                        "Abortion & sex", "Misc. policy"))) %>%
-  ggplot(aes(x = p, y = b1)) + 
-  geom_vline(xintercept = .5, linetype = 2, color = "gray") + 
-  geom_point(aes(fill = small.cat), shape = 21) + 
-  geom_text_repel(aes(label = var.name), size = 1.75) + 
-  theme_bw() + 
-  labs(x = "Phi estimate (Updating)", y = "Beta estimate (Consistency)", fill = "", shape = "") + 
-  theme(axis.title=element_text(size=9)) +
-  facet_wrap(~small.cat) + 
-  theme(legend.position = "none")
-
-```
-
 ```{r religphi, message=FALSE, fig.width=6.8, fig.height=8.5, fig.cap="Phi estimates for items about religious activity and beliefs, social life, subjective SES, and suicide."}

### 2021-12-15 code/clean10.R

@@ -217,16 +217,16 @@ gss.wide10 <- gss.long10 %>%
   spread(v, value)
 
 
-weight06 <- gss06 %>% mutate(idnum = paste0("06", "-", id_1)) %>% select(idnum, wtpannr123) 
-weight08 <- gss08 %>% mutate(idnum = paste0("08", "-", id_1)) %>% select(idnum, wtpannr123)
+weight06 <- gss06 %>% mutate(idnum = paste0("06", "-", id_1)) %>% select(idnum, wtpannr12, wtpannr123) 
+weight08 <- gss08 %>% mutate(idnum = paste0("08", "-", id_1)) %>% select(idnum, wtpannr12, wtpannr123)
 weight10 <- gss10 %>% mutate(idnum = paste0("10", "-", id_1),
-                             wtpannr123 = WTPANNR123) %>% select(idnum, wtpannr123)
+                             wtpannr123 = WTPANNR123) %>% select(idnum, wtpannr12, wtpannr123)
 all.weights <- bind_rows(weight06, weight08, weight10)
 
 
 gss.wide <- bind_rows(gss.wide06, gss.wide08, gss.wide10) %>% left_join(all.weights)
 gss.long <- gss.wide %>% 
-  gather(key = "variable", value = "value", -c(idnum, ds, minage, wtpannr123)) %>%
+  gather(key = "variable", value = "value", -c(idnum, ds, minage, wtpannr12, wtpannr123)) %>%
   separate(variable, into = c("var", "wave")) %>%
   spread(var, value) %>%
   mutate(ds = recode(ds, "gss06"="06", "gss08"="08", "gss10"="10"))

### 2021-12-15 code/phi_beta_model.R

@@ -10,6 +10,7 @@ load("~/Desktop/panel_change/data/attitude_vars.Rdata")
 #Make a list of variables in the gss
 gss_vars <- attitude_vars$var
 
+
 #Function that creates a subset of the panel that is just a variable of 
 #interest, taken as a character vector, and omits missing cases
 cleanGSS <- function(var) {
@@ -18,7 +19,8 @@ cleanGSS <- function(var) {
     mutate(wave = paste("w", wave, sep = "")) %>%
     spread(wave, var) %>%
     na.omit() %>% filter(!is.infinite(minage)) %>%
-    mutate(var = var)
+    mutate(var = var) %>%
+    mutate(wtpannr123 = as.numeric(wtpannr123))
   return(data)
 }
 
@@ -40,7 +42,7 @@ for (i in 1:length(gss_vars)) {
               start = list(a = 0, b = 1, p = 0),
               weights = wtpannr123)
   bic1 <- BIC(nls1)
-  
+
   #Model that constrains phi to .5
   nls2 <- nls(y3 ~ a + b*.5*y2 + b*(.5)*y1,
               data = all.data,
@@ -70,6 +72,22 @@ res.df <- bind_rows(results) %>%
          bicdiff = bic1 - bic2,
          or_aum = exp(-bicdiff/2),
          prob_aum = or_aum/(or_aum + 1)) 
+
+new.res <- bind_rows(results) %>%
+  mutate(aum_pref = ifelse(bic1 < bic2, 1, 0),
+         bicdiff = bic1 - bic2,
+         or_aum = exp(-bicdiff/2),
+         prob_aum = or_aum/(or_aum + 1)) 
+
+
+
+test <- left_join(res.df, new.res, by = c("var" = "var")) 
+
+test %>% filter(aum_pref.x != aum_pref.y) %>% View()
+
+test %>%
+  ggplot(aes(x = bicdiff.x, y = bicdiff.y)) +
+  geom_point()
 
 #make a vector of variables that prefer the constraints
 stable <- res.df %>% 
@@ -269,4 +287,7 @@ label.points <- clean.pestimates %>%
 
 plot.data <- left_join(plot.data, label.points)
 save(plot.data, file = "~/Desktop/panel_change/data/summary_plot_data.Rdata")
+
+
+
 

### 2021-12-15 panel_change_mre.R

@@ -1,0 +1,58 @@
+
+#Settled dispositions model
+y_star <- rnorm(1000, 0, 1)
+y1 <- y_star + rnorm(1000, 0, .3)
+y2 <- y_star + rnorm(1000, 0, .3)
+y3 <- y_star + rnorm(1000, 0, .3)
+
+df <- data.frame(y1 = y1, y2 = y2, y3 = y3)
+
+nls1 <- nls(y3 ~ a + b*p*y2 + b*(1-p)*y1,
+            data = df,
+            start = list(a = 0, b = 1, p = 0))
+#p estimate is ~0.5, meaning waves 1 and 2 are equally predictive of wave3
+
+
+nls2 <- nls(y3 ~ a + b*.5*y2 + b*(.5)*y1,
+            data = df,
+            start = list(a = 0, b = 1))
+
+#Calculate BICs for each model
+bic1 <- BIC(nls1)
+bic2 <- BIC(nls2)
+
+#Calculate probability of active updating model
+bicdiff=bic2-bic1
+or=exp(bicdiff/2) 
+prob_aum=or/(or+1)
+prob_aum
+
+
+
+#Active updating model
+#A Random Walk
+y_star <- rnorm(1000, 0, 1)
+y1 <- y_star + rnorm(1000, 0, .3)
+y2 <- y1 + rnorm(1000, 0, .3)
+y3 <- y2 + rnorm(1000, 0, .3)
+
+df <- data.frame(y1 = y1, y2 = y2, y3 = y3)
+
+nls1 <- nls(y3 ~ a + b*p*y2 + b*(1-p)*y1,
+            data = df,
+            start = list(a = 0, b = 1, p = 0))
+
+nls2 <- nls(y3 ~ a + b*.5*y2 + b*(.5)*y1,
+            data = df,
+            start = list(a = 0, b = 1))
+#p estimate is ~1, meaning wave 1 gives no additional information when wave 2 present
+
+bic1 <- BIC(nls1)
+bic2 <- BIC(nls2)
+
+#Calculate probability of active updating model
+bicdiff=bic2-bic1
+or=exp(bicdiff/2) 
+prob_aum=or/(or+1)
+prob_aum
+

### 2023-10-27 correction.Rmd
correction files

@@
 ---
-title: "Correction"
+title: "Correction: Measuring Stability and Change in Personal Culture Using Panel Data"
 author:
@@
 - Stephen Vaisey
-date: "12/14/2021"
-output: pdf_document
+date: "1/3/2022"
+output:
+  pdf_document:
+    number_sections: yes
+    keep_tex: true
+    latex_engine: xelatex
+  bookdown::word_document2: 
+    fig_caption: yes
+    reference_docx: word-styles-reference-02.docx
+header-includes:
+  - \usepackage{setspace}
+  - \doublespacing
+indent: true
 bibliography: panel_change_bib.bib
@@
 We recently discovered that an error in the R source code for a function we used in the original analysis for our 2020 ASR paper, "Measuring Stability and Change in Personal Culture Using Panel Data," led to erroneous results, and we would like to correct the record.
-This bug, in which survey weights were incorrectly applied to residuals twice (rather than once) in calculating the log likelihood of non-linear least squares models, produces changes in how we classify some items as showing either “settled dispositions” or “active updating” throughout the paper, but it does not affect the overall pattern of results or our substantive interpretations of those results.
-We stand by the general claims we make in the paper.
+Correcting this bug in the function, in which survey weights were incorrectly applied to residuals twice (rather than once) in calculating the log likelihood of non-linear least squares models,  changes how we classify some items as showing either “settled dispositions” or “active updating” throughout the paper.
+It does not affect the overall pattern of results or our substantive interpretations of those results.
+We stand by the general claims we made in the paper, which we reproduced with alternative methods elsewhere [@vaisey2021].
 
-In our original paper, we estimated two non-linear least squares models -- one that modeled changes within people over time as random departures from a settled baseline (a settled dispositions model) and one that modeled changes within people over time as durable updates (an active updating model) -- for 183 variables from the General Social Survey's rotating three-wave panels.
+In our original paper, we estimated two non-linear least squares models -- one that modeled changes within people over time as random departures from a settled baseline, which we called a "settled dispositions model," and one that modeled changes within people over time as durable updates, which we called an "active updating model" -- for 183 items, including several composite scales, from the General Social Survey's rotating three-wave panels.
 We then compared the fit of these two models using the Bayesian Information Criterion to evaluate which model better fit the observed data.
@@
 We conducted this analysis using R version 3.5.3.
-In versions of R prior to 4.0.3, there was an error in the way R calculated the log likelihood of non-linear least squares models, which resulted in an incorrect calculation of the Bayesian Information Criterion.
+In versions of R prior to 4.0.3, there were two errors in the way R calculated the log likelihood of non-linear least squares models, which resulted in an incorrect calculation of the Bayesian Information Criterion.
 This led to errors in how we judged the two models against each other.
@@
 The formula also did not divide the sum of the log of the weights by N.
-This latter issue produced incorrect log likelihood values, but because this term is constant across both models, it did not affect comparisons.
+This latter issue produced log likelihood values that were too large, but because this term was constant across both models, it did not affect comparisons.
 
@@
 
-Because the error lies in how survey weights were applied to calculation of the BIC, there is no clear pattern in how results changed when the correct formula was applied.
+Because the error lies in how survey weights were applied to the calculation of the BIC, there is no clear pattern in how results changed when the correct formula was applied.
 Re-running our analyses with the corrected calculation produces slightly different results than what we presented in the published paper.
@@
 
-Corrected versions of Figures 2, 3, 4, 5, and A1-A4 are included below. Figures 1 and 6 were not affected by the change.
+Corrected versions of Figures 2, 3, 4, 5, and A1-A4 are included below.
+Figures 1 and 6 were not affected by the change.
+Code to reproduce the analyses, including the old and corrected results, can be found at: https://github.com/krkiley/panel_change.
 
@@
 
-## Figures
+# Figures
 
@@
   labs(x = "", y = "Phi estimate (Updating)", color = "") +
-  theme(axis.text=element_text(size=7),
+  theme(axis.text=element_text(size=9),
         axis.title=element_text(size=9),
@@
     legend.title = element_blank(),
-    axis.text=element_text(size=7),
+    axis.text=element_text(size=9),
     legend.text=element_text(size=9)
@@
     strip.text.x = element_blank(),
-    strip.text.y =element_text(size=8),
+    strip.text.y =element_text(size=9),
     strip.placement = "outside",
@@
     legend.title = element_blank(),
-    axis.text=element_text(size=7),
-    legend.text=element_text(size=7)
+    axis.text=element_text(size=9),
+    legend.text=element_text(size=9)
   )  +
@@
     strip.text.x = element_blank(),
-        strip.text.y =element_text(size=8),
+        strip.text.y =element_text(size=9),
     strip.placement = "outside",
@@
     strip.text.x = element_blank(),
-            strip.text.y =element_text(size=8),
+            strip.text.y =element_text(size=9),
     strip.placement = "outside",
@@
     strip.text.x = element_blank(),
-                strip.text.y =element_text(size=8),
+                strip.text.y =element_text(size=9),
     strip.placement = "outside",
@@
 ```
+
+# References

### 2023-10-27 correction.tex
correction files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Correction: Measuring Stability and Change in Personal Culture Using Panel Data},
  pdfauthor={Kevin Kiley; Stephen Vaisey},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\doublespacing
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Correction: Measuring Stability and Change in Personal Culture
Using Panel Data}
\author{Kevin Kiley \and Stephen Vaisey}
\date{1/3/2022}

\begin{document}
\maketitle

We recently discovered that an error in the R source code for a function
we used in the original analysis for our 2020 ASR paper, ``Measuring
Stability and Change in Personal Culture Using Panel Data,'' led to
erroneous results, and we would like to correct the record. Correcting
this bug in the function, in which survey weights were incorrectly
applied to residuals twice (rather than once) in calculating the log
likelihood of non-linear least squares models, changes how we classify
some items as showing either ``settled dispositions'' or ``active
updating'' throughout the paper. It does not affect the overall pattern
of results or our substantive interpretations of those results. We stand
by the general claims we made in the paper, which we reproduced with
alternative methods elsewhere (Vaisey and Kiley 2021).

In our original paper, we estimated two non-linear least squares models
-- one that modeled changes within people over time as random departures
from a settled baseline, which we called a ``settled dispositions
model,'' and one that modeled changes within people over time as durable
updates, which we called an ``active updating model'' -- for 183 items,
including several composite scales, from the General Social Survey's
rotating three-wave panels. We then compared the fit of these two models
using the Bayesian Information Criterion to evaluate which model better
fit the observed data. We found the settled dispositions model was
preferred for 75 questions, meaning there was no evidence of durable
change for these items. We also found that even for questions where
there was evidence of durable change, most change was not durable.

We conducted this analysis using R version 3.5.3. In versions of R prior
to 4.0.3, there were two errors in the way R calculated the log
likelihood of non-linear least squares models, which resulted in an
incorrect calculation of the Bayesian Information Criterion. This led to
errors in how we judged the two models against each other.

In versions of R prior to 4.0.3, the log likelihood of non-linear least
squares models was calculated as:

\[ - \frac{N}{2}(log(2\pi) + 1 - log(N) - \sum_{i=1}^N{log(w) + log(\sum_{i=1}^N{[w^2*(y - \hat{y})]^2})}) \]

where \(N\) is the number of observations, \(w\) are survey weights,
\(y\) is the observed value of the outcome, and \(\hat{y}\) is the
predicted value of the outcome. The key issue for our analysis was that
the formula applied survey weights to residuals twice (\(w^2\) instead
of \(w\)). The formula also did not divide the sum of the log of the
weights by N. This latter issue produced log likelihood values that were
too large, but because this term was constant across both models, it did
not affect comparisons.

The corrected calculation of the log likelihood for non-linear least
squares models, implemented in version 4.0.3,\footnote{Documentation of
  the error can be found:
  \url{https://bugs.r-project.org/show_bug.cgi?id=16100\#c3}} is:

\[ - \frac{N}{2}(log(2\pi) + 1 - log(N) - \frac{\sum_{i=1}^N{log(w)}}{N} + log(\sum_{i=1}^N{[w*(y - \hat{y})]^2})) \]

Because the error lies in how survey weights were applied to the
calculation of the BIC, there is no clear pattern in how results changed
when the correct formula was applied. Re-running our analyses with the
corrected calculation produces slightly different results than what we
presented in the published paper. Seventeen variables where we found a
better fit for the stable disposition model prefer the active updating
model under the corrected calculation. Fourteen variables where we found
that the active updating model fit the data better prefer the settled
dispositions model under the corrected calculation. This resulted in a
net change of three more variables preferring the active updating model
to the settled dispositions model (75, compared to 72 in the original
analysis). The revised calculation also found fewer variables with a
statistically significant difference between the \(\phi\) estimates for
people 30 and under and people over 30 (9 instead of 22), but the
overall pattern of younger people showing more active updating is still
consistent.

The items we labeled as ``settled dispositions'' under the incorrect
calculation that are ``active updating'' under the corrected calculation
are: abnomore, confinan, conjudge, contv, courts, fund, hapmar,
misanthropy, neverrelact, polattak, reborn, socommun, socrel, spkath,
spkrac, teensex, and xmovie.

The items we labeled as ``active updating'' under the incorrect
calculation that are ``settled dispositions'' under the corrected
calculation are: aged, colrac, divlaw, getahead, marwht, meovrwrk,
pillok, polhitok, racdif2, socfrend, suicide1, trust, uswary, and
workblks.

While fixing this calculation changes how these variables are
classified, the erroneous calculation was unrelated to our estimates of
\(\phi\) or \(\beta\), and the overall substantive conclusions of our
paper, which we reproduced with different methods elsewhere (Vaisey and
Kiley 2021), remain the same.

Corrected versions of Figures 2, 3, 4, 5, and A1-A4 are included below.
Figures 1 and 6 were not affected by the change. Code to reproduce the
analyses, including the old and corrected results, can be found at:
\url{https://github.com/krkiley/panel_change}.

We thank Hugo Mercier for bringing the discrepancy to our attention. We
thank Sebastian Meyer for patching the function in R and for explaining
the change in personal communications.

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/phihist-1.pdf}
\caption{Distribution of phi estimates and probabilities that items show
evidence of active updating.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/agegroupcomparison-1.pdf}
\caption{Comparison of phi estimates for individuals over and equal to
or less than 30 years old, with 95 percent confidence intervals.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/phibetascatter-1.pdf}
\caption{Distribution of beta and phi estimates for GSS items, by
whether model prefers phi = .5 constraint.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/summary-1.pdf}
\caption{Summary of phi estimates for all items, by topical group.
Numbers in parentheses indicate the number of items in each topical
group.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/religphi-1.pdf}
\caption{Phi estimates for items about religious activity and beliefs,
social life, subjective SES, and suicide.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/poliphi-1.pdf}
\caption{Phi estimates for items about guns, law, crime and policing;
politics and government; and public spending. The item `Police can hit
citizens,' which has phi = .43, has been removed for ease of viewing}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/civlibphi-1.pdf}
\caption{Phi estimates for items about civil liberties, confidence in
leadership, health, morale, and social trust.}
\end{figure}

\begin{figure}
\centering
\includegraphics{correction_files/figure-latex/racgenphi-1.pdf}
\caption{Phi estimates for items about race, gender, sex, sexuality, and
abortion.}
\end{figure}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-vaisey2021}{}%
Vaisey, Stephen, and Kevin Kiley. 2021. {``A Model-Based Method for
Detecting Persistent Cultural Change Using Panel Data.''}
\emph{Sociological Science} 8 (5): 83--95.
\url{https://doi.org/10.15195/v8.a5}.

\end{CSLReferences}

\end{document}

### 2023-10-27 code/phi_beta_model.R

@@ -6,6 +6,7 @@
 #### Load in a long version of the GSS panels.
 load("~/Desktop/panel_change/data/gsslong.Rdata")
 load("~/Desktop/panel_change/data/attitude_vars.Rdata")
+load("~/Desktop/panel_change/data/good_vars.Rdata")
 
 #Make a list of variables in the gss
 gss_vars <- attitude_vars$var
@@ -73,29 +74,13 @@ res.df <- bind_rows(results) %>%
          or_aum = exp(-bicdiff/2),
          prob_aum = or_aum/(or_aum + 1)) 
 
-new.res <- bind_rows(results) %>%
-  mutate(aum_pref = ifelse(bic1 < bic2, 1, 0),
-         bicdiff = bic1 - bic2,
-         or_aum = exp(-bicdiff/2),
-         prob_aum = or_aum/(or_aum + 1)) 
-
-
-
-test <- left_join(res.df, new.res, by = c("var" = "var")) 
-
-test %>% filter(aum_pref.x != aum_pref.y) %>% View()
-
-test %>%
-  ggplot(aes(x = bicdiff.x, y = bicdiff.y)) +
-  geom_point()
-
 #make a vector of variables that prefer the constraints
 stable <- res.df %>% 
   filter(aum_pref == 0)
 stable_vars <- stable$var
 
 #Save these data sets
-save(res.df, file = "~/Desktop/panel_change/data/results_df.Rdata")
+save(res.df, file = "~/Desktop/panel_change/correction/results_df.Rdata")
 
 #select only items where AUM is preferred
 aum_vars <- res.df %>% filter(aum_pref == 1)
@@ -219,7 +204,7 @@ for (i in 1:length(age_vars)) {
      t.young, t.old)
 }
 
-save(age.results, file = "~/Desktop/panel_change/data/age_results.Rdata")
+save(age.results, file = "~/Desktop/panel_change/correction/age_results.Rdata")
 
 bind_rows(age.results) %>%
   mutate(ci.lower = estimate - 1.96 * std.error,
@@ -267,7 +252,7 @@ clean.pestimates <- bind_rows(p.estimates) %>%
   mutate(ci.lower = ifelse(var %in% stable_vars, .5, ci.lower),
          ci.upper = ifelse(var %in% stable_vars, .5, ci.upper)) 
 
-save(clean.pestimates, file = "~/Desktop/panel_change/data/figure_data.Rdata")
[…]
